---
title: "beige_vs_white_downsample_lazarescu_adipo_only"
author: "Sarah Hazell Pickering"
date: "2025-08-26"
output: html_document
---


Using Azimuth to transfer Adipocyte and progenitor cell identities found in lazarescu et al 2025 to the beige and white timecourse.

See publ data dir for reference generation.  

```{r libs, message=FALSE, warning=FALSE}
library(Azimuth)
library(BPCells)
library(Seurat)
library(ggplot2)
library(here)
library(tidyr)
library(ggpubr)
```

```{r params}
module = "beige_vs_white"
dataname = "downsample"
load_data = T #Load prebuilt Azimuth reference
subset_data = F

rmd_filename = "beige_vs_white_lazarescu_adipo"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))

indir = here("output", module,dataname)
publ_data = here("../other_data/lazarescu2025")
odir = here(indir,"lazarescu")
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4", "#85C2EA","#B3B3B3", "#F18486","#E31A1C")
clusters = clusters = MetBrewer::met.brewer("Redon", 12)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
other=ggsci::pal_npg()(7)
refclusters = c(rev(paletteer::paletteer_d("MetBrewer::Redon")),  RColorBrewer::brewer.pal("Dark2", n=8))
```

## Load Query Datasets

```{r}
if (subset_data){
  qso = readRDS(here(indir, "10%_complete_analysis.rds"))
  k.weight=5
} else {
  qso = readRDS(here(indir, "complete_analysis.rds"))
  k.weight=22
}
qso
```

```{r}
DimPlot(qso, group.by = "seurat_clusters", shuffle = T) + scale_colour_manual(values = clusters)
DimPlot(qso, group.by = "time", shuffle = T) + scale_colour_manual(values = pal)
```
## set reference dir
```{r}
azdir = here(publ_data, "azimuth_reference/adipo_only/pca_switch")
if (subset_data){
  azdir = here(azdir, "10%_subset")
}
```

## Map clusters

Not many anchors found for mapping :/ so we're adding a kweight limit to the Run Azimuth code aka. 
trace("RunAzimuth.Seurat", where=asNamespace("Azimuth"), edit=T)
adding kanchors=k.weight as a param to MappingScore on ln 107

```{r}
annot <- RunAzimuth(
    query = qso,
    reference = azdir,
    annotation.levels = "author_cell_type",
    k.weight=k.weight
    )
```

```{r}
colnames(annot@meta.data)
table(annot$predicted.author_cell_type)

summary(annot$predicted.author_cell_type.score)
```
## Plot Overlap

```{r}
DimPlot(annot, group.by = "predicted.author_cell_type", shuffle = T) + scale_colour_manual(values=refclusters)
FeaturePlot(annot,"predicted.author_cell_type.score") + scale_colour_gradientn(colours =  spectral)
```
## Add uncertain category

```{r filt_lazarescu_cell_type}
annot$filt.lazarescu.celltype = "uncertain"
annot$filt.lazarescu.celltype[annot$predicted.author_cell_type.score > 0.75] =  
  annot$predicted.author_cell_type[annot$predicted.author_cell_type.score > 0.75]

DimPlot(annot, group.by = "filt.lazarescu.celltype", shuffle = T) +
  scale_colour_manual(values = c("#00A087", "#E64B35","grey"))
```


```{r filt_lazarescu_cell_type_90}
annot$filt.lazarescu.celltype = "uncertain"
annot$filt.lazarescu.celltype[annot$predicted.author_cell_type.score > 0.9] =  
  annot$predicted.author_cell_type[annot$predicted.author_cell_type.score > 0.9]

DimPlot(annot, group.by = "filt.lazarescu.celltype", shuffle = T) + 
  scale_colour_manual(values = c("#00A087", "#E64B35","grey"))
ggsave(here(fig_dir, "lazarescu_celltype_with_uncertainty.pdf"), device = "pdf")
ggsave(here(fig_dir, "lazarescu_celltype_with_uncertainty.png"), device = "png")
```

```{r}
Idents(annot) = "condition"
DimPlot(annot, group.by = "filt.lazarescu.celltype", shuffle = T, cells = WhichCells(annot, idents=c("asc","white"))) + 
  scale_colour_manual(values = c("#00A087", "#E64B35","grey"))
DimPlot(annot, group.by = "filt.lazarescu.celltype", shuffle = T, cells = WhichCells(annot, idents=c("asc","beige"))) + 
  scale_colour_manual(values = c("#00A087", "#E64B35","grey"))
```


```{r}
VlnPlot(annot,"predicted.author_cell_type.score", group.by = "predicted.author_cell_type", pt.size = 0)

VlnPlot(annot,"predicted.author_cell_type.score", group.by = "predicted.author_cell_type", 
        split.by = "cond.time",  pt.size = 0, cols=pal)

VlnPlot(annot,"predicted.author_cell_type.score", group.by = "predicted.author_cell_type", 
        split.by = "seurat_clusters",  pt.size = 0, cols=pal)
```

## Pval test celltype score

```{r}
celltype_score = select(annot@meta.data, "seurat_clusters","cond.time",
                        "predicted.author_cell_type.score","predicted.author_cell_type") 
head(celltype_score)

stats = compare_means(predicted.author_cell_type.score~cond.time, data=celltype_score, group.by ="predicted.author_cell_type")

compute_stars <- function(p) {
  # Accepts a numeric vector; returns character vector of significance codes
  out <- rep("", length(p))
  out[!is.finite(p)] <- NA_character_           # NA/Inf/-Inf → NA
  out[p > 0.01]  <- "ns"
  out[p < 0.01] <- "*"
  out[p < 0.001] <- "**"
  out[p < 0.001] <- "***"
  return(out)
}

length(compute_stars(stats$p.adj))
nrow(stats)

stats = mutate(stats, p.adj.sig = compute_stars(p.adj))
stats
```

Every comparison significant via wilcoxon/t.test test for SA5 vs cond.time; but have a high p so all is statistically significant.  Need biological replication to prove its robust.  

## Cohens D

```{r}
library(effsize)
pairwise.cohen.d = function(comparisons, all_counts){
  per_time = list()

  for (comp in comparisons){
    d = cohen.d(filter(all_counts, group==comp[[1]]) %>% pull(counts), 
                filter(all_counts, group== comp[[2]]) %>% pull(counts),
                na.rm = T, pooled = T)
    per_time[[paste(comp, collapse = ".v.")]] = d$estimate *-1 #reverse direction
  }
  return(per_time)
}
```


```{r}
cond.time = levels(celltype_score$cond.time)
comparisons = list(cond.time[1:2], 
                   cond.time[2:3],
                   cond.time[3:4],
                   cond.time[4:5],
                   cond.time[c(1,5)])
celltype_score$counts = celltype_score$predicted.author_cell_type.score
celltype_score$group = celltype_score$cond.time

pairwise.cohen.d(comparisons , filter(celltype_score, predicted.author_cell_type == "SA5"))
```
Negligible differences in confidence scores between SA cells in white vs beige conditions at day3 (cohens d 0.144).  


```{r}
table(annot$predicted.author_cell_type, annot$cond.time)

table(annot$predicted.author_cell_type, annot$cond.time, annot$predicted.author_cell_type.score > 0.90)
table(annot$predicted.author_cell_type, annot$cond.time, annot$predicted.author_cell_type.score > 0.85)
table(annot$predicted.author_cell_type, annot$cond.time, annot$predicted.author_cell_type.score > 0.80)
```



```{r}
variable.genes = VariableFeatures(annot)
head(variable.genes)
length(variable.genes)
write.table(variable.genes, here(odir, "beige_vs_white.variable.genes.txt"),
            row.names = F, quote=F)
```


```{r save}
if (subset_data == FALSE){
  saveRDS(annot, file = here(odir, "complete_analysis.annot.rds"))
  sub = subset(annot, downsample=1000)
  saveRDS(sub,here(odir,"10%_complete_analysis.annot.rds"))
} else if (subset_data){ #if subset, don't save complete analysis
  saveRDS(annot,here(odir, "10%_subset_analysis.annot.rds"))
}
```

