---
title: "beige_vs_white_downsample"
author: "Sarah Hazell Pickering"
date: "2024-08-23"
output: html_document
---


Basic analysis of gene-level counts using seurat v5.  White vs beige and downsampling so that we have the same number of cells between conditions at each timepoint.  

Update umap number neighbours = 20 (rather than 30).

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(here)
library(patchwork)
```

```{r params}
module = "beige_vs_white"
dataname = "downsample"
load_data = T #loading data
subset_data = F

rmd_filename = "beige_vs_white_downsample"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)
indir = here("output", "adipogenesis", "initial_r2_only")
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4", "#B2DF8A","#85C2EA","#B3B3B3", "#F18486","#FDBF6F","#E31A1C")
clusters =   tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```


## Read data

```{r}
if (subset_data){
  filt = readRDS(here(indir, "10%_subset_analysis.rds"))
} else {
  filt = readRDS(here(indir, "complete_analysis.rds"))
  DefaultAssay(filt) = "RNA"
}
Idents(filt) = "sample"
table(Idents(filt))
filt$orig.ident= filt$sample
filt$cond.time = paste0(filt$condition,".", filt$time )
filt$cond.time = factor(filt$cond.time, levels = paste(c(rep("white",2), "asc",rep("beige",2)),sep=".",
                                                                c("day3","day1","day0","day1","day3")))
filt
```

```{r no_cells_presbs, dev=c('png','pdf')}
ggplot(filt@meta.data) + geom_bar(aes(x=cond.time, fill=sample)) + scale_fill_manual(values=pal) +
  theme(axis.text.x = element_text(angle=15, hjust=1)) + scale_y_continuous(expand = c(0, 0.05))
ggplot(filt@meta.data) + geom_bar(aes(x=cond.time, fill=cond.time)) + scale_fill_manual(values=pal[c(1,2,4,6,7)]) +
  theme(axis.text.x = element_text(angle=15, hjust=1)) + scale_y_continuous(expand = c(0, 0.05))
```

```{r}
filt$time.rep = paste0(filt$time,".", filt$rep )

ident_tab = group_by(filt@meta.data, sample, time.rep, condition) %>% count()
ident_tab

min_cells = group_by(ident_tab, time.rep) %>% summarise(min=min(n))
min_cells
```


```{r}
temp_sbs = list()
for (time.rep in min_cells$time.rep){
  print(time.rep)
  idents = ident_tab$sample[ident_tab$time.rep == time.rep]
  dwns = min_cells$min[min_cells$time.rep == time.rep]
  temp_sbs[[time.rep]] = subset(filt, idents=idents, downsample =dwns)
} 
temp_sbs
```

```{r}
sbs = merge(x=temp_sbs[[1]],y=temp_sbs[2:4])
sbs = JoinLayers(sbs)
sbs$sample = factor(sbs$sample, levels=levels(filt$sample))
sbs$orig.ident = sbs$sample
sbs$cond.time = paste0(sbs$condition,".", sbs$time )
sbs$cond.time = factor(sbs$cond.time, levels = paste(c(rep("white",2), "asc",rep("beige",2)),sep=".",
                                                                c("day3","day1","day0","day1","day3")))

sbs
table(sbs$sample)
rm(filt)
```

```{r no_cells_dwns, dev=c('png','pdf')}
ggplot(sbs@meta.data) + geom_bar(aes(x=sample, fill=sample)) + scale_fill_manual(values=pal) +
  theme(axis.text.x = element_text(angle=15, hjust=1)) + scale_y_continuous(expand = c(0, 0.05))
ggplot(sbs@meta.data) + geom_bar(aes(x=cond.time,fill=sample)) + scale_fill_manual(values=pal) +
  theme(axis.text.x = element_text(angle=15, hjust=1)) + scale_y_continuous(expand = c(0, 0.05))
ggplot(sbs@meta.data) + geom_bar(aes(x=cond.time,fill=cond.time)) + scale_fill_manual(values=pal[c(1,3,4,5,7)]) +
  theme(axis.text.x = element_text(angle=15, hjust=1))+ scale_y_continuous(expand = c(0, 0.05))
```


## Basic Normalisation Scale the read depth

Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. This is then natural-log transformed using log1p
```{r lognorm}
sbs= NormalizeData(sbs, normalization.method = "LogNormalize", scale.factor = 10000)
sbs#161Mb
```


```{r variable_features}
sbs <- FindVariableFeatures(sbs, selection.method = "vst", nfeatures = 2000)
sbs #163Mb
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(sbs), 10)

# plot variable features with and without labels
plot1 = VariableFeaturePlot(sbs)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```


```{r}
all.genes = rownames(sbs)
sbs = ScaleData(sbs, features=all.genes)
```

## PCA

```{r pca, dev=c('png','pdf')}
sbs = RunPCA (sbs, ndims.print=2)
DimPlot(sbs, reduction = "pca", cols = pal, shuffle = T) + theme_hp
DimPlot(sbs, reduction = "pca", cols = pal[c(4,2,1)], shuffle = T, group.by = "time") + theme_hp
```


```{r}
FeaturePlot(sbs, features= c("percent.mt","nCount_RNA","CHI3L1","ACTA2"))
DimHeatmap(sbs, dims = 1:6, cells = 500, balanced = TRUE)
```

## Determine the dimensionality

They encourage to complete downstream analyses with different numbers of PCs.  
```{r elbow}
ElbowPlot(sbs, ndims = 20)
ElbowPlot(sbs, ndims = 50)
```

## Cluster
```{r cluster}
sbs <- FindNeighbors(sbs, dims = 1:30)
sbs <- FindClusters(sbs, resolution = 0.5)
table(sbs@meta.data$seurat_clusters)
```

## Cluster composition

```{r cluster_composition}
composition = group_by(sbs@meta.data, sample,seurat_clusters) %>% count()
composition = composition[order(composition$n, decreasing = T),]
colnames(composition) = c("sample","seurat_clusters","n_cells")
#calculate proportions directly, then plot percentage just for the three cyc;ling cells
composition = group_by(composition, sample) %>% mutate(cells_per_sample = n_cells/sum(n_cells),
                                                           time =gsub(".*_day","day", gsub("_rep.*","",sample)))
composition = group_by(composition, seurat_clusters) %>% mutate(cells_per_cluster = n_cells/sum(n_cells))
head(composition)
```


```{r composition_plot, dev=c('png','pdf')}
ggplot(composition) + geom_bar(stat="identity",aes(y=sample, x=n_cells, fill=seurat_clusters)) + 
  scale_y_discrete(limits=rev) + scale_fill_manual(values=clusters)+
  scale_x_continuous(expand = c(0, 0))
ggplot(composition) + geom_bar(stat="identity",aes(y=sample, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=clusters)+
  scale_x_continuous(expand = c(0, 0))
ggplot(composition) + geom_bar(stat="identity",aes(fill=sample, x=n_cells, y=seurat_clusters)) +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)+
  scale_x_continuous(expand = c(0, 0))
ggplot(composition) + geom_bar(stat="identity",aes(fill=sample, x=n_cells, y=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev) + scale_fill_manual(values=pal)+
  scale_x_continuous(expand = c(0, 0))
```

```{r composition_plot_per_time}
ggplot(composition) + geom_bar(stat="identity",aes(y=time, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=clusters)+
  scale_x_continuous(expand = c(0, 0))
```


## UMAP
```{r umap}
sbs = RunUMAP(sbs, dims=1:30, n.neighbors = 20L)
DimPlot(sbs, reduction="umap", label=T, shuffle = T, cols=clusters)  + theme_hp
DimPlot(sbs, reduction="umap", split.by = "sample", ncol=3, shuffle = T,  cols=clusters) + theme_hp
DimPlot(sbs, group.by="sample", cols=pal, shuffle = T) + theme_hp
DimPlot(sbs, cols=pal[c(4,3,1)], shuffle = T, group.by="time")  + theme_hp
DimPlot(sbs, group.by="cond.time", cols=pal[c(1,3,4,5,7)], shuffle = T) + theme_hp
```

## t-SNE
```{r tsne}
sbs = RunTSNE(sbs, reduction="pca", dims=1:30)
DimPlot(sbs, reduction = "tsne",shuffle=T,  cols=clusters)  + theme_hp
DimPlot(sbs, reduction = "tsne", group.by= "sample", shuffle = T, cols=pal)  + theme_hp
DimPlot(sbs,  reduction = "tsne", cols=pal[c(4,2,1)], shuffle = T, group.by="time")  + theme_hp
```


## Find biomarkers
Min.pct sets the minimum percentage of cells for a marker to be found in

Does densify, which increases speed at the cost of memory help?  
```{r densify}
norm.markers = FindAllMarkers(sbs, only.pos = T, min.pct=0.25, 
                              logfc.threshold = 0.5, densify=T)
head(norm.markers); dim(norm.markers)
norm.markers = filter(norm.markers, p_val_adj < 0.01)
table(norm.markers$cluster)

filt.markers = as.data.frame(norm.markers %>% filter( pct.1 > 0.5) %>%
    arrange(avg_log2FC))

top.markers = as.data.frame(filt.markers %>% group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC))
top.markers
```

```{r}
time_per_cluster = group_by(composition, seurat_clusters, time) %>% summarise(n_cells = sum(n_cells)) %>% 
  slice_max(n=1, order_by = n_cells) %>% select(seurat_clusters, time)
time_per_cluster
top.markers = merge(top.markers, time_per_cluster, by.x="cluster", by.y="seurat_clusters")
head(top.markers)
```

 
 
```{r top_markers}
DotPlot(sbs, features=rev(unique(top.markers$gene)), group.by = "seurat_clusters") & 
  scale_colour_gradientn(colours = spectral) & coord_flip() 

top.markers$gene = factor(top.markers$gene, levels=rev(unique(top.markers$gene)))
overlap.to.plot = filter(top.markers, gene %in% top.markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(top.markers$gene))

ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point(colour="black",aes(size = avg_log2FC*1.5)) + geom_point() +
  scale_color_manual(values=clusters) + facet_grid(~time, scales="free_x", space="free_x")
```

```{r top_marker_genes, dev=c('png','pdf')}
include_all_plot = filter(norm.markers, gene %in% overlap.to.plot$gene & p_val_adj< 0.01)
include_all_plot$gene = factor(include_all_plot$gene, levels=levels(top.markers$gene))
include_all_plot = merge(include_all_plot, time_per_cluster, by.x="cluster", by.y="seurat_clusters")
a = ggplot(include_all_plot, aes(x=cluster, y=gene, colour=cluster, size=avg_log2FC)) + 
  geom_point(colour="black",aes(size = avg_log2FC*1.5)) + geom_point() +scale_color_manual(values=clusters)
a

a + facet_grid(~time, scales="free_x", space="free_x")
```



Visualisation methods

```{r marker_vis, dev=c('png','pdf')}
VlnPlot(sbs, features = c("MGP","ADIPOQ"))+ scale_fill_manual(values=clusters)
FeaturePlot(sbs, features = c("MGP","ADIPOQ","MT-ATP8","ACTA2")) 
top.markers$gene = as.character(top.markers$gene)

for (i in 1:(length(top.markers$gene)/4)){
  start = ((i-1)*4)+1; print(start)
  end = i*4; print(end)
  print(FeaturePlot(sbs, features = top.markers$gene[start:end]))
}

FeaturePlot(sbs, features = c("percent.mt", "nCount_RNA","nFeature_RNA"))
FeaturePlot(sbs, features = c("PC_1","PC_2","PC_3","PC_4")) 
```


```{r}
filt.markers %>%
    group_by(cluster) %>% 
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(subset(sbs,downsample=100), features= top10$gene) + scale_colour_manual(values=clusters)
```


## Cluster GO

```{r load_go}
molsig <- clusterProfiler::read.gmt(here("data/gene_sets/msigdb.v2024.1.Hs.symbols.gmt"))
head(molsig); nrow(molsig)
##Pick gene sets to look in bcos all is a lot
prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))
```

```{r}
filt.markers$cluster = factor(filt.markers$cluster, levels =0:max(as.integer(filt.markers$cluster)))
all.genes = rownames(sbs)
```


```{r basic_go}
go = compareCluster(gene~cluster, data=filt.markers, 
                      fun="enricher", TERM2GENE = some.molsig, universe = all.genes)
raw_go_terms = go %>% arrange(p.adjust) %>% group_by(cluster) %>% slice(1:2) %>% as.data.frame() %>% pull(ID)
```


```{r go_replace}
phrases= read.delim(here("data/gene_sets/abbreviations.txt"))
recap= read.delim(here("data/gene_sets/capitalisations.txt"))
```


Function to intelligently abbreviate GO terms and correct the capitalisation of initialisations based on the above lists.  To do: save these lists to file to make reproducible and concise.   

```{r abrev_go}
drama_shorten = function(ont) {
  for (x in 1:length(phrases)){
    ont = gsub( names(phrases)[x],phrases[x], ont, ignore.case = T)
  }
  a = abbreviate(gsub("_"," ", gsub("^[[:alpha:]]*_","",tolower(ont))),minlength=41, dot=T, named = F)
  a = stringr::str_to_sentence(a)
  for (i in 1:length(recap))
    a = gsub(sprintf("\\b%s\\b", names(recap)[i]),recap[i],a, ignore.case = T)
  return(a)
}
#raw_go_terms
drama_shorten(raw_go_terms)
```


```{r default_go_plots}
go = mutate(go, cluster = factor(cluster, levels=0:max(as.integer(cluster))))
head(go)[,c(2,4:9)]

p = dotplot(go, x="cluster", showCategory=2) 
p+ scale_y_discrete(labels=drama_shorten(levels(p$data$Description)))  + theme_hp
p = dotplot(go, x="cluster", showCategory=2, includeAll=F) 
p+ scale_y_discrete(labels=drama_shorten(levels(p$data$Description)))  + theme_hp
```


```{r set_up_go_table}
dfgo = as.data.frame(go)
dfgo$short_desc= drama_shorten(dfgo$Description)
dfgo = merge(dfgo, time_per_cluster,by.x="cluster", by.y="seurat_clusters")
go_to_plot = filter(dfgo, p.adjust < 0.01) %>% group_by(Cluster) %>%
              slice_min(n = 2, order_by = data.frame(p.adjust, pvalue))
head(go_to_plot)
```



## Save
```{r save}
if (subset_data == F){
  saveRDS(sbs, file = here(odir, "complete_analysis.rds"))
  write.table(norm.markers, here(odir, "marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(dfgo, here(odir, "ORA_marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(composition, here(odir, "cluster_composition.txt"),
              sep="\t", row.names = F)
  Idents(sbs) = "sample"
  sbs = subset(sbs, downsample=1000)
  saveRDS(sbs, file = here(odir, "10%_complete_analysis.rds"))
} else if (subset_data){
  saveRDS(sbs, file = here(odir,"10%_subset_analysis.rds"))
}
```


