---
title: "Downsample & moncle white & beige"
author: "Sarah Hazell Pickering"
date: "2024-09-19"
output: html_document
---

 We need a way to control for the random variation of monocle trajectory graph construction.  The best way to do this is to downsample our data 10 times and rerun. We can then using that information to inform the variance estimates or average the psueodtime values to provide a more robust estimate.


FastMNN integration at the gene level.  No loop closing.  

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(dplyr)
library(ggplot2)
library(here)
library(monocle3)
library(viridis)
```

```{r params}
module = "beige_vs_white"
method_name = "fastmnn"
dataname = paste0("tss_",method_name)
load_data = T #loading data
subset_data = F

if (exists("snakemake")){
    seed = snakemake@params[["seed"]]
} else(
    seed = as.numeric(gsub(".* |:","",Sys.time()))
)
print(seed)

rmd_filename = "tss_fastmnn_monocle_umap"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
indir = here("output", module, dataname) 
odir =  here(indir,paste0("subset_", seed))
```

```{r setup}
set.seed(seed)
fig_dir = here("analysis", module, "figures", paste0(rmd_filename,"_subset_", seed))
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
clusters = MetBrewer::met.brewer("Redon", 12)
names(clusters) = 0:11
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```


```{r}
if (subset_data){
  sbs = readRDS(here(indir, "10%_complete_analysis.rds"))
  Idents(sbs) = "seurat_clusters"
} else {
  sbs = readRDS(here(indir,"complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
sbs
```

# Subset
```{r}
Idents(sbs) = "cond.time"
table(Idents(sbs))
dns = subset(sbs, downsample = 3300, seed=seed)
table(Idents(dns))
```


# White 
## Split seurat object
```{r}
white = subset(dns, idents=c("asc.day0", "white.day1","white.day3"))
white
```

```{r}
Idents(white) = "initial_clusters"
table(Idents(white))
white$seurat_clusters = white$initial_clusters
```

```{r}
DimPlot(white, reduction="umap") + scale_colour_manual(values=clusters)
```

## Convert to monocle object
```{r warning=FALSE}
cds = SeuratWrappers::as.cell_data_set(white)
cds
```

```{r}
p= plot_cells(cds, show_trajectory_graph = F, cell_stroke=0.35, reduction_method = "UMAP", color_cells_by="seurat_clusters") + 
  scale_colour_manual(values=clusters)
p$layers[[1]]$aes_params$colour = 'transparent'
p
```

Overwriting the UMAP slot with tSNE because there might be a bug
```{r}
cds = cluster_cells(cds)
p= plot_cells(cds, show_trajectory_graph = F, color_cells_by = "partition", label_cell_groups = F)
p$layers[[1]]$aes_params$colour = 'transparent'
p
```

```{r}
cds = learn_graph(cds, close_loop = F, use_partition = F)
p=plot_cells(cds, color_cells_by = "seurat_clusters",
           label_leaves=FALSE,
           label_branch_points=FALSE, cell_stroke=0.35,
           trajectory_graph_color = "black")+ 
  scale_colour_manual(values=clusters)
p$layers[[1]]$aes_params$colour = 'transparent'
p
```


## Order cells

```{r}
pick_root_node = function(cds,factor_name, start_bin){
  cell_ids = which(colData(cds)[,factor_name]==start_bin)
  closest_vertex <-  cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  summary(closest_vertex)
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <- igraph::V(principal_graph(cds)[["UMAP"]])$name[as.numeric(names
    (which.max(table(closest_vertex[cell_ids,]))))]
return(root_pr_nodes)
}
pick_root_node(cds, "time","day0")
pick_root_node(cds, "time","day1")
pick_root_node(cds, "time","day3")
```


## Starting from proliferating cells

```{r}
cds = order_cells(cds, root_pr_nodes = pick_root_node(cds, "seurat_clusters", "8"))
p=plot_cells(cds, color_cells_by = "pseudotime", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="red") + scale_colour_viridis()
p$layers[[1]]$aes_params$colour = 'transparent'
p
```


## Adipogenic end point
```{r}
cds = order_cells(cds, root_pr_nodes = pick_root_node(cds, "seurat_clusters", "4"))
p=plot_cells(cds, color_cells_by = "pseudotime", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="red") + scale_colour_viridis()
p$layers[[1]]$aes_params$colour = 'transparent'
p
```
```{r trajectory_only}
p = plot_cells(cds, label_leaves = F, label_branch_points = F, label_roots = F,
                trajectory_graph_color="black", label_cell_groups = F, cell_size=0,
            graph_label_size = 5) 
p
saveRDS(p, file=here(odir, "white_trajectory_plot.rds"))
```


## Save

```{r}
cds$raw_pseudotime = pseudotime(cds)
cds$monocle_pseudotime = scales::rescale(cds$raw_pseudotime, to = c(1, 0)) #rescale from 1 to 0 (cos were tracing backwards from  day3)
cds$pscondition = "white"
if (subset_data == FALSE){
  write.table(colData(cds), here(odir,"white_monocle_pseudotime.tsv"), sep="\t", quote=F)
} 
```

# Beige

## Split seurat object
```{r}
Idents(dns) = "cond.time"
table(Idents(dns))
beige = subset(dns, idents=c("asc.day0", "beige.day1","beige.day3"))
beige
```

```{r}
Idents(beige) = "initial_clusters"
table(Idents(beige))
beige$seurat_clusters = beige$initial_clusters
```

```{r}
DimPlot(beige, reduction="umap") + scale_colour_manual(values=clusters)
```

## Convert to monocle object
```{r warning=FALSE}
cds = SeuratWrappers::as.cell_data_set(beige)
cds
```

```{r}
p= plot_cells(cds, show_trajectory_graph = F, cell_stroke=0.35, reduction_method = "UMAP", color_cells_by="seurat_clusters") + 
  scale_colour_manual(values=clusters)
p$layers[[1]]$aes_params$colour = 'transparent'
p
```


```{r}
cds = cluster_cells(cds)
p= plot_cells(cds, show_trajectory_graph = F, color_cells_by = "partition", label_cell_groups = F)
p$layers[[1]]$aes_params$colour = 'transparent'
p
```

```{r}
cds = learn_graph(cds, close_loop = F)
p=plot_cells(cds, color_cells_by = "seurat_clusters",
           label_leaves=FALSE,
           label_branch_points=FALSE, cell_stroke=0.35,
           trajectory_graph_color = "black")+ 
  scale_colour_manual(values=clusters)
p$layers[[1]]$aes_params$colour = 'transparent'
p
```


## Order cells

```{r}
pick_root_node = function(cds,factor_name, start_bin){
  cell_ids = which(colData(cds)[,factor_name]==start_bin)
  closest_vertex <-  cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  summary(closest_vertex)
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <- igraph::V(principal_graph(cds)[["UMAP"]])$name[as.numeric(names
    (which.max(table(closest_vertex[cell_ids,]))))]
return(root_pr_nodes)
}
pick_root_node(cds, "time","day0")
pick_root_node(cds, "time","day1")
pick_root_node(cds, "time","day3")
```


## Adipogenic end point
```{r}
cds = order_cells(cds, root_pr_nodes = pick_root_node(cds, "seurat_clusters", "4"))
p=plot_cells(cds, color_cells_by = "pseudotime", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="red") + scale_colour_viridis()
p$layers[[1]]$aes_params$colour = 'transparent'
p
```

```{r trajectory_only_beige}
p = plot_cells(cds, label_leaves = F, label_branch_points = F, label_roots = F,
                trajectory_graph_color="black", label_cell_groups = F, cell_size=0,
            graph_label_size = 5) 
p
saveRDS(p, file=here(odir, "beige_trajectory_plot.rds"))
```
## Save

```{r}
cds$raw_pseudotime = pseudotime(cds)
cds$monocle_pseudotime = scales::rescale(cds$raw_pseudotime, to = c(1, 0)) #rescale from 1 to 0 (cos were tracing backwards from  day3)
cds$pscondition = "beige"
if (subset_data == FALSE){
  write.table(colData(cds), here(odir,"beige_monocle_pseudotime.tsv"), sep="\t", quote=F)
} 
```
