---
title: "FastMNN integration - Beige vs White (gene-level)"
author: "Sarah Hazell Pickering"
date: "2024-04-11"
output: html_document
---

1. Integrate using Fastmnn
2. Find marker genes and GO terms for new FASTMNN clusters


```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(RColorBrewer)
library(Seurat)
library(SeuratWrappers)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(ggsci)
library(here)
library(patchwork)
```

```{r params}
module = "beige_vs_white"
method_name = "fastmnn"
dataname = paste0("bvw_",method_name)
load_data = T #loading data
subset_data = F

rmd_filename = "bvw_fastmnn_integration"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)
indir = here("output", module, "downsample")
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4", "#85C2EA","#B3B3B3", "#F18486","#E31A1C")
clusters = MetBrewer::met.brewer("Redon", 12)
spectral =  rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
```

```{r setup-variables-rpca}
red_name = paste("pca", sep=".", method_name)
umap_name = paste("umap", sep=".", method_name)
clu_name = paste( method_name, sep="_", "clusters")
tsne_name = paste("tsne", sep=".", method_name)
```

## Read data

```{r load_data}
if (subset_data){
  sbs = readRDS(here(indir, "10%_complete_analysis.rds"))
  Idents(sbs) = "seurat_clusters"
} else {
  sbs = readRDS(here(indir, "complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
table(Idents(sbs))
sbs[["RNA"]] =split(sbs[["RNA"]], f = sbs$time)
sbs = FindVariableFeatures(sbs, nfeatures = 2000)
sbs
```


```{r}
sbs$initial_clusters = sbs$seurat_clusters
sbs[["initial.umap"]] = sbs[["umap"]]
sbs[["initial.tsne"]] = sbs[["tsne"]]
```

```{r}
DimPlot(sbs, cols=clusters, label=T)
```
## Integrate layers

```{r setup-variables-fastmnn}
method_name="fastmnn"
red_name = paste("pca", sep=".", method_name)
umap_name = paste("umap", sep=".", method_name)
clu_name = paste( method_name, sep="_", "clusters")
tsne_name = paste("tsne", sep=".", method_name)
```

```{r}
sbs = IntegrateLayers(sbs, method= FastMNNIntegration, orig.reduction = "pca",
                       new.reduction = red_name)
```

```{r}
sbs <- FindNeighbors(sbs, reduction =  red_name, dims = 1:30, verbose=F)
sbs <- FindClusters(sbs, resolution = 0.5, cluster.name =  clu_name)
sbs <- RunUMAP(sbs, reduction = red_name, dims = 1:20, spread=1.3, reduction.name = umap_name)
```


```{r}
DimPlot(sbs, reduction = umap_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = umap_name,  group.by = "cond.time", cols=pal)
DimPlot(sbs, reduction = umap_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.umap",  group.by = clu_name, label=T, cols=clusters)
```

```{r}
sbs = RunTSNE(sbs, reduction = red_name, dims = 1:30, reduction.name = tsne_name)
```

```{r}
DimPlot(sbs, reduction = tsne_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = tsne_name,  group.by = "cond.time", cols=pal)
DimPlot(sbs, reduction = tsne_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.tsne",  group.by = clu_name, label=T, cols=clusters)
```


```{r}
sbs[["umap"]] = sbs[[umap_name]]
sbs[["tsne"]] = sbs[[tsne_name]]
```


## No cells
```{r no_cells}
ggplot(sbs@meta.data) + geom_bar(aes(x=cond.time, fill=cond.time)) + 
  scale_fill_manual(values=pal) + scale_y_continuous(expand=c(0,0), name="Cells")
```

```{r}
DimPlot(sbs, reduction = "tsne", group.by="initial_clusters", cols=  tvthemes::gravityFalls_pal()(13))
```


## Cluster composition

```{r cluster_composition}
composition = group_by(sbs@meta.data, cond.time,fastmnn_clusters, time, condition) %>%
  dplyr::count(name="n_cells") %>%
  arrange(desc(n_cells))
head(composition)
```


```{r cluster_comp_plots}
ggplot(composition) + geom_bar(stat="identity",aes(x=cond.time, y=n_cells, fill=fastmnn_clusters)) + 
   scale_fill_manual(values=clusters, name="BvW RPCA\nclusters")+ scale_y_continuous(expand=c(0,0)) +
  theme(axis.text.x=element_text(angle=15, hjust=1))
ggplot(composition) + geom_bar(stat="identity",aes(x=cond.time, y=n_cells, fill=fastmnn_clusters), position = "fill") +
  scale_fill_manual(values=clusters)+ scale_y_continuous(expand=c(0,0)) +
  theme(axis.text.x=element_text(angle=15, hjust=1))
ggplot(composition) + geom_bar(stat="identity",aes(fill=cond.time, y=n_cells, x=fastmnn_clusters)) +
  scale_fill_manual(values=pal)+ scale_y_continuous(expand=c(0,0)) +
  theme(axis.text.x=element_text(angle=15, hjust=1))
ggplot(composition) + geom_bar(stat="identity",aes(fill=cond.time, y=n_cells, x=fastmnn_clusters), position = "fill") +
   scale_fill_manual(values=pal) + scale_y_continuous(expand=c(0,0)) +
  theme(axis.text.x=element_text(angle=15, hjust=1))
```

## UMAP

```{r umap}
DimPlot(sbs, reduction="umap", label=T, shuffle = T, cols=clusters)
DimPlot(sbs, reduction="umap", split.by = "cond.time", ncol=3, shuffle = T, cols=clusters)
DimPlot(sbs, group.by="cond.time", cols=pal, shuffle = T)
```



## t-SNE
```{r tsne}
DimPlot(sbs, reduction = "tsne",shuffle=T, cols=clusters)
DimPlot(sbs, reduction="tsne", split.by = "cond.time", ncol=3, shuffle = T, cols=clusters)
DimPlot(sbs, reduction = "tsne", group.by= "cond.time", shuffle = T, cols=pal)
```


## Find biomarkers
Min.pct sets the minimum percentage of cells for a marker to be found in

Does densify, which increases speed at the cost of memory help?  
```{r densify}
sbs = JoinLayers(sbs)
norm.markers = FindAllMarkers(sbs, only.pos = T, min.pct=0.25, 
                              logfc.threshold = 0.5, densify=T)
norm.markers = arrange(norm.markers, p_val_adj, avg_log2FC)
head(norm.markers); dim(norm.markers)


top.markers = as.data.frame(norm.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC))
top.markers
```
### Dotplots

```{r}
time_per_cluster = group_by(composition, fastmnn_clusters, time) %>% summarise(n_cells = sum(n_cells)) %>% 
  slice_max(n=1, order_by = n_cells) %>% select(fastmnn_clusters, time)
time_per_cluster
top.markers = merge(top.markers, time_per_cluster, by.x="cluster", by.y="fastmnn_clusters")
head(top.markers)
```

 
 
```{r marker_expr}
DotPlot(sbs, features=rev(unique(top.markers$gene)), group.by = "fastmnn_clusters") & 
  scale_colour_gradientn(colours = spectral) & coord_flip()
```


```{r marker_dots}
top.markers$gene = factor(top.markers$gene, levels=rev(unique(top.markers$gene)))
overlap.to.plot = filter(top.markers, gene %in% top.markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(top.markers$gene))

ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point(colour="black",aes(size = avg_log2FC*1.5)) + geom_point() +
  scale_color_manual(values=clusters) + facet_grid(~time, scales="free_x", space="free_x")
```

```{r top_marker_genes, dev=c('png','pdf')}
include_all_plot = filter(norm.markers, gene %in% overlap.to.plot$gene & p_val_adj< 0.01)
include_all_plot$gene = factor(include_all_plot$gene, levels=levels(top.markers$gene))
include_all_plot = merge(include_all_plot, time_per_cluster, by.x="cluster", by.y="fastmnn_clusters")
a = ggplot(include_all_plot, aes(x=cluster, y=gene, colour=cluster, size=avg_log2FC)) + 
  geom_point(colour="black",aes(size = avg_log2FC*1.5)) + geom_point() +scale_color_manual(values=clusters)
a

a + facet_grid(~time, scales="free_x", space="free_x")
```

###  TSNEs

```{r marker_vis}
to_plot = grep("ACTA2|ADIPOQ$|r1@FASN|DCN", value=T, Features(sbs))

FeaturePlot(sbs, features =to_plot, reduction = "tsne") 

for (i in 1:(length(top.markers$gene)/4)){
  start = ((i-1)*4)+1; print(start)
  end = i*4; print(end)
  print(FeaturePlot(sbs, features =as.character(top.markers$gene[start:end], 
                                                reduction = "tsne")))
}

FeaturePlot(sbs, features = c("percent.mt", "nCount_RNA","nFeature_RNA"), reduction = "tsne")
FeaturePlot(sbs, features = c("PC_1","PC_2","PC_3","PC_4"), reduction = "tsne")
```

### heatmap

```{r}
norm.markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) -> top10
DoHeatmap(subset(sbs,downsample=100), features= top10$gene)
```



## Cluster GO

```{r load_go}
molsig <- clusterProfiler::read.gmt("/projects/imb-pkbphil/sp/annotations/molsigdb/msigdb.v2024.1.Hs.symbols.gmt")
head(molsig); nrow(molsig)
##Pick gene sets to look in bcos all is a lot
prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))
```

```{r}
norm.markers$cluster = factor(norm.markers$cluster, levels =0:11)
all.genes = rownames(sbs)
```


```{r basic_go}
go = compareCluster(gene~cluster, data=norm.markers, 
                      fun="enricher", TERM2GENE = some.molsig, universe = all.genes)
head(go)[,c(2,4:9)]
```

```{r go_replace}
phrases= read.delim(here("data/gene_sets/abbreviations.txt"))
recap= read.delim(here("data/gene_sets/capitalisations.txt"))
```

```{r abrev_go}
drama_shorten = function(ont) {
  for (x in 1:nrow(phrases)){
    ont = gsub( phrases$phrase[x],phrases$abbreviation[x], ont, ignore.case = T)
  }
  a = abbreviate(gsub("_"," ", gsub("^[[:alpha:]]*_","",tolower(ont))),minlength=41, dot=T, named = F)
  a = stringr::str_to_sentence(a)
  for (i in 1:nrow(recap))
    a = gsub(sprintf("\\b%s\\b", recap$lowercase[i]),recap$capital[i],a, ignore.case = T)
  return(a)
}
```


```{r default_go_plots}
go = mutate(go, cluster = factor(cluster, levels=0:max(as.integer(cluster))))
head(go)[,c(2,4:9)]

p = dotplot(go, x="cluster", showCategory=2, includeAll=F) 
p+ scale_y_discrete(labels=drama_shorten(levels(p$data$Description)))  + theme_hp
```


```{r set_up_go_plot}
dfgo = as.data.frame(go)
dfgo$short_desc= drama_shorten(dfgo$Description)
dfgo = merge(dfgo, time_per_cluster,by.x="cluster", by.y="fastmnn_clusters")
go_to_plot = filter(dfgo, p.adjust < 0.01) %>% group_by(Cluster) %>%
              slice_min(n = 2, order_by = data.frame(p.adjust, pvalue))
go_to_plot$short_desc = factor(go_to_plot$short_desc, levels=rev(unique(go_to_plot$short_desc)))
```


```{r cluster_go, dev=c('png','pdf')}
p= ggplot(go_to_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
   geom_point(colour="black",aes(size = Count+(Count*1.5))) + geom_point() +scale_color_manual(values=clusters)+ theme(axis.title.y = element_blank())
p
```

```{r cluster_go_overlapping, dev=c('png','pdf')}
include_all_plot = filter(dfgo, Description %in% go_to_plot$Description & p.adjust < 0.01)
include_all_plot$short_desc = factor(include_all_plot$short_desc, levels=levels(go_to_plot$short_desc))
#include_all_plot = merge(include_all_plot, time_per_cluster,by.x="cluster", by.y="seurat_clusters")
a = ggplot(include_all_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
  geom_point(colour="black",aes(size = Count*2)) + geom_point()+
  theme(axis.text.x=element_text(angle=30, hjust=1)) +scale_color_manual(values=clusters)
a
a + facet_grid(~time, scales="free_x", space = "free_x")
```

## Add Progenitor clusters 

```{r}
prog_cols = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0")
pro_clusters = "output/progenitors/rpca"
if (subset_data){
  pro = readRDS(here(pro_clusters, "10%_complete_analysis.rds"))
} else{
  pro = readRDS(here(pro_clusters, "complete_analysis.rds"))
}
pro
table(pro$rpca_clusters)
head(pro$rpca_clusters)
tail(pro$rpca_clusters)
```


```{r}
pro_clusters = pro$rpca_clusters
names(pro_clusters) = gsub("day0.subjectS6","asc.day0",names(pro_clusters))
head(pro_clusters[rownames(sbs@meta.data)])
sbs@meta.data$progen_clusters = pro_clusters[rownames(sbs@meta.data)]
table(sbs$progen_clusters, useNA="ifany")
```


```{r}
sbs$if_progen_cluster = paste0("t",sbs$initial_clusters)
sbs$if_progen_cluster[!is.na(sbs$progen_clusters)] = paste0("p",sbs$progen_clusters[!is.na(sbs$progen_clusters)])
un.pc = unique(sbs$if_progen_cluster)
sbs$if_progen_cluster = factor(sbs$if_progen_cluster , 
                               levels=un.pc[order(gsub("[[:digit:]]*$","",un.pc),
                                                  as.integer(gsub("p|t","", un.pc)))] )
table(sbs$if_progen_cluster, useNA="ifany")
```

```{r}
sbs$time.progen_clusters = sbs$time
sbs$time.progen_clusters[!is.na(sbs$progen_clusters)] = paste0("p",sbs$progen_clusters[!is.na(sbs$progen_clusters)])
sbs$time.progen_clusters = factor(sbs$time.progen_clusters,
                                  levels=c(paste0("p",0:7), "day0","day1","day3")) #approx pseuodtime ordering
table(sbs$time.progen_clusters)
rm(pro)
```

### Progen plots

```{r}
DimPlot(sbs, reduction="umap", group.by= "time.progen_clusters", label=F, shuffle = T, 
        cols=c(prog_cols,pal[c(3,2,1)]))
DimPlot(sbs, reduction="umap", group.by= "time.progen_clusters", label=F, shuffle = T, 
        cols=c(prog_cols,pal[c(3,2,1)]), split.by = "time", ncol = 2)
DimPlot(sbs, reduction="umap", group.by= "time.progen_clusters", label=F, shuffle = T, 
        cols=c(prog_cols,pal[c(3,2,1)]), split.by = "time.progen_clusters", ncol = 3)
```

```{r prog_tsne}
DimPlot(sbs, reduction="tsne", group.by= "time.progen_clusters", label=F, shuffle = T, 
        cols=c(prog_cols,pal[c(3,2,1)]))
DimPlot(sbs, reduction="tsne", group.by= "time.progen_clusters", label=F, shuffle = T, 
        cols=c(prog_cols,pal[c(3,2,1)]), split.by = "time", ncol = 2)
DimPlot(sbs, reduction="tsne", group.by= "time.progen_clusters", label=F, shuffle = T, 
        cols=c(prog_cols,pal[c(3,2,1)]), split.by = "time.progen_clusters", ncol = 3)
```


## Save



```{r save}
if (subset_data == F){
  saveRDS(sbs, file = here(odir, "complete_analysis.rds"))
   write.table(composition, here(odir, "cluster_composition.tsv"),
            sep="\t", row.names = F, quote=F)
  write.table(norm.markers, here(odir, "marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(go, here(odir, "ORA_marker_genes.txt"),
            sep="\t", row.names = F)
  Idents(sbs) = "cond.time"
  dwns = subset(sbs, downsample=1000)
  saveRDS(dwns, file = here(odir, "10%_complete_analysis.rds"))
} else if (subset_data){
  saveRDS(sbs, file = here(odir, "10%_subset_analysis.rds"))
}
```


