---
title: "tss_integration_trial"
output: html_document
---


Integrate time points using RPCA, FastMNN and Harmony.  We're not integrating out differences between white and beige cells.  We're trying to bring the day0 cells close enough to the day1 and day3 cells (in UMAP or TSNE space reduced dimensionality space) that we can estimate cellular differentiation trajectories.  


```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(SeuratWrappers)
library(dplyr)
library(ggplot2)
library(ggsci)
library(here)
library(patchwork)
```

```{r params}
module = "beige_vs_white"
dataname = "tss_integration_trial"
load_data = T #loading data
subset_data = F

rmd_filename = "tss_integration_trial"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)
indir = here("output", module, "tss_bvw_initial")
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4", "#85C2EA","#B3B3B3", "#F18486","#E31A1C")
clusters = tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```

## Read data

```{r}
if (subset_data){
  sbs = readRDS(here(indir, "10%_complete_analysis.rds"))
  Idents(sbs) = "seurat_clusters"
} else {
  sbs = readRDS(here(indir, "complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
sbs[["RNA"]] =split(sbs[["RNA"]], f = sbs$time)
sbs = FindVariableFeatures(sbs, nfeatures = 2000)
sbs
```


```{r}
sbs$initial_clusters = sbs$seurat_clusters
sbs[["initial.umap"]] = sbs[["umap"]]
sbs[["initial.tsne"]] = sbs[["tsne"]]
```

```{r}
DimPlot(sbs, cols=clusters, label=T)
```


## RPCA Integration

```{r setup-variables-rpca}
method_name="rpca"
red_name = paste("pca", sep=".", method_name)
umap_name = paste("umap", sep=".", method_name)
clu_name = paste( method_name, sep="_", "clusters")
tsne_name = paste("tsne", sep=".", method_name)
```

```{r}
sbs = IntegrateLayers(sbs, method= RPCAIntegration, orig.reduction = "pca",
                       new.reduction = red_name, 
                      assay = "RNA", scale.layer="scale.data", layers="data"
                      )
```

```{r basic_analysis}
sbs <- FindNeighbors(sbs, reduction =  red_name, dims = 1:30, verbose=F)
sbs <- FindClusters(sbs, resolution = 0.5, cluster.name =  clu_name)
sbs <- RunUMAP(sbs, reduction = red_name, dims = 1:20, spread = 1.3, reduction.name = umap_name)
```

```{r}
DimPlot(sbs, reduction = umap_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = umap_name,  group.by = "orig.ident", cols=pal)
DimPlot(sbs, reduction = umap_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.umap",  group.by = clu_name, label=T, cols=clusters)
```

```{r}
sbs = RunTSNE(sbs, reduction = red_name, dims = 1:30, reduction.name = tsne_name)
```

```{r}
DimPlot(sbs, reduction = tsne_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = tsne_name,  group.by = "orig.ident", cols=pal)
DimPlot(sbs, reduction = tsne_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.tsne",  group.by = clu_name, label=T, cols=clusters)
```

## Harmony Integration

```{r setup-variables-harmony}
method_name="harmony"
red_name = paste("pca", sep=".", method_name)
umap_name = paste("umap", sep=".", method_name)
clu_name = paste( method_name, sep="_", "clusters")
tsne_name = paste("tsne", sep=".", method_name)
```

```{r}
sbs = IntegrateLayers(sbs, method= HarmonyIntegration, orig.reduction = "pca",
                       new.reduction = red_name)
```

```{r}
sbs <- FindNeighbors(sbs, reduction =  red_name, dims = 1:30, verbose=F)
sbs <- FindClusters(sbs, resolution = 0.5, cluster.name =  clu_name)
sbs <- RunUMAP(sbs, reduction = red_name, dims = 1:20, spread = 1.3, reduction.name = umap_name)
```


```{r}
DimPlot(sbs, reduction = umap_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = umap_name,  group.by = "orig.ident", cols=pal)
DimPlot(sbs, reduction = umap_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.umap",  group.by = clu_name, label=T, cols=clusters)
```

```{r}
sbs = RunTSNE(sbs, reduction = red_name, dims = 1:30, reduction.name = tsne_name)
```

```{r}
DimPlot(sbs, reduction = tsne_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = tsne_name,  group.by = "orig.ident", cols=pal)
DimPlot(sbs, reduction = tsne_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.tsne",  group.by = clu_name, label=T, cols=clusters)
```

## FastMNN

```{r setup-variables-fastmnn}
method_name="fastmnn"
red_name = paste("pca", sep=".", method_name)
umap_name = paste("umap", sep=".", method_name)
clu_name = paste( method_name, sep="_", "clusters")
tsne_name = paste("tsne", sep=".", method_name)
```

```{r}
sbs = IntegrateLayers(sbs, method= FastMNNIntegration, orig.reduction = "pca",
                       new.reduction = red_name)
```

```{r}
sbs <- FindNeighbors(sbs, reduction =  red_name, dims = 1:30, verbose=F)
sbs <- FindClusters(sbs, resolution = 0.5, cluster.name =  clu_name)
sbs <- RunUMAP(sbs, reduction = red_name, dims = 1:20, spread = 1.3, reduction.name = umap_name)
```


```{r}
DimPlot(sbs, reduction = umap_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = umap_name,  group.by = "orig.ident", cols=pal)
DimPlot(sbs, reduction = umap_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.umap",  group.by = clu_name, label=T, cols=clusters)
```

```{r}
sbs = RunTSNE(sbs, reduction = red_name, dims = 1:30, reduction.name = tsne_name)
```

```{r}
DimPlot(sbs, reduction = tsne_name,  group.by = clu_name, label=T, cols=clusters)
DimPlot(sbs, reduction = tsne_name,  group.by = "orig.ident", cols=pal)
DimPlot(sbs, reduction = tsne_name,  group.by = "initial_clusters", label=T, cols=clusters)
DimPlot(sbs, reduction = "initial.tsne",  group.by = clu_name, label=T, cols=clusters)
```

```{r save}
if (subset_data == FALSE){
  saveRDS(sbs, file = here(odir, "complete_analysis.rds"))
  sub = subset(sbs, downsample=1000)
  saveRDS(sub,here(odir, "10%_complete_analysis.rds"))
} else if (subset_data){ #if subset, don't save complete analysis
  saveRDS(sbs,here(odir, "10%_subset_analysis.rds"))
}
```
