---
title: "DE_per_time"
author: "Sarah Hazell Pickering"
date: "2025-02-11"
output: html_document
---

TSS analysis bvw initial

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(RColorBrewer)
library(Seurat)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(ggrepel)
library(ggsci)
library(here)
library(tidyr)
```

```{r params}
module = "beige_vs_white"
dataname = "tss_bvw_initial"
load_data = T #loading data
subset_data = F

rmd_filename = "tss_bvw_initial_DE_per_time"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)

```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4","#85C2EA","#B3B3B3", "#F18486","#E31A1C")
clusters = #ggsci::pal_simpsons()(13)[c(1:7,9:13)]
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```


```{r params}
module="adipogenesis"
dataname = "tss_bvw_initial"
library(here); i_am(file.path("analysis", paste0(dataname,".Rmd")))

odir = here("output", module, dataname) 

load_data = T #loading data 
subset_data = F
```

```{r libs, warning=F, message=F, echo=F}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)
library(ggrepel)
library(Seurat)
library(BPCells)
library(RColorBrewer)
library(clusterProfiler)
library(enrichplot)
```

```{r}
pal = c("#1F78B4","#85C2EA","#B3B3B3", "#F18486","#E31A1C")
clusters = #ggsci::pal_simpsons()(13)[c(1:7,9:13)]
  tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
#define new theme
theme_hp = theme_classic(base_size=15) + 
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25),
        strip.background = element_blank(), # remove boxes around facet labels
        )
#set for all ggplots
theme_set(theme_hp)
knitr::opts_chunk$set(dev=c("png","pdf"))
```


## Read data

```{r}
if (subset_data){
  sbs = readRDS(here(odir, "10%_complete_analysis.rds"))
} else {
  sbs = readRDS(here(odir, "complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
sbs$cond.time = factor(sbs$cond.time, levels=c("white.day3","white.day1","asc.day0","beige.day1","beige.day3"))
Idents(sbs) = "cond.time"
sbs
table(Idents(sbs))
```

```{r}
DimPlot(sbs, cols= pal, shuffle=T)
```

Formatting for TSS ids so we can show them in a shorter form on the volcano plot for e.g. 
gene_name is used on the plot.  
MT loci have multple genes per TSS so I removed the additional r1@MT- and just show additional MT gene names with commas. 
  e.g. P:r1@MT-ND2,r1@MT-CO1,r1@MT-TI,r1@MT-TM,r1@MT-TW is shortned to P:r1@MT-ND2,CO1,TI,TM,TW

```{r}
format_tss_ids = function(tss_id){
  df = data.frame(tss_id=tss_id)
  df =  mutate(df, 
         simple_id = gsub("r[0-9]*@ADDG[0-9M]*\\.[A-Z],","",
                        gsub(",r[0-9]*@ADDG[0-9M]*\\.[A-Z]$","",tss_id)),
         feature_name = if_else(grepl("@MT-.*,r[0-9]*@MT-",tss_id), 
                             gsub(",r1@MT-",",", simple_id),#compress MT loci
                              gsub("@ADDG.{8}","@..", simple_id)),
         #for GO analysis need gene name; pick first, unless it has an obvious no symbol
         best_gene_name = gsub(",.*","",gsub("[PD]:r[0-9]*@((MIR|AL|LINC|FP)[0-9\\.-]*,r[0-9]*@)?","", feature_name)),
         gene_names = gsub("([DP]:)?r[0-9]*@", "", simple_id),
         proximity = gsub(":r[0-9]*@.*", "", simple_id)
         )
  return (df)
}
all.genes = format_tss_ids(row.names(sbs)) 
head(all.genes)
#test cases
all.genes[grep("LINC00339", all.genes$tss_id),]
all.genes[grep("MT-CO2", all.genes$tss_id),]
all.genes[grep("MTCO", all.genes$tss_id),]
```

Adding some sorting information for DE analysis; the condition, the tss_ids from above and sorting the output by average logFC (with significant hits at the top).  

```{r}
de_formatting = function(df){
  df = bind_cols(df, format_tss_ids(rownames(df)))
   df = mutate(df, 
          sig=if_else(p_val_adj < 0.01 & 
                (pct.1 + pct.2) > 0.2 & abs(avg_log2FC) > 0.25,
                "sig","not_sig"), 
         condition = if_else(avg_log2FC > 0, "beige", "white"),
         sig_cond = if_else(sig == "not_sig", "not_sig",
                            if_else(avg_log2FC > 0, "beige", "white")
                            ))%>%
  arrange(sig == "not_sig", desc(abs(avg_log2FC)))
   return(df)
}
```


## Day3
```{r}
day3_test = FindMarkers(sbs, ident.1="beige.day3",ident.2="white.day3")
day3_test = de_formatting(day3_test)
table(day3_test$sig_cond)
head(day3_test)
```


```{r day3_volcano_all}
ggplot(day3_test, aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = slice_min(day3_test, n=20, order_by = p_val), aes(label=feature_name), show.legend = F) + 
  scale_colour_manual(values=c(pal[5],"black",pal[1])) + guides(colour = guide_legend(override.aes = list(size = 2)))  +
  ggtitle("day3") + coord_cartesian(x=c(-1,1))
```


### Show only proximal elements

```{r day3_volcano_proximal}
ggplot(filter(day3_test, proximity=="P"), aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = filter(day3_test, proximity=="P" ) %>% slice_min(n=10, order_by = p_val), 
                  aes(label=feature_name), show.legend = F) + 
  scale_colour_manual(values=c(pal[5],"black",pal[1])) + guides(colour = guide_legend(override.aes = list(size = 2)))  +
  ggtitle("day3") + coord_cartesian(x=c(-1,1))

ggplot(filter(day3_test, proximity=="P"), aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = filter(day3_test, proximity=="P" ) %>% slice_min(n=20, order_by = p_val), 
                  aes(label=feature_name), show.legend = F) + 
  scale_colour_manual(values=c(pal[5],"black",pal[1])) + guides(colour = guide_legend(override.aes = list(size = 2)))  +
  ggtitle("day3") + coord_cartesian(x=c(-1,1))

```



## day 1
```{r}
day1_test = FindMarkers(sbs, ident.1="beige.day1",ident.2="white.day1")
day1_test = de_formatting(day1_test)
table(day1_test$sig_cond)
head(day1_test, n=20)
```



```{r day1_volcano}
ggplot(day1_test, aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = slice_min(day1_test, n=15, order_by = p_val), aes(label=feature_name)) + 
  scale_colour_manual(values=c(pal[4],"black",pal[2]))+ guides(colour = guide_legend(override.aes = list(size = 2))) +
  ggtitle("day1")+ coord_cartesian(x=c(-2,2))

ggplot(day1_test, aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = slice_min(day1_test, n=30, order_by = p_val), aes(label=feature_name)) + 
  scale_colour_manual(values=c(pal[4],"black",pal[2])) + guides(colour = guide_legend(override.aes = list(size = 2))) +
  ggtitle("day1 zoom") + coord_cartesian(x=c(-1,1), y=c(100, 150))

ggplot(day1_test, aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = slice_min(day1_test, n=70, order_by = p_val), aes(label=feature_name)) + 
  scale_colour_manual(values=c(pal[4],"black",pal[2])) + guides(colour = guide_legend(override.aes = list(size = 2))) +
  ggtitle("day1 zoom") + coord_cartesian(x=c(-1,1), y=c(0, 100))
```
### only proximal 

```{r day1_volcano_proximal}
ggplot(filter(day1_test, proximity=="P"), aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = filter(day1_test, proximity=="P" & sig == "sig") %>% slice_min(n=10, order_by = p_val), 
                  aes(label=feature_name), show.legend = F) + 
  scale_colour_manual(values=c(pal[4],"black",pal[2])) + guides(colour = guide_legend(override.aes = list(size = 2)))  +
  ggtitle("day1") + coord_cartesian(x=c(-1,1))

ggplot(filter(day1_test, proximity=="P"), aes(x=avg_log2FC, y=-log10(p_val), colour=sig_cond)) + geom_point(size=0.5) + 
  geom_text_repel(data = filter(day1_test, proximity=="P" & sig == "sig") %>% slice_min(n=30, order_by = p_val), 
                  aes(label=feature_name), show.legend = F) + 
  scale_colour_manual(values=c(pal[4],"black",pal[2])) + guides(colour = guide_legend(override.aes = list(size = 2)))  +
  ggtitle("day1") + coord_cartesian(x=c(-1,1))
```

```{r}
filter(day1_test, sig_cond=="beige" & proximity=="P") %>% pull(best_gene_name) %>% length()
```

## Quick GO 


```{r}
molsig <- clusterProfiler::read.gmt("/projects/imb-pkbphil/sp/annotations/molsigdb/msigdb.v2024.1.Hs.symbols.gmt")
head(molsig); nrow(molsig)

prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))
rm(molsig)
```


```{r}

go_day1 = enricher(filter(day1_test, sig_cond=="beige") %>% pull(best_gene_name),
               universe = all.genes$best_gene_name,
            TERM2GENE = some.molsig)
go_day1
#dotplot(go_day1)
```



```{r}
go_day3 = enricher(filter(day3_test, sig_cond=="beige") %>% pull(best_gene_name),
               universe = all.genes$best_gene_name,
            TERM2GENE = some.molsig)
head(go_day3, n=10)[c(3:6,8:9)]
dotplot(go_day3, showCategory=8) + theme_hp
```



## Save tables
```{r}
if (subset_data == F){
  write.table(filter(day3_test, p_val_adj < 0.01), here(odir, "DE_day3_wvb.tsv"),
              sep="\t",row.names = F, quote = F)
  write.table(filter(day1_test, p_val_adj < 0.01), here(odir, "DE_day1_wvb.tsv"),
              sep="\t",row.names = F, quote = F)
}
```

