---
title: "Donor-specific progentor markers"
output: html_document
date: "2024-07-18"
---

```{r}
module="progenitors"
dataname = "rpca"
library(here);i_am(file.path("analysis",module,"donor-specific_progenitor_markers.Rmd"))
odir = here("output", module, dataname) 

load_data = T #Loading initial seurat analysis
subset_data =  F
```

```{r libs, warning=F, message=F, echo=T}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)

library(Seurat)
library(BPCells)

library(clusterProfiler)
library(enrichplot)
```

```{r themes, warning=F, message=F, echo=T}
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters =  c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0")
spectral =  rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))

#define new theme
theme_hp = theme_classic(base_size=15) + 
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25),
        strip.background = element_blank(), # remove boxes around facet labels
        )
		#set for all ggplots
theme_set(theme_hp)
knitr::opts_chunk$set(dev=c("png","pdf"))
```

```{r}
coi = c(0,1,2,7)
```

## Load RPCA cluster markers

```{r}
norm.markers = read.delim(here(odir, "marker_genes.txt"))
norm.markers = arrange(norm.markers, cluster, desc(avg_log2FC))
head(norm.markers)
table(norm.markers$cluster)
```
So to start with here are the DE genes per cluster


## Some basic filtering

```{r}
nonclusters = filter(norm.markers, !cluster %in% coi) %>% pull(gene)
unique = filter(norm.markers, cluster %in% coi & !gene %in% nonclusters)
nrow(unique); table(unique$cluster)
head(unique)
```

Removing markers that overlap with other clusters (any amount); Still many genes left.  


## Load seurat object; basic UMAP
```{r}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
DimPlot(filt, cols = clusters)
```


## Load manually selected markers

```{r}
selected = readxl::read_xlsx(here("data/selected_markers/Progenitor_markers.xlsx"))
head(selected)
```

## Cluster specificity? 
```{r}
for (clu in unique(selected$cluster)){
  print(clu)
  features = selected$Gene[selected$cluster == clu]
  if (length(features) > 1){
      v =VlnPlot(filt, features, stack = T, flip = T, combine=F, split.by = "seurat_clusters", cols=clusters)
    print(v)
  } else{
    v =VlnPlot(filt, features,  split.by = "seurat_clusters", cols=clusters)
    print(v)
  }
}
```
### Consistency across subjects? 
```{r}
for (clu in unique(selected$cluster)){
  print(clu)
  features = selected$Gene[selected$cluster == clu]
  if (length(features) > 1){
      v =VlnPlot(filt, features, stack = T, flip = T, combine=F, split.by = "subject", cols=pal)
    print(v)
  } else{
    v =VlnPlot(filt, features,  split.by = "subject", cols=pal)
    print(v)
  }
}
```


### Subject specificity
```{r}
for (clu in unique(selected$cluster)){
  print(clu)
  features = selected$Gene[selected$cluster == clu]
  if (length(features) > 1){
      v =VlnPlot(filt, features, stack = T, flip = T, combine=F, group.by="subject", split.by = "subject", cols=pal)
    print(v)
  } else{
    v =VlnPlot(filt, features, group.by="subject",  split.by = "subject", cols=pal)
    print(v)
  }
}
```


```{r}
for (gene  in selected$Gene){
  p = FeaturePlot(filt, gene) + scale_colour_gradientn(colours = spectral)
  print(p)
}
```


```{r}
FeaturePlot(filt, "CLEC3B") + scale_colour_gradientn(colours = spectral)
FeaturePlot(filt, "AKAP12") + scale_colour_gradientn(colours = spectral)
```
## Using unintegrated data

```{r}
for (gene  in selected$Gene){
  p = FeaturePlot(filt, gene, reduction = "initial.umap") + scale_colour_gradientn(colours = spectral)
  print(p)
}
```


## Load GO terms

```{r}
molsig <- clusterProfiler::read.gmt("/projects/imb-pkbphil/sp/annotations/molsigdb/msigdb.v2024.1.Hs.symbols.gmt")
head(molsig); nrow(molsig)
grep("EXTRACELLULAR_MATRIX", unique(molsig$term), value=T)

colnames(molsig) = c("term","gene")
some.molsig = molsig[molsig$term %in% c("GOCC_CELL_SURFACE","GOCC_COLLAGEN_CONTAINING_EXTRACELLULAR_MATRIX"),]
some.molsig$term = factor(some.molsig$term)
table(some.molsig$term)
rm(molsig)
```
## Surface markers 
( just based on molsigDB GO terms; which is not very inclusive)
Cluster 0 has very few unique surface markers.  

```{r}
surface = filter(unique, gene %in% some.molsig$gene)
table(surface$cluster)
surface
```




```{r}
write.table(surface, here(odir, "surface_markers_progenitor_clusters.tsv"), sep="\t", row.names = F, quote = F)
```

