---
title: "Progenitor seurat qc and analysis"
output: html_document
---

previous code: mito_on_mad_take2_full
now updating seuratv5

```{r params}
module = "progenitors"
dataname = "initial"#<- to load / to save -> dataname = "regress_out"
library(here); i_am(file.path("analysis",paste(module,sep="_",dataname,"seurat_analysis.Rmd")))
indir = here("data/cellranger/count_no_introns/day0")
odir = here("output", module, dataname) 


if (!dir.exists(odir)){dir.create(odir, recursive = T)}
load_data = T
subset_data = F #subset data to 10% total cells?
```


```{r libs, message=FALSE, warning=FALSE, echo=T}
library(dplyr, quietly = T)
library(ggplot2)
library(ggsci)

library(Seurat)
library(BPCells)
library(patchwork)
library(clusterProfiler, quietly = T)
library(enrichplot)
```


```{r colors}
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
library(RColorBrewer)
library(ggsci)
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
spectral =  rev(brewer.pal(n = 11, name = "Spectral"))

#clusters = c("#FED439FF", "#709AE1FF", "#8A9197FF", "#D2AF81FF",
 #            "#FD7446FF", "#D5E4A2FF", "#197EC0FF", "#46732EFF", "#71D0F5FF")
```

## Read data

This filtered matrix has been filtered for cells with low UMI counts already (EmptyDrops method).  
If first run, save bpcells objects, otherwise load data.  Can work on a subset if subset_data=T

```{r read_or_load}
samples = list.dirs(here(indir), full.names = F, recursive = F)
samples

if (load_data==FALSE){# load count matrices
  
  data_list = c()
  
  for (s in samples){
    #if this has been run before delete bp dir
    bc_matrix = Read10X(data.dir = here(indir, s, "outs/filtered_feature_bc_matrix"))
    so = CreateSeuratObject(counts=bc_matrix, project=s, names.field = 2, names.delim = "-")
    write_matrix_dir(mat=so[["RNA"]]$counts, dir = here(odir, "bpcells", s))
    gene_mat = open_matrix_dir(here(odir, "bpcells", s))
    so[["RNA"]]$counts = gene_mat
    
    data_list = c(data_list,so)
  }
  names(data_list) = samples
  aggr = merge(data_list[[1]], y= data_list[2:length(data_list)], 
               add.cell.ids = gsub("_",".",samples), project=dataname )
  
  saveRDS(aggr,here(odir,"unfiltered_object.rds"))

  
  rm(data_list, bc_matrix, so)
} else if(load_data){
  aggr = readRDS(here(odir, "unfiltered_object.rds"))
}

aggr
head(aggr@meta.data) #auto metadata
```


```{r}
dataname = "regess_out"
odir = here("output", module, dataname) 
if (!dir.exists(odir)){dir.create(odir, recursive = T)}
```


```{r}
aggr$subject = gsub("_.*","", gsub("day0.subject","",colnames(aggr)))
Idents(aggr) = "subject"
aggr$orig.ident = aggr$subject
table(Idents(aggr))
aggr <- JoinLayers(aggr)
aggr
```

## Quality control

Number of genes (features) detected per cell high, due to the depth of our sequencing and the stem-cell nature of the samples.  There is a wide spread of the number of genes (nFeature)and reads (nCount) per cell.  The percentage mitochondrial genes is super low for most cells, but some few have up to 80% mt.   


```{r before_filt1, dev=c("png","pdf")}
aggr[["percent.mt"]] = PercentageFeatureSet(aggr, pattern="^MT-")
VlnPlot(aggr, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0, cols=pal)
```


```{r before_filt2}
VlnPlot(subset(x = aggr, downsample = 1000), features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, cols=pal)
RidgePlot(aggr, feature= "nFeature_RNA", cols=pal) + geom_vline(aes(xintercept=2000))
```

```{r before_filt3}
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "percent.mt", cols=pal)
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", cols=pal)
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal)
```
We set an adaptive Mt.percent filter that's the 3* mad maximum. 
```{r}
mt_info = group_by(aggr@meta.data, subject) %>% 
  summarise(mad = mad(percent.mt),
            median = median(percent.mt), 
            thresh = median(percent.mt) + mad(percent.mt)*3,
            th95 = quantile(percent.mt, 0.95))
mt_info
thresh = max(mt_info$thresh)
```
```{r}
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal) + 
  geom_hline(aes(yintercept=thresh), linetype="dashed") +
  geom_vline(aes(xintercept=2000), linetype="dashed")
```


```{r filter}
ncol(aggr)
filt <- subset(aggr,  percent.mt < thresh & nFeature_RNA > 2000)
ncol(filt)

FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size=0.5, cols=pal)
FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size=0.5, cols=pal)
FeatureScatter(filt, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal) 
```


## Normalise

```{r normalise}
filt= NormalizeData(filt, normalization.method = "LogNormalize", scale.factor = 10000)
```

## Identification of Variable Features

```{r variable_features}
filt <- FindVariableFeatures(filt, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
prev_var_feat = VariableFeatures(filt)
top10 <- head(VariableFeatures(filt), 10)

# plot variable features with and without labels
plot1 = VariableFeaturePlot(subset(filt, downsample=1000))
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

## Scaling

Centres around zero and sets the variance to 1 - basically a z-score transformation.  Here we use all genes, but by default Seurat uses only identified variable features.

```{r scaling}
all.genes = rownames(filt)
filt = ScaleData(filt, features=all.genes, vars.to.regress=c("nCount_RNA", "percent.mt"))
filt = FindVariableFeatures(filt)
new_var_feat = VariableFeatures(filt)
summary(new_var_feat == prev_var_feat)
```

## PCA

PC1 is separating a small number of cells from the main group.  MKI67 is very much a prolif/cell cycle gene.  

```{r pca}
filt = RunPCA(filt)
print(filt[["pca"]], dims = 1:5, nfeatures = 5)
FeaturePlot(filt, reduction = "pca", feature="MKI67", cols=c("lightgrey","red"))
FeaturePlot(filt, reduction = "pca", feature="nCount_RNA", cols=c("lightgrey","red"))
DimPlot(filt, reduction="pca", shuffle=T, split.by = "orig.ident", ncol=3)

FeaturePlot(filt, reduction = "pca", feature="nCount_RNA", cols=c("lightgrey","red"), dims = c(3,4))
```


## Determine the dimensionality

```{r elbow}
ElbowPlot(filt, ndims = 50)
```

## Clustering

```{r clust}
filt <- FindNeighbors(filt, dims = 1:20)
filt <- FindClusters(filt, resolution = 0.5)
```

```{r}
table(Idents(filt))
VlnPlot(filt, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0, cols=clusters)
```


### Cluster Composition
```{r}
composition = group_by(filt@meta.data, subject,seurat_clusters) %>% count()
composition = composition[order(composition$n, decreasing = T),]
colnames(composition) = c("donor_id","seurat_clusters","n_cells")
head(composition)
```


```{r donor_composition, dev=c('png','pdf')}
ggplot(composition) + geom_bar(stat="identity",aes(y=donor_id, x=n_cells, fill=seurat_clusters)) + 
  scale_y_discrete(limits=rev)+ scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15) + scale_fill_manual(values=clusters)
ggplot(composition) + geom_bar(stat="identity",aes(y=donor_id, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15)  + scale_fill_manual(values=clusters)
```

```{r cluster_composition, dev=c('png','pdf')}
ggplot(composition) + geom_bar(stat="identity",aes(fill=donor_id, x=n_cells, y=seurat_clusters)) +
    scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)  + scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15)
ggplot(composition) + geom_bar(stat="identity",aes(fill=donor_id, x=n_cells, y=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)  + scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15)
```



## Non-linear dimensionality reduction

```{r umap, dev=c('png','pdf')}
filt = RunUMAP(filt, dims=1:20)
DimPlot(filt, reduction="umap", label=T, cols=clusters)
DimPlot(filt, reduction="umap", group.by = "subject", label=T, cols=pal, label.color = "white", label.box=T)
DimPlot(filt, reduction="umap", split.by = "subject",group.by = "seurat_clusters", ncol=3, cols = clusters)
FeaturePlot(filt, reduction = "umap",  features=c("nCount_RNA", "percent.mt")) & scale_colour_gradientn(colours = spectral)
```

## Integrate

```{r setup-variables-rpca}
method_name="rpca"
red_name = paste("integrated", sep=".", method_name)
umap_name = paste("umap", sep=".", method_name)
clu_name = paste( method_name, sep="_", "clusters")
```

```{r}
filt[["RNA"]] <- split(filt[["RNA"]], f = filt$orig.ident)
filt = IntegrateLayers(filt, method= RPCAIntegration, orig.reduction = "pca",
                       new.reduction = red_name)
```

```{r}
filt <- FindNeighbors(filt, reduction =  red_name, dims = 1:20, verbose=F)
filt <- FindClusters(filt, resolution = 0.5, cluster.name =  clu_name)
filt <- RunUMAP(filt, reduction = red_name, dims = 1:20, reduction.name = umap_name)
```

```{r}
DimPlot(filt, reduction = umap_name,  group.by = clu_name, label=T,cols=clusters)
FeaturePlot(filt, reduction = umap_name,  features=c("nCount_RNA", "percent.mt")) & scale_colour_gradientn(colours = spectral)
DimPlot(filt, reduction = umap_name,  group.by = "orig.ident", label=T, cols=pal, label.color = "white", label.box = T )
```


### Save 

```{r save} 
if (subset_data == FALSE){
  saveRDS(filt, file = here(odir, "complete_analysis.rds"))
  write.table(composition, here(odir, "cluster_composition.txt"),
              sep="\t", row.names = F)
  Idents(filt) = "orig.ident"
  sub = subset(filt, downsample=1000)
  Idents(filt) = "seurat_clusters"
  saveRDS(sub,here(odir,"10%_complete_analysis.rds"))
} else if (subset_data){ #if subset, don't save complete analysis
  saveRDS(filt,here(odir, "10%_subset_analysis.rds"))
  write.table(cluster.markers, file.path(odir,dataname, "10%_subset_marker_genes.txt"),
            sep="\t", row.names = F)
}
```

