---
title: "progenitors_rpca_cluster_stability"
author: "Sarah Hazell Pickering"
date: "2024-08-16"
output: html_document
---

Exploring cluster stability with clustree on resolutions 0.1 -> 1.2 and sch posthoc heirarchical clustering and statistcial testing

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(clustree)
library(dplyr)
library(ggplot2)
library(ggsci)
library(here)
library(tidyr)
library(scSHC)
```

```{r params}
module = "progenitors"
dataname = "integration_trial"
load_data = T #Loading initial seurat analysis
subset_data = F

rmd_filename = "progenitors_rpca_clustree"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0", tvthemes::gravityFalls_pal()(12))
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```

## load data

```{r}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
table(filt$rpca_clusters)
```

```{r}
filt[["mnn.reconstructed"]] =NULL
filt = JoinLayers(filt)
Idents(filt) = "rpca_clusters"
filt
head(filt@meta.data) 
table(filt$rpca_clusters)

red_to_rm = grep("harmony|fastmnn", names(filt@reductions), value = T)
for (red in red_to_rm){
  print(paste0("red removed ", red))
  filt[[red]] = NULL
}
#set default reductions
filt[["umap"]] = filt[["umap.rpca"]]
filt[["tsne"]] = filt[["rpca.tsne"]]
filt
```

## Testing different resolutions

```{r}
res_to_test = c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8, 1, 1.2)
filt = FindNeighbors(filt, dims=1:20, reduction = "integrated.rpca")
for (res in res_to_test){
  filt = FindClusters(filt, resolution = res, cluster.name = paste0("res",res,"_clusters"))
}

head(filt@meta.data)
table(filt$res0.5_clusters)
table(filt$rpca_clusters)
```
## save and or load

```{r eval=F}
saveRDS(filt, here(odir,"multiple_resolutions.rds"))
filt = readRDS(here(odir,"multiple_resolutions.rds"))
```

## UMAPs at different res
```{r}
DimPlot(filt, group.by = "res0.1_clusters")
DimPlot(filt, group.by = "res0.5_clusters") + scale_color_manual(values=clusters)
DimPlot(filt, group.by = "res0.4_clusters") + scale_color_manual(values=clusters) #Keep res0.5 clu7 separate
DimPlot(filt, group.by = "res0.3_clusters") + scale_color_manual(values=clusters) #Or merge clu7
DimPlot(filt, group.by = "res0.2_clusters") + scale_color_manual(values=clusters) #merge cell cycle and subject2
```

## Tree plots

```{r tree}
clustree(filt, prefix="res", suffix ="_clusters")
clustree(filt, prefix="res", suffix ="_clusters", node_colour="sc3_stability")
```

## Cluster tests with scSHC

I imagine we're going to need batch 
```{r}
counts = as.matrix(filt[["RNA"]]$counts)
new_clusters = testClusters(counts, cluster_ids = as.character(filt$res1.2_clusters), 
                            batch = as.character(filt$subject),
                            parallel = F, cores = 1)
```


### SHC clusters
```{r}
table(new_clusters[[1]], filt$res1.2_clusters)
table(new_clusters[[1]], filt$subject)
table(filt$res1.2_clusters, filt$subject)
```

```{r}
print(new_clusters[[2]])
```

### Compare to res 0.3

```{r}
table(new_clusters[[1]], filt$res0.3_clusters)
```

```{r}
filt$shc_clusters = new_clusters[[1]]
DimPlot(filt, group.by=c("shc_clusters","res1.2_clusters"), cols=clusters)
DimPlot(filt, group.by=c("shc_clusters","res1.2_clusters"), cols=clusters, reduction="tsne")
DimPlot(filt, group.by=c("res1.2_clusters"), cols=c(clusters[1:10], rep("grey", 10)), reduction="tsne")
DimPlot(filt, group.by=c("res1.2_clusters"), cols=c(clusters[1:10], rep("grey", 10)), reduction="umap")
DimPlot(filt, group.by=c("shc_clusters","res0.3_clusters"), cols=clusters, reduction="tsne")
DimPlot(filt, group.by=c("shc_clusters","res0.3_clusters"), cols=clusters,)
DimPlot(filt, group.by=c("res0.3_clusters","res0.8_clusters"), cols=clusters, reduction="tsne")
```

### Expore why cluster p2 is split by shc 
1. is it the shc heirarchy
or an artefact of the res1.2 clustering? 

** the tnse split cluster is replicated at res0.3 without splitting p2 in umap; so the tsne profile is not informative**
```{r}
DimPlot(filt, group.by="res0.3_clusters", cols=c("grey","grey","red", rep("grey",4)), alpha=0.5)
DimPlot(filt, group.by="res0.3_clusters", cols=c("grey","grey","red", rep("grey",4)), alpha=0.5, reduction="tsne")
```

Is it a 1.2 clustering artefact? 
No res1.2 clustering does not mysteriously connect the end of p2 and the main bubble. There are slightly more outlier cells than at re0.3, but not enough for me to conclude that the res1.2 clustering itself is an outlier. Therefore it is probably the shc heriarchy that is connecting the pointy end of p2 to the main blob.  This likely reflects a gene expression similarity between these two areas. 
```{r}
DimPlot(filt, group.by="res1.2_clusters", cols=clusters, alpha=0.5)
DimPlot(filt, group.by="res1.2_clusters", cols=c(rep("grey",6),"red", rep("grey",15)), alpha=0.5)
```



```{r}
write.table(new_clusters[[2]], here(odir, "res1.2_cluster_quality_summary.tsv"), row.names = F, quote = F)
```

## Independent shc clustering

```{r}
denovo = scSHC(counts, batch=filt$subject)
table(denovo[[1]], filt$res0.3_clusters)
```


```{r}
denovo[[2]]
```

```{r}
filt$denovo_shc = denovo[[1]]
DimPlot(filt, c("res0.3_clusters","denovo_shc"), cols=clusters)
```

