---
title: "progenitor_rpca_comopsition_test"
author: "Sarah Hazell Pickering"
date: "2025-02-05"
output: html_document
---


```{r}
module="progenitors"
dataname = "rpca"
library(here);i_am(file.path("analysis",paste(module,sep="_",dataname, "clusters_and_markers.Rmd")))

odir = here("output", module, dataname) 

load_data = T #Loading initial seurat analysis
subset_data =  F
```



```{r libs}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)

library(Seurat)
library(BPCells)
```


```{r themes}
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
spectral =  rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
		theme_set(theme_classic(base_size=15) + 
		  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25)))
```

```{r}
clcomp = read.delim(here(odir, "cluster_composition.txt"), colClasses = list(seurat_clusters="factor"))
head(clcomp)
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
```

## Proportion test
 the null tested is that the proportions in each group are the same. 
With such a large n, even clusters 5 and 6 that have a really similar proportion of cells come out as significant.  e.g. standard deviation is < 1%

```{r}
donor_counts = group_by(clcomp, donor_id) %>% summarise(total=sum(n_cells))
donor_counts
```


```{r}
clu_tests = list()
for (clu in 0:(length(levels(clcomp$seurat_clusters))-1)){
  clu_counts = filter(clcomp, seurat_clusters == clu) %>% pull(n_cells)
  res = prop.test(clu_counts, donor_counts$total)
  clu_tests[[paste0("p",clu)]] = res
}
clu_tests

```


## Pairwise prop test
because all the individual comparsions are not necessarily significant
Now we've got to apply p-value adjustment.  
```{r}
subject_pairs = list(1:2,c(1,3),c(1,4),c(1,5),c(1,6),
                     2:3, c(2,4),c(2,5),c(2,6),
                     3:4,c(3,5),c(3,6),
                     4:5, c(4,6), 5:6)
subject_pairs = sapply(subject_pairs, function(pair) {
  paste0("S", pair[1], ".v.S", pair[2])})

clu_tests = data.frame(seurat_clusters =rep(levels(clcomp$seurat_clusters), each=15), 
                       pvalue=NA, 
                       subject_pairs=subject_pairs           )
head(clu_tests)
```


```{r}
pairwise.prop.test = function(names, counts, totals, name_col="donor_id"){
  #counts table with name_col usually donor id and n_cells for the cluster in question
  #names is a vector e.g. c("S1","S2")
  
  counts_filt = rename(counts, subject = all_of(name_col)) %>%filter( subject%in% names) %>% pull(n_cells)
  totals_filt = rename(totals, subject = all_of(name_col)) %>% filter(subject %in% names) %>% pull(total)
  res = prop.test(counts_filt, totals_filt)
  return(res$p.value)
}
pairwise.prop.test(c("S1","S2"), filter(clcomp, seurat_clusters == clu), donor_counts)
```


```{r}
for (clu in 0:(length(levels(clcomp$seurat_clusters))-1)){
  clu_counts = filter(clcomp, seurat_clusters == clu)
  pairs = c()
  for (pair in subject_pairs){
    subjects = stringr::str_split_1(pair[1], pattern=".v.")
    pvalue = pairwise.prop.test(subjects, clu_counts, donor_counts)
    print(pvalue)
    clu_tests$pvalue[clu_tests$seurat_clusters == clu &
                     clu_tests$subject_pairs == pair] =  pvalue
  }
}
clu_tests = mutate(clu_tests, p.adjust=p.adjust(pvalue, method="fdr"))
head(clu_tests)
```

```{r}
adjp_mat = pivot_wider(clu_tests, id_cols=subject_pairs, names_from=seurat_clusters, values_from = p.adjust, names_prefix = "p", values_fn = function(x){round(x,4)})
adjp_mat

group_by(clu_tests,subject_pairs) %>% summarise(round(mean(p.adjust), 4))
group_by(clu_tests,seurat_clusters) %>% summarise(round(mean(p.adjust), 4))
```

```{r}
write.table(adjp_mat, here(odir, "pairwise_donor_composition_test.tsv"),
            sep="\t", quote=F, row.names = F)
```




## permutation-based p-value

```{r}
input_tab = select(filt@meta.data, seurat_clusters, subject) 
head(input_tab)
set.seed(123)
input_tab = input_tab[sample(nrow(input_tab)),]
```


```{r}
compute_sd <- function(data) {
  comp = group_by(data, seurat_clusters, subject) %>%
      summarise(n_cells = n())
  comp = group_by(comp, subject) %>% mutate(cells_per_subject = n_cells/sum(n_cells))
  sd_cell_number = group_by(comp, seurat_clusters) %>% 
    summarise(sd=sd(cells_per_subject), mean = mean(cells_per_subject)) 
  return(sd_cell_number)
}
obs = compute_sd(input_tab)
obs
```

may need to test replace and non-replace in this instance

```{r message=FALSE}
perm_stats = obs 
for (i in 1:1000) {
  # Shuffle the subject labels
  df_sampled <- mutate(input_tab, shuffled_subject = sample(c("S1", "S2", "S3", "S4", "S5", "S6"),
                                                            size=nrow(input_tab), prob =donor_counts$total, 
                                                            replace=T))
  #print(table(df_sampled$shuffled_subject))
  df_sampled =  select(df_sampled, -subject) %>%
                      rename(subject = shuffled_subject)
    #print(head(df_sampled))
  column = paste0("s",i,"_sd")
  # Compute the test statistic for the permutation
  perm_stats[column] <- compute_sd(df_sampled)$sd
}
head(perm_stats)
```

```{r}
iter_tab =  select(perm_stats, seurat_clusters, grep("_sd", colnames(perm_stats), value=T)) %>%
  pivot_longer(ends_with("sd"), names_to=c("iteration",".value"), names_pattern="(.*)_(.*)") 
head(iter_tab)
```

```{r}
obs$pvalue = NA
obs$confidencei = NA 
for (i in 1:nrow(obs)){
  row = obs[i,]
  single_test = filter(iter_tab, seurat_clusters==row$seurat_clusters) %>% pull(sd)
  obs$pvalue[[i]] = mean(single_test >= row$sd)
  obs$confidencei[[i]] = quantile(single_test, 0.99)
}
obs
```


## Plot 99% CI

```{r}
#add in subject for plotting
#Also need the actual cells per cluster
ci_tab = merge(obs, clcomp, by=c("seurat_clusters"), suffixes=c(".total",""))
ci_tab
```


```{r}
ggplot(clcomp, aes(x=seurat_clusters, fill=donor_id, y=cells_per_donor)) + 
  geom_bar(position="dodge", stat='identity') +  scale_fill_manual(values=pal) 

ggplot(clcomp, aes(x=seurat_clusters, fill=donor_id, y=cells_per_donor)) + 
  geom_bar(position="dodge", stat='identity') +  scale_fill_manual(values=pal) +
  geom_errorbar(data=ci_tab, aes(ymin=cells_per_donor-confidencei, ymax=cells_per_donor+confidencei), 
                 position=position_dodge(0.9), width=0.4, linewidth=0.3)
```

```{r}

```




## Cohen's d
```{r}
library(effsize)
s0.3 = sample(c(0,1), size=100, replace=T, prob=c(1-0.3, 0.3))
s0.4 = sample(c(0,1), size=100, replace=T, prob=c(1-0.43, 0.43))
summary(s0.3)
summary(s0.4)
cohen.d(sample(c(0,1), size=100, replace=T, prob=c(0.3, 1-0.3)), 
        sample(c(0,1), size=100, replace=T, prob=c(0.43, 1-0.43)))
```





## compute for cell numbers
but of course we have different cell numbers per subject; so this is not super useful 
```{r}
input_tab = select(filt@meta.data, seurat_clusters, subject) 
head(input_tab)
set.seed(123)
input_tab = input_tab[sample(nrow(input_tab)),]

compute_sd <- function(data) {
  sd_cell_number = group_by(data, seurat_clusters, subject) %>%
      summarise(n_cells = n(), prop=n()/sum(n())) %>% group_by(seurat_clusters) %>% summarise(sd=sd(n_cells), mean = mean(n_cells)) 
  return(sd_cell_number)
}
obs = compute_sd(input_tab)
obs
```

may need to test replace and non-replace in this instance

```{r message=FALSE}
perm_stats = obs 
for (i in 1:1000) {
  # Shuffle the subject labels
  df_sampled <- mutate(input_tab, shuffled_subject = sample(c("S1", "S2", "S3", "S4", "S5", "S6"),
                                                            size=nrow(input_tab), prob =donor_counts$total, 
                                                            replace=T))
  #print(table(df_sampled$shuffled_subject))
  df_sampled =  select(df_sampled, -subject) %>%
                      rename(subject = shuffled_subject)
    #print(head(df_sampled))
  column = paste0("s",i,"_sd")
  # Compute the test statistic for the permutation
  perm_stats[column] <- compute_sd(df_sampled)$sd
}
head(perm_stats)
```

```{r}
iter_tab =  select(perm_stats, seurat_clusters, grep("_sd", colnames(perm_stats), value=T)) %>%
  pivot_longer(ends_with("sd"), names_to=c("iteration",".value"), names_pattern="(.*)_(.*)") 
head(iter_tab)
```

```{r}
obs$pvalue = NA
obs$confidencei = NA 
for (i in 1:nrow(obs)){
  row = obs[i,]
  single_test = filter(iter_tab, seurat_clusters==row$seurat_clusters) %>% pull(sd)
  obs$pvalue[[i]] = mean(single_test >= row$sd)
  obs$confidencei[[i]] = quantile(single_test, 0.99)
}
obs
```




## Plot 99% CI

```{r}
#add in subject for plotting
#Also need the actual cells per cluster
ci_tab = merge(obs, clcomp, by=c("seurat_clusters"), suffixes=c(".total",""))
ci_tab
```


```{r}
ggplot(clcomp, aes(x=seurat_clusters, fill=donor_id, y=n_cells)) + 
  geom_bar(position="dodge", stat='identity') +  scale_fill_manual(values=pal) +
  geom_errorbar(data=ci_tab, aes(ymin=n_cells-confidencei, ymax=n_cells+confidencei), 
                 position=position_dodge(0.9), width=0.4, linewidth=0.3)
```


