---
title: "progenitors_rpca_cluster_stability"
author: "Sarah Hazell Pickering"
date: "2024-08-16"
output: html_document
---
Subset :) Mainly developing the function to annotate current clusters with shc info
Exploring cluster stability with clustree on resolutions 0.1 -> 1.2 and sch posthoc heirarchical clustering and statistcial testing

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(clustree)
library(dplyr)
library(ggplot2)
library(ggsci)
library(here)
library(tidyr)
library(scSHC)
```

```{r params}
module = "progenitors"
dataname = "integration_trial"
load_data = T #Loading previous multiple res data?
subset_data = T

rmd_filename = "progenitors_rpca_cluster_stability_subset"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0", tvthemes::gravityFalls_pal()(12))
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```

## load data

```{r}
if (load_data){
    filt = readRDS(here(odir,"multiple_resolutions.rds"))
    if (subset_data){
        Idents(filt) = "res0.3_clusters"
        filt = subset(filt, downsample=500)
    } 
} 

if (!load_data){
  if (subset_data){
    filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
    filt = readRDS(here(odir, "complete_analysis.rds"))
    }
}

filt
table(filt$res0.3_clusters)
```

```{r}
if (!load_data){
    filt[["mnn.reconstructed"]] =NULL
filt = JoinLayers(filt)
Idents(filt) = "rpca_clusters"
filt
head(filt@meta.data) 
table(filt$rpca_clusters)

red_to_rm = grep("harmony|fastmnn", names(filt@reductions), value = T)
for (red in red_to_rm){
  print(paste0("red removed ", red))
  filt[[red]] = NULL
}
#set default reductions
filt[["umap"]] = filt[["umap.rpca"]]
filt[["tsne"]] = filt[["rpca.tsne"]]
filt
}

```

## Testing different resolutions

```{r}
if (!load_data){
    res_to_test = c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8, 1, 1.2)
    filt = FindNeighbors(filt, dims=1:20, reduction = "integrated.rpca")
    for (res in res_to_test){
      filt = FindClusters(filt, resolution = res, cluster.name = paste0("res",res,"_clusters"))
    }
    
    head(filt@meta.data)
    table(filt$res0.5_clusters)
    table(filt$rpca_clusters)
}

```
## save and or load

```{r}
if(!subset_data){
    saveRDS(filt, here(odir,"multiple_resolutions.rds"))
}
```

## UMAPs at different res
```{r}
DimPlot(filt, group.by = "res0.1_clusters")
DimPlot(filt, group.by = "res0.5_clusters") + scale_color_manual(values=clusters)
DimPlot(filt, group.by = "res0.4_clusters") + scale_color_manual(values=clusters) #Keep res0.5 clu7 separate
DimPlot(filt, group.by = "res0.3_clusters") + scale_color_manual(values=clusters) #Or merge clu7
DimPlot(filt, group.by = "res0.2_clusters") + scale_color_manual(values=clusters) #merge cell cycle and subject2
```

## Tree plots

```{r tree}
clustree(filt, prefix="res", suffix ="_clusters")
clustree(filt, prefix="res", suffix ="_clusters", node_colour="sc3_stability")
```

## Cluster tests with scSHC

I imagine we're going to need batch 
```{r}
counts = as.matrix(filt[["RNA"]]$counts)
new_clusters = testClusters(counts, cluster_ids = as.character(filt$res1.2_clusters), 
                            batch = as.character(filt$subject),
                            parallel = F, cores = 1)
```


### SHC clusters
```{r}
table(new_clusters[[1]], filt$res1.2_clusters)
table(new_clusters[[1]], filt$subject)
table(filt$res1.2_clusters, filt$subject)
```

```{r}
print(new_clusters[[2]])
```

### Compare to res 0.3

```{r}
table(new_clusters[[1]], filt$res0.3_clusters)
```

```{r}
filt$shc_clusters = new_clusters[[1]]
DimPlot(filt, group.by=c("shc_clusters","res1.2_clusters"), cols=clusters)
DimPlot(filt, group.by=c("shc_clusters","res1.2_clusters"), cols=clusters, reduction="tsne")
DimPlot(filt, group.by=c("shc_clusters","res0.3_clusters"), cols=clusters, reduction="tsne")
DimPlot(filt, group.by=c("shc_clusters","res0.3_clusters"), cols=clusters,)
DimPlot(filt, group.by=c("res0.3_clusters","res0.8_clusters"), cols=clusters, reduction="tsne")
```

## Create an annotation function

```{r}
#simple method:
freq = table(new_clusters[[1]], filt$res0.3_clusters)
freq
annot = paste0("new_p",max.col(freq)-1)
annot

contributing_clus = (freq/rowSums(freq) > 0.25)
contributing_clus
colnames(contributing_clus)

annot[7] = "new_p3/2"

shc_clu = c()
for (i in 1:nrow(freq)){
    row = freq[i,]
    print(row)
    cat("\n")
    primary_clu = names(which.max(row))
    primary_clu
    contributing_clus = row/sum(row) > 0.25
    if (sum(contributing_clus) >1){
        secondary_clus = names(row)[row/sum(row) > 0.25 & names(row) != primary_clu]
        shc_clu = c(shc_clu, paste0("new_p", primary_clu,"//p", secondary_clus))
    } else{
        shc_clu = c(shc_clu, paste0("new_p", primary_clu))
    }
        
  
}
shc_clu
```


##Extract pvals per cluster

```{r}
dend = new_clusters[[2]]
as.character(dend)
str(dend)
```



## Check marker genes and ontology

