---
title: "differential_abundance"
author: "Sarah Hazell Pickering"
date: "2024-09-16"
output: html_document
---
```{r params}
module="progenitors"
dataname = "rpca"
library(here); i_am(file.path("analysis",paste(module,sep="_",
                                               paste0(dataname, "_differential_abundance.Rmd"))))
odir = here("output", module, dataname) 

load_data = T #Loading initial seurat analysis
subset_data =  F

if (!dir.exists(odir)){dir.create(odir, recursive = T)}
```



```{r libs, warning=F, message=F, echo=T}
library(dplyr)
library(ggplot2)
library(edgeR)
library(tidyr)

library(Seurat)
library(BPCells)

pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters =  c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0")
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
```



```{r}
comp = read.delim(here(odir, "cluster_composition.txt"))
comp$donor_id = factor(comp$donor_id)
comp$seurat_clusters = factor(comp$seurat_clusters)
comp = group_by(comp,seurat_clusters) %>% mutate( sd_cells_per_donor=sd(cells_per_donor))
head(comp)
wide = pivot_wider(comp, names_from = donor_id, values_from = n_cells, id_cols = seurat_clusters,
                   names_sort=T) %>% as.data.frame()
rownames(wide) = wide$seurat_clusters
wide$seurat_clusters = NULL
head(wide)
```


## Effect sizes

```{r}
ggplot(comp, aes(x=donor_id, y=cells_per_donor, fill=seurat_clusters)) + geom_bar(stat="identity") +
  scale_fill_manual(values=clusters)

rank_and_plot = function(comp, cluster){
  to_plot = filter(comp, seurat_clusters == cluster) %>% arrange(cells_per_donor)
  to_plot$donor_id = factor(to_plot$donor_id, levels=to_plot$donor_id)
  ggplot(to_plot, aes(x=donor_id, y=cells_per_donor, fill=seurat_clusters)) + geom_bar(stat="identity") +
  scale_fill_manual(values=clusters[cluster+1]) + theme_classic(base_size=18)
}
rank_and_plot(comp, 0)

for (i in 0:7){
  print(rank_and_plot(comp, i))
}
```

for each cluster (x axis) I want to plot the range and mean (boxplot + scatter basically) of the percentage change

```{r}
ggplot(comp, aes(x=seurat_clusters, colour=seurat_clusters, y=cells_per_donor)) + 
  geom_boxplot() +geom_point()+ geom_line(aes(group=donor_id)) + 
  scale_colour_manual(values=clusters) + theme_classic(base_size=18)

ggplot(comp, aes(x=seurat_clusters, colour=seurat_clusters, y=cells_per_donor)) + 
  geom_boxplot() + geom_point() + 
  scale_colour_manual(values=clusters) + theme_classic(base_size=18)

ggplot(comp, aes(x=seurat_clusters, y=sd_cells_per_donor, fill=seurat_clusters)) + 
  geom_bar(stat="summary", fun="mean") +
  scale_fill_manual(values=clusters) + theme_classic(base_size=18)
```




## Read data

```{r load_data}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
  Idents(filt) = "seurat_clusters"
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
head(filt@meta.data) 
table(Idents(filt))
```


```{r}
Idents(filt) = "subject"
subset1 = subset(filt,downsample=1000, seed=1)
table(subset1$seurat_clusters, subset1$subject)

subset2 = subset(filt,downsample=1000, seed=2)
table(subset2$seurat_clusters, subset2$subject)

subset3 = subset(filt, downsample=1000, seed=3)
table(subset3$seurat_clusters, subset3$subject)
```

```{r}
all_prop = merge(unclass(table(subset1$seurat_clusters, subset1$subject)), unclass(table(subset2$seurat_clusters, subset2$subject)),
                      by="row.names", suffixes=c("_ds1","_ds2"))
third_rep = unclass(table(subset3$seurat_clusters, subset3$subject))
colnames(third_rep) = paste0(colnames(third_rep), "_ds3")
all_prop = merge(all_prop, third_rep,
                      by.x="Row.names",by.y="row.names")

rownames(all_prop) = all_prop$Row.names
all_prop$Row.names = NULL
head(all_prop)
```

## Test every cluster 
```{r}
sample_info = data.frame(sample_id = colnames(all_prop),
                         donor_id = gsub("_ds.*","", colnames(all_prop)),
                         subset_id = gsub(".*_ds", "ds", colnames(all_prop)))
y.ab = DGEList(all_prop, samples = sample_info)
y.ab

design = model.matrix(~0+donor_id, y.ab$samples)
colnames(design) = gsub("donor_id","", colnames(design))
head(design)
```


```{r}
y.ab = estimateDisp(y.ab, design, trend.method="none")
summary(y.ab$common.dispersion)
plotBCV(y.ab, cex=1)
```


```{r}
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)
summary(fit.ab$df.prior)
plotQLDisp(fit.ab, cex=1)
```

```{r}

head(design)
contrasts = makeContrasts(S1=S1-((S2+S3+S4+S5+S6)/5),
                          S2=S2-((S1+S3+S4+S5+S6)/5),
                          S3=S3-((S1+S2+S4+S5+S6)/5),
                          S4=S4-((S1+S2+S3+S5+S6)/5),
                          S5=S5-((S1+S2+S3+S4+S6)/5),
                          S6=S6-((S1+S2+S3+S4+S5)/5),
                          levels=design)
res <- glmQLFTest(fit.ab, contrast = contrasts)
summary(decideTests(res, coef=1))
```


```{r}
topTags(res)
```

If we run the DA with all 8 clusters, the most significant is cluster 4; the donor S2 specific cluster.  This might be affectig the proportions of the others.  

### Without cluster 4


```{r}
all_prop = all_prop[rownames(all_prop)!= "4",]
sample_info = data.frame(sample_id = colnames(all_prop),
                         donor_id = gsub("_ds.*","", colnames(all_prop)),
                         subset_id = gsub(".*_ds", "ds", colnames(all_prop)))
y.ab = DGEList(all_prop, samples = sample_info)
y.ab

design = model.matrix(~0+donor_id, y.ab$samples)
colnames(design) = gsub("donor_id","", colnames(design))
head(design)
```


```{r}
y.ab = estimateDisp(y.ab, design, trend.method="none")
summary(y.ab$common.dispersion)
plotBCV(y.ab, cex=1)
```


```{r}
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)
summary(fit.ab$df.prior)
plotQLDisp(fit.ab, cex=1)
```

```{r}

head(design)
contrasts = makeContrasts(S1=S1-((S2+S3+S4+S5+S6)/5),
                          S2=S2-((S1+S3+S4+S5+S6)/5),
                          S3=S3-((S1+S2+S4+S5+S6)/5),
                          S4=S4-((S1+S2+S3+S5+S6)/5),
                          S5=S5-((S1+S2+S3+S4+S6)/5),
                          S6=S6-((S1+S2+S3+S4+S5)/5),
                          levels=design)
res <- glmQLFTest(fit.ab, contrast = contrasts)
summary(decideTests(res, coef=1))
```


```{r}
top = topTags(res, n = nrow(res$table))
top
```

After removing cluster4, cluster 3 is the most different between donors, followed by cluster 0.  Cluster 7 is also fairly significant.  



