---
title: "rpca_published_annotation"
author: "Sarah Hazell Pickering"
date: "2024-08-13"
output: html_document
---

```{r params}
module="progenitors"
dataname = "rpca"
library(here);i_am(file.path("analysis",paste(module,sep="_",dataname, "clusters_and_markers.Rmd")))
odir = here("output", module, dataname) 

load_data = T #Loading initial seurat analysis
subset_data =  F

if (!dir.exists(odir)){dir.create(odir, recursive = T)}
```


```{r libs, warning=F, message=F, echo=T}
library(biomaRt)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)

library(Seurat)
library(BPCells)
library(HGNChelper)

library(clusterProfiler)

pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
spectral =  rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
```


```{r}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
all.genes = rownames(filt)
```


## Read markers

```{r}
norm.markers = read.delim(file.path(odir, "marker_genes.txt"))
norm.markers$cluster = factor(norm.markers$cluster, levels=0:max(as.integer(norm.markers$cluster)))
uniq.norm.markers = norm.markers[!norm.markers$gene %in% names(which(table(norm.markers$gene) > 1)),]
head(norm.markers)
table(norm.markers$cluster)
```


```{r}
liu = read.delim(here("data/other_papers/liu_2022/41421_2022_402_MOESM3_ESM.csv"), skip=1, header=T, sep=",")
liu = liu[grep("^c[[:digit:]]*$", liu$cluster),] #eliminate metadata at the end
liu = liu[order(liu$p_val, -liu$avg_logFC),]
liu = separate_longer_delim(liu,gene,delim=" /// " )
nrow(liu); tail(liu, n=10)
liu$gene =  checkGeneSymbols(liu$gene, unmapped.as.na = F)$Suggested.Symbol
```

```{r}
sb_clusters= read.delim(here("data/other_papers/strieder-barboza/TableS2_strieder-barboza.csv"),skip=2, sep=",")
sb_clusters$X = NULL
sb_clusters= separate_longer_delim(sb_clusters,gene,delim=" /// " )
head(sb_clusters); table(sb_clusters$cluster)
sb_clusters$gene = checkGeneSymbols(sb_clusters$gene, unmapped.as.na = F)$Suggested.Symbol
```

```{r}
vj_clusters= read.delim(here("data/other_papers/vijay2020/Vijiay_Suppplementary_table15.csv"),skip=2, sep=",")
vj_clusters$Gene = checkGeneSymbols(vj_clusters$Gene, unmapped.as.na = F)$Suggested.Symbol
```

```{r}
whtk = readxl::read_excel(here("data/other_papers/whytock2022/SuppTable5_DEGs_in_SVF.xlsx"), skip = 2)
new_colnames = rep(grep("\\.{3}", invert=T, value=T, colnames(whtk) ), each=9)
colnames(whtk) = paste(new_colnames, whtk[1,], sep="/")
whtk = whtk[2:nrow(whtk),]
head(whtk)
whtk_clusters = pivot_longer(whtk, cols=colnames(whtk), names_to=c("cluster",".value"), names_sep="/") 
whtk_clusters$`NA` = NULL
whtk_clusters$Gene =  checkGeneSymbols(whtk_clusters$Gene, unmapped.as.na = F)$Suggested.Symbol
head(whtk_clusters)
table(whtk_clusters$cluster)
```




## Summarise modules to show

```{r}
pub_modules = list(c0_asc_preadipo_Liu = liu$gene[liu$cluster == "c0"],
                   c1_asc_intermediate_Liu = liu$gene[liu$cluster == "c1"],
                   c3_asc_progenitor_Liu = liu$gene[liu$cluster == "c3"],
                   c5_asc_mhc_Liu = liu$gene[liu$cluster == "c5"],
                   c20_pericyte_Liu = liu$gene[liu$cluster == "c20"], #included as outgroups 
                   c15_prolif.nkt_Liu = liu$gene[liu$cluster == "c15"],
                   SM_asc_Strieder.Barboza = sb_clusters$gene[sb_clusters$cluster == "SM-ASC"],
                   FA_asc_Strieder.Barboza = sb_clusters$gene[sb_clusters$cluster == "FA-ASC"],
                   IM_asc_Strieder.Barboza = sb_clusters$gene[sb_clusters$cluster == "IM-ASC"],
                   EC_Strieder.Barboza = sb_clusters$gene[sb_clusters$cluster == "EC"],
                   SP1_early_ASC_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP1"],
                   SP2_mature_ASC_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP2"],
                   SP3_early_ASC_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP3"],
                   SP4_ECM_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP4"],
                   SP5_Inflam_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP5"],
                   Stem1_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Stem 1" & whtk_clusters$avg_log2FC > 0],
                   Stem2_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Stem 2" & whtk_clusters$avg_log2FC > 0],
                   Stem3_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Stem 3" & whtk_clusters$avg_log2FC > 0],
                   PreAd1_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Pre-Ad 1" & whtk_clusters$avg_log2FC > 0],
                   PreAd2_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Pre-Ad 2" & whtk_clusters$avg_log2FC > 0])
str(pub_modules)
adipo_only = list(c0_asc_preadipo_Liu = liu$gene[liu$cluster == "c0"],
                   c1_asc_intermediate_Liu = liu$gene[liu$cluster == "c1"],
                   c3_asc_progenitor_Liu = liu$gene[liu$cluster == "c3"],
                   c5_asc_mhc_Liu = liu$gene[liu$cluster == "c5"],
                   SM_asc_Strieder.Barboza = sb_clusters$gene[sb_clusters$cluster == "SM-ASC"],
                   FA_asc_Strieder.Barboza = sb_clusters$gene[sb_clusters$cluster == "FA-ASC"],
                   IM_asc_Strieder.Barboza = sb_clusters$gene[sb_clusters$cluster == "IM-ASC"],
                   SP1_early_ASC_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP1"],
                   SP2_mature_ASC_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP2"],
                   SP3_early_ASC_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP3"],
                   SP4_ECM_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP4"],
                   SP5_Inflam_VJ = vj_clusters$Gene[vj_clusters$Cluster == "SP5"],
                   Stem1_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Stem 1" & whtk_clusters$avg_log2FC > 0],
                   Stem2_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Stem 2" & whtk_clusters$avg_log2FC > 0],
                   Stem3_whytock = whtk_clusters$Gene[whtk_clusters$cluster == "Stem 3" & whtk_clusters$avg_log2FC > 0])
str(adipo_only)
```


```{r warning=F}
filt= AddModuleScore(filt, pub_modules, name="pub", search = T, verbose=F)
```

Not found in object: AC245595.1
The following features are still not present in the object: H4C3, H2AZ1, H1-5, H1-3, H1-2, H3C2, H2AC12, H2AC11, H2AC16, MACROH2A1
No updated symbols foundWarning: The following features are still not present in the object: MIR145, CTD-2269F5.1
Warning: No updated symbols foundWarning: The following features are still not present in the object: AP001347.6, MARCHF3, AC007246.3, AL163953.3
Warning: No updated symbols foundWarning: The following features are still not present in the object: H3-3
Warning: No updated symbols foundWarning: The following features are still not present in the object: RP11-347P5.1, H2AZ2, AC090498.1

## Marker genes  and expression

```{r}
colnames(filt@meta.data)[grep("pub",colnames(filt@meta.data))] = names(pub_modules)
names(pub_modules)
DotPlot(filt, features=names(pub_modules), col.min=0, cols=c("lightgrey","red")) + coord_flip() #& scale_colour_gradientn(colours = spectral)
DotPlot(filt, features=names(adipo_only), col.min=0, cols=c("lightgrey","red")) + coord_flip() #& scale_colour_gradientn(colours = spectral)
```

```{r}
FeaturePlot(filt, features = c("SP5_Inflam_VJ","SP4_ECM_VJ","SP2_mature_ASC_VJ","SP1_early_ASC_VJ"))& scale_colour_gradientn(colours = spectral)
FeaturePlot(filt, features = c("FA_asc_Strieder.Barboza","SM_asc_Strieder.Barboza", 
                               "IM_asc_Streider.Barboza","EC_Strieder.Barboza")) & 
  scale_colour_gradientn(colours = spectral)
```
```{r}
FeaturePlot(filt, features = c("c3_asc_progenitor_Liu","c0_asc_preadipo_Liu", 
                               "c1_asc_intermediate_Liu","c5_asc_mhc_Liu")) & 
  scale_colour_gradientn(colours = spectral)
```

```{r}
FeaturePlot(filt, features = c("Stem1_whytock","Stem2_whytock", 
                               "Stem3_whytock","PreAd1_whytock")) & 
  scale_colour_gradientn(colours = spectral)
```


```{r}
RidgePlot(filt, features = c("SP5_Inflam_VJ"))

```



## Massier meta-analysis clusters

```{r}
massier = read.delim(here("data/other_papers/massier2023/Massier2023_subq_FAP_markers.txt"))
nrow(massier); table(massier$cluster)
summary(grepl("///",massier$gene))
massier= separate_longer_delim(massier,gene,delim=" /// ")
massier = filter(massier, p_val_adj < 0.01, avg_log2FC > 0.25, pct.1 > 0.1)
nrow(massier); table(massier$cluster)
head(massier)
massier$gene = checkGeneSymbols(massier$gene, unmapped.as.na = F)$Suggested.Symbol
summary(massier)
```

```{r}
massier$clu_name = paste0("sf", massier$cluster)
massier_modules = list()
for (clu in 0:max(massier$cluster)){
 massier_modules[[paste0("sf", clu)]] = filter(massier, cluster == clu) %>% pull(gene) 
}
str(massier_modules)
filt= AddModuleScore(filt, massier_modules, name="massier", search = T, verbose=F)
```


```{r}
colnames(filt@meta.data)[grep("massier",colnames(filt@meta.data))] = names(massier_modules)
DotPlot(filt, features=rev(names(massier_modules)), col.min=0, cols=c("lightgrey","red")) + coord_flip()
```

## Overrepresentation of marker genes

### Make tables
```{r}
liu_t2g = liu[ grep("c[0135]$",liu$cluster ), c("cluster","gene")]
liu_t2g$cluster = paste("Liu2022", liu_t2g$cluster)
head(liu_t2g)

sb_t2g = sb_clusters[c("cluster","gene")]
sb_t2g$cluster = paste("Strieder-Barboza", sb_t2g$cluster)
head(sb_t2g)

vj_t2g = vj_clusters[c("Cluster","Gene")]
colnames(vj_t2g) = c("cluster","gene")
vj_t2g$cluster = paste("Vijay", vj_t2g$cluster)
head(vj_t2g)

wtk_t2g = whtk_clusters[grep("Stem",whtk_clusters$cluster ), c("cluster","Gene")]
colnames(wtk_t2g) = c("cluster","gene")
wtk_t2g$cluster = paste("Whytock", wtk_t2g$cluster)
head(wtk_t2g)
```


### ORA
```{r}
pub_go = compareCluster(gene~cluster, data = norm.markers,  
                        fun="enricher", TERM2GENE = rbind(liu_t2g, sb_t2g, vj_t2g, wtk_t2g), 
                        universe = all.genes)
head(pub_go)[,c(2,4:10)]
```

```{r}
nrow(pub_go)
```

```{r}
filter(pub_go, cluster == 1) %>% head()
filter(pub_go, cluster == 5) %>% head()
filter(pub_go, cluster == 6) %>% head()
```



```{r}
dotplot(pub_go)
```
### Dot plots
```{r cluster_go, dev=c("png","pdf")}
dfgo = as.data.frame(pub_go)
dfgo$cluster = factor(dfgo$cluster, levels=0:max(as.integer(dfgo$cluster)))
go_to_plot = dfgo %>% group_by(cluster) %>%
              slice_min(n = 2, order_by = p.adjust) %>% arrange(cluster)
go_to_plot$Description= factor(go_to_plot$Description, levels=rev(unique(go_to_plot$Description)))
go_to_plot$cluster = factor(go_to_plot$cluster, levels=unique(go_to_plot$cluster))

p = ggplot(go_to_plot, aes(x=cluster, y=Description, colour=cluster, size=Count)) + 
  geom_point()+  theme_bw(base_size=15) + 
  theme(axis.text.x=element_text(angle=30, hjust=1)) +
  scale_color_manual(values=clusters[as.integer(levels(go_to_plot$cluster))+1])
p
```

```{r cluster_go_overlapping, dev=c("png","pdf")}
include_all_plot = filter(dfgo, Description %in% go_to_plot$Description)
include_all_plot$Description = factor(include_all_plot$Description, levels=levels(go_to_plot$Description))
a = ggplot(include_all_plot, aes(x=cluster, y=Description, colour=cluster, size=Count)) + 
  geom_point()+  theme_bw(base_size=15) + 
  theme(axis.text.x=element_text(angle=30, hjust=1)) +
  scale_color_manual(values=clusters[as.integer(levels(go_to_plot$cluster))+1])
a
```

## Palani et al. CD29
All of our cells are ITGB1+ (CD29)

```{r}
RidgePlot(filt, features = "ITGB1", slot="counts")
RidgePlot(filt, features = "ITGB1", slot="data")
```


## Massier GO 

```{r}
mas_t2g = massier[c("clu_name","gene")]
colnames(mas_t2g) = c("cluster","gene")
mas_t2g$cluster = paste("Massier", mas_t2g$cluster)
head(mas_t2g)
```


```{r}
pub_go = compareCluster(gene~cluster, data = norm.markers,  
                        fun="enricher", TERM2GENE = mas_t2g, 
                        universe = all.genes)
head(pub_go)[,c(2,4:10)]
```

```{r}
dotplot(pub_go, showCategory=3)
```

### Dot plots
```{r massier_dot, dev=c("png","pdf")}
dfgo = as.data.frame(pub_go)
dfgo$cluster = factor(dfgo$cluster, levels=0:max(as.integer(dfgo$cluster)))
go_to_plot = dfgo %>% group_by(cluster) %>%
              slice_min(n = 2, order_by = p.adjust) %>% arrange(cluster)
go_to_plot$Description= factor(go_to_plot$Description, levels=rev(unique(go_to_plot$Description)))
go_to_plot$cluster = factor(go_to_plot$cluster, levels=unique(go_to_plot$cluster))

p = ggplot(go_to_plot, aes(x=cluster, y=Description, colour=cluster, size=Count)) + 
  geom_point()+  theme_bw(base_size=15) + 
  theme(axis.text.x=element_text(angle=30, hjust=1)) +
  scale_color_manual(values=clusters[as.integer(levels(go_to_plot$cluster))+1])
p
```

```{r massier_dot_overlap, dev=c("png","pdf")}
include_all_plot = filter(dfgo, Description %in% go_to_plot$Description)
include_all_plot$Description = factor(include_all_plot$Description, levels=levels(go_to_plot$Description))
a = ggplot(include_all_plot, aes(x=cluster, y=Description, colour=cluster, size=Count)) + 
  geom_point()+  theme_bw(base_size=15) + 
  theme(axis.text.x=element_text(angle=30, hjust=1)) +
  scale_color_manual(values=clusters[as.integer(levels(go_to_plot$cluster))+1])
a
```
sf3  = cd34- & overlaps with cluster 3
sf4 = Areg type
sf14 = marked by COL4A1 & overlaps clusters 2 and 7
sf16 = late committed preadipocytes

```{r}
FeaturePlot(filt, features=c("sf3","sf4","sf14","sf16")) & scale_color_gradientn(colours=spectral)
```


