---
title: "progenitors_rpca_cluster_stability"
author: "Sarah Hazell Pickering"
date: "2024-08-16"
output: html_document
---

Exploring cluster stability with clustree on resolutions 0.1 -> 1.2 and sch posthoc heirarchical clustering and statistcial testing

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(clustree)
library(dplyr)
library(ggplot2)
library(ggsci)
library(here)
library(tidyr)
library(scSHC)
```


## Setup

```{r params}
module = "progenitors"
dataname = "rpca"
load_data = T #Loading from previous clustree analysis? 
subset_data = F
resolution=0.8
clu_name = paste0("res", resolution, "_clusters")

rmd_filename = "progenitors_rpca_clustree"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#0A7EA8","#EF5780","#0A8550","#1C31D3","#EE9614","#670F39")
clusters = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0", tvthemes::gravityFalls_pal()(12))
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```


## load data

```{r}
if (load_data){
    filt = readRDS(here(odir,"multiple_resolutions.rds"))
    if (subset_data){
        Idents(filt) = "res0.3_clusters"
        filt = subset(filt, downsample=500)
    } 
} 

if (!load_data){
  if (subset_data){
    filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
    filt = readRDS(here(odir, "complete_analysis.rds"))
    }
}

filt
table(filt$seurat_clusters)
```


## Testing different resolutions


```{r}
if (!load_data){
    res_to_test = c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8, 1, 1.2)
    filt = FindNeighbors(filt, dims=1:20, reduction = "integrated.rpca")
    for (res in res_to_test){
      filt = FindClusters(filt, resolution = res, cluster.name = paste0("res",res,"_clusters"))
    }
    
    head(filt@meta.data)
    table(filt$res0.5_clusters)
    table(filt$rpca_clusters)
}

```

## Save if this is the first time

```{r}
if(!subset_data & !load_data){
    saveRDS(filt, here(odir,"multiple_resolutions.rds"))
}
```

## UMAPs at different res
```{r}
DimPlot(filt, group.by = "res0.8_clusters")+ scale_color_manual(values=clusters)
DimPlot(filt, group.by = "res0.5_clusters") + scale_color_manual(values=clusters)
DimPlot(filt, group.by = "res0.4_clusters") + scale_color_manual(values=clusters) 
DimPlot(filt, group.by = "res0.3_clusters") + scale_color_manual(values=clusters) 
DimPlot(filt, group.by = "res0.2_clusters") + scale_color_manual(values=clusters) 
```

## Tree plots

```{r tree}
clustree(filt, prefix="res", suffix ="_clusters")
clustree(filt, prefix="res", suffix ="_clusters", node_colour="sc3_stability")
```

## Cluster tests with scSHC

```{r}
counts = as.matrix(filt[["RNA"]]$counts)
new_clusters = testClusters(counts, cluster_ids = as.character(filt@meta.data[,clu_name]), 
                            batch = as.character(filt$subject),
                            parallel = F, cores = 1)
```


### SHC clusters
```{r}
table(new_clusters[[1]], filt@meta.data[,clu_name])
table(new_clusters[[1]], filt$subject)
table(filt@meta.data[,clu_name], filt$subject)
```

```{r}
print(new_clusters[[2]])
```

### Compare to res 0.3

```{r}
table(new_clusters[[1]], filt$res0.3_clusters)
```

```{r}
filt$shc_clusters = new_clusters[[1]]
DimPlot(filt, group.by=c("shc_clusters",clu_name), cols=clusters)
DimPlot(filt, group.by=c("shc_clusters",clu_name), cols=clusters, reduction="tsne")
DimPlot(filt, group.by=c("shc_clusters","res0.3_clusters"), cols=clusters,)
```


## Create an annotation function

```{r}
#simple method:
freq = table(new_clusters[[1]], filt$res0.3_clusters)
freq
annot = paste0("new_p",max.col(freq)-1)
names(annot) = rownames(freq)
annot
```


```{r}
annotate_clusters = function(new_clusters, old_clusters){
    freq = table(new_clusters, old_clusters)
    
    shc_clu = c()
    for (i in 1:nrow(freq)){
        row = freq[i,]
        primary_clu = names(which.max(row))
        contributing_clus = row/sum(row) > 0.2
        if (sum(contributing_clus) >1){
            secondary_clus = names(row)[row/sum(row) > 0.2 & names(row) != primary_clu]
            shc_clu = c(shc_clu, paste0("new_p", primary_clu,"//p", paste0(secondary_clus, collapse="//p")))
        } else{
            shc_clu = c(shc_clu, paste0("new_p", primary_clu))
        }
    }
    return(shc_clu)
}
annotate_clusters(new_clusters[[1]], filt$res0.3_clusters)
```


```{r}
annot0.8 = annotate_clusters(new_clusters[[1]], filt$res0.8_clusters)
names(annot0.8) = rownames(freq)
annot0.8
```

## extract pavlue

data.tree object

```{r}
new_clusters[[2]]
pvals = as.data.frame(new_clusters[[2]])
colnames(pvals) = "tree"
pvals[1,]


annot = annotate_clusters(new_clusters[[1]], filt$res0.3_clusters)
names(annot) = gsub("new", "Cluster",rownames(freq))
annot
```


```{r}
align_vector = function(x){
    old_ids = c()
    for (shc_label in x){
        print(shc_label)
        old_ids = c(old_ids,annot[[shc_label]])
    }
    return(old_ids)
}

pvals = mutate(pvals, type = if_else (grepl("Cluster",tree), "cluster", "node"),
               pval =if_else (grepl(":",tree), 
                              gsub(".*:", "", gsub(" ","", tree)), 
                                NA),
               shc_label= gsub(":.*", "", gsub(".*--","", gsub(" ","", tree))),
               tree_order = 1:nrow(pvals)
)
pvals$cluster_match = NA
pvals$cluster_match[pvals$type == "cluster"] = annot[pvals$shc_label[pvals$type=="cluster"]]
pvals = select(pvals, tree_order, tree, cluster_match,pval, type, shc_label)
pvals
```

```{r}
filter(pvals, type=="cluster")
filter(pvals, type=="node")
```

## save statistics
```{r}
saveRDS(new_clusters, here(odir, paste0(clu_name, "_shc_clusters.rds")))
write.table(pvals, here(odir, paste0(clu_name,"_statistics.tsv")), row.names = F, quote = F)
```


