---
title: "progenitors_rpca_cell_cycle"
author: "Sarah Hazell Pickering"
date: "2024-08-12"
output: html_document
---


To classify cell cycle stages we used TinyAtlas list from https://github.com/hbc/tinyatlas for G2M and S phase genes, calculated module scores using cellCycleScoring from seurat 

An alternative to seurat's phase sorting mechanism is to use Model based clustering to classify cells into cell cycle states.  In the case where the expression threshold for both phases was met whichever phase had the higher aggregated expression was used.  Mclust from CRAN: Model-based clustering based on parameterized finite Gaussian mixture models. Models are estimated by EM algorithm initialized by hierarchical model-based agglomerative clustering. The optimal model is then selected according to BIC. 

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(dplyr)
library(ggplot2)
library(ggsci)
library(here)
library(mclust)
library(biomaRt)
```

```{r params}
module = "progenitors"
dataname = "rpca"
load_data = T #Loading initial seurat analysis
subset_data = F

rmd_filename = "progenitors_rpca_cell_cycle"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)

```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)

theme_zoom = theme_classic(30) &  
    theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks = element_line(linewidth = 0.5),
        strip.background = element_blank()) 
clusters = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0", tvthemes::gravityFalls_pal()(5))
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```

## load data

```{r}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
```

```{r C_inset_umap}
DimPlot(filt, cols=clusters) + 
    theme_hp + scale_colour_manual(values=clusters, name="Progenitor\ncluster", label = function(x) paste0("p",x))
DimPlot(filt, cols=clusters, pt.size=0.75) + coord_cartesian(x=c(2,9),y=c(-6,1))+
    theme_zoom + scale_colour_manual(values=clusters ,guide="none")
```


```{r}
cc_file <- "https://raw.githubusercontent.com/hbc/tinyatlas/master/cell_cycle/Homo_sapiens.csv"
cell_cycle_genes <- read.csv(cc_file)
head(cell_cycle_genes)
table(cell_cycle_genes$phase)
```


```{r}
mart <- biomaRt::useMart(biomart = "ensembl",
  dataset = "hsapiens_gene_ensembl",
  host = "https://apr2020.archive.ensembl.org") #ensembl98 no longer available 99 closest
annot = biomaRt::getBM(c("external_gene_name", "description", "ensembl_gene_id", "gene_biotype"), 
              filters = "ensembl_gene_id",
              values = as.character(cell_cycle_genes$geneID),
              mart = mart,
              useCache = F)
annot$description = gsub("\\[.*\\]", "", annot$description)
head(annot)
cell_cycle_genes = merge(cell_cycle_genes, annot, by.x="geneID", by.y="ensembl_gene_id")
head(cell_cycle_genes)
write.table(here("data/gene_sets/cell_cycle_genes_ensembl.tsv"))
```

```{r}
filt <- CellCycleScoring(filt,
                      g2m.features = cell_cycle_genes$external_gene_name[cell_cycle_genes$phase == "G2/M"],
                      s.features = cell_cycle_genes$external_gene_name[cell_cycle_genes$phase == "S"])
summary(filt$G2M.Score)
```

```{r Aphase}
phase_plots = FeaturePlot(filt, features=c( "G2M.Score","S.Score"), blend=T, ncol=2, combine=F, pt.size=0.5) 
phase_plots[[3]] & theme_hp & theme(legend.position="none") & labs(title=NULL)
phase_plots[[4]] & theme_zoom & labs(title=NULL)
```


```{r Cphasescatter}
FeatureScatter(filt, feature1="G2M.Score", feature2="S.Score", group.by="rpca_clusters", cols=clusters, pt.size=0.75)+ 
    theme_hp + labs(title=NULL) + 
    scale_colour_manual(values=clusters, name="Progenitor\ncluster", label = function(x) paste0("p",x),
                        guide = guide_legend(override.aes = list(size=3)))
    
```


```{r }
ggplot(filt@meta.data) + geom_point(aes(x=G2M.Score, y=S.Score, colour=rpca_clusters))
```

```{r Bvlnplot}
VlnPlot(filt, features=c("G2M.Score", "S.Score"), group.by="rpca_clusters", cols=clusters, pt.size = 0) & 
    theme_hp & scale_x_discrete(name = "Progenitor cluster", label=function(x) paste0("p",x)) &
    scale_fill_manual(values=clusters, guide="none") &
    labs( y="Average expression score")
```

## Proportion plot

```{r composition}
composition = read.delim(here(odir, "cluster_composition.txt"))
composition$seurat_clusters = factor(composition$seurat_clusters, levels=c(0,1:max(composition$seurat_clusters)))
head(composition)

clu_to_plot = c(4,5)
ggplot(filter(composition, seurat_clusters %in% clu_to_plot)) + 
  geom_bar(stat="identity",aes(x=donor_id, y=n_cells, fill=seurat_clusters)) + 
  scale_y_continuous(expand = c(0.01, 0))+ 
   scale_fill_manual(values=clusters[clu_to_plot+1])

ggplot(filter(composition, seurat_clusters %in% clu_to_plot)) + 
  geom_bar(stat="identity",aes(x=donor_id, y=cells_per_donor, fill=seurat_clusters), position = "dodge") + 
  scale_y_continuous(expand = c(0.01, 0))+ 
   scale_fill_manual(values=clusters[clu_to_plot+1])
```


```{r Dproportion}
ggplot(filter(composition, seurat_clusters %in% clu_to_plot)) + 
  geom_bar(stat="identity",aes(x=donor_id, y=cells_per_donor, fill=seurat_clusters)) + 
  scale_y_continuous(expand = expansion(mult=c(0, 0.05)))+ 
   scale_fill_manual(values=clusters[clu_to_plot+1], labels=function(x) paste0("p", x)) +
    labs(x="Subject", y="Proportion of cells", fill="Progenitor\nclusters")
```

Default seurat threshold is too low, setting any cell with high than 0 expression of these genes as cycling.  
```{r default_phase}
DimPlot(filt,
        group.by= "Phase")+ theme_hp
```

## An alternative method for thresholding:

```{r}
g2m_model = Mclust(filt$G2M.Score, 2)
plot(g2m_model, what="classification", main="Mclust Classification")
filt@meta.data$g2m_model = as.factor(g2m_model$classification)
ggplot(filt@meta.data) + geom_histogram(aes(x=G2M.Score, fill=g2m_model), bins=100)
```

```{r}
s_model = Mclust(filt$S.Score, 2)
filt@meta.data$s_model = as.factor(s_model$classification)
ggplot(filt@meta.data) + geom_histogram(aes(x=S.Score , fill=s_model), bins=100)
```

## Decision tree

```{r}
filt$Phase = NA
filt$Phase [filt$G2M.Score > 0 & filt$g2m_model == 2 &
                (filt$s_model == 1 | filt$S.Score < filt$G2M.Score)] = "G2M"
filt$Phase [filt$S.Score > 0 & filt$s_model == 2 &
                (filt$g2m_model == 1 |  filt$G2M.Score < filt$S.Score)] = "S"
filt$Phase[(filt$g2m_model == 1 & filt$s_model == 1)  |
             (filt$S.Score < 0 & filt$G2M.Score < 0) ] = "G1"
```


```{r}
table(filt$Phase, useNA = "ifany")
ggplot(filt@meta.data) + geom_point(aes(x=G2M.Score, y=S.Score, colour=Phase))+ theme_hp
ggplot(filt@meta.data) + geom_point(aes(x=G2M.Score, y=S.Score, colour=rpca_clusters)) + scale_color_manual(values=clusters) + theme_hp
```

```{r umap_phase}
DimPlot(filt, group.by = "Phase")+ theme_hp
DimPlot(filt, group.by = "Phase", reduction = "pca") + theme_hp
```



