---
title: "progenitors_rpca_cell_cycle"
author: "Sarah Hazell Pickering"
date: "2024-08-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

To classify cell cycle stages we used TinyAtlas list from https://github.com/hbc/tinyatlas for G2M and S phase genes, calculated module scores using cellCycleScoring from seurat and used Model based clustering to classify cells into cell cycle states.  In the case where the expression threshold for both phases was met whichever phase had the higher aggregated expression was used.  

Mclust from CRAN: Model-based clustering based on parameterized finite Gaussian mixture models. Models are estimated by EM algorithm initialized by hierarchical model-based agglomerative clustering. The optimal model is then selected according to BIC. 



```{r params}
module="progenitors"
dataname = "rpca"
library(here);i_am(file.path("analysis",paste(module,sep="_",dataname, "clusters_and_markers.Rmd")))
odir = here("output", module, dataname) 

load_data = T #Loading initial seurat analysis
subset_data =  F

if (!dir.exists(odir)){dir.create(odir, recursive = T)}
```


```{r libs, warning=F, message=F, echo=T}
library(biomaRt)
library(dplyr)
library(ggplot2)
library(ggsci)

library(Seurat)
library(BPCells)

library(mclust)#for classifying

pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
spectral =  rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
```


```{r}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
```

```{r}
cc_file <- "https://raw.githubusercontent.com/hbc/tinyatlas/master/cell_cycle/Homo_sapiens.csv"
cell_cycle_genes <- read.csv(cc_file)
head(cell_cycle_genes)
table(cell_cycle_genes$phase)
```


```{r}
mart <- biomaRt::useMart(biomart = "ensembl",
  dataset = "hsapiens_gene_ensembl",
  host = "https://sep2019.archive.ensembl.org") #ensembl98
annot = biomaRt::getBM(c("external_gene_name", "description", "ensembl_gene_id", "gene_biotype"), 
              filters = "ensembl_gene_id",
              values = as.character(cell_cycle_genes$geneID),
              mart = mart,
              useCache = F)
annot$description = gsub("\\[.*\\]", "", annot$description)
head(annot)
cell_cycle_genes = merge(cell_cycle_genes, annot, by.x="geneID", by.y="ensembl_gene_id")
head(cell_cycle_genes)
```

```{r}
filt <- CellCycleScoring(filt,
                      g2m.features = cell_cycle_genes$external_gene_name[cell_cycle_genes$phase == "G2/M"],
                      s.features = cell_cycle_genes$external_gene_name[cell_cycle_genes$phase == "S"])
```

Default seurat threshold is too low, setting any cell with high than 0 expression of these genes as cycling.  
```{r}
DimPlot(filt,
        group.by= "Phase")
FeaturePlot(filt, features=c("S.Score", "G2M.Score")) & scale_color_gradientn(colours = spectral)
ggplot(filt@meta.data) + geom_point(aes(x=G2M.Score, y=S.Score, colour=Phase), size=0.5, alpha=0.75)
```

```{r}
g2m_model = Mclust(filt$G2M.Score, 2)
plot(g2m_model, what="classification", main="Mclust Classification")
filt@meta.data$g2m_model = as.factor(g2m_model$classification)
ggplot(filt@meta.data) + geom_histogram(aes(x=G2M.Score, fill=g2m_model), bins=100)
```

```{r}
s_model = Mclust(filt$S.Score, 2)
filt@meta.data$s_model = as.factor(s_model$classification)
ggplot(so@meta.data) + geom_histogram(aes(x=S.Score , fill=s_model), bins=100)
```

## Decision tree

```{r}
filt$Phase = NA
filt$Phase [filt$G2M.Score > 0 & filt$g2m_model == 2 &
                (filt$s_model == 1 | filt$S.Score < filt$G2M.Score)] = "G2M"
filt$Phase [filt$S.Score > 0 & filt$s_model == 2 &
                (filt$g2m_model == 1 |  filt$G2M.Score < filt$S.Score)] = "S"
filt$Phase[(filt$g2m_model == 1 & filt$s_model == 1)  |
             (filt$S.Score < 0 & filt$G2M.Score < 0) ] = "G1"
```


```{r}
table(filt$Phase, useNA = "ifany")
ggplot(filt@meta.data) + geom_point(aes(x=G2M.Score, y=S.Score, colour=Phase))
```

```{r}
DimPlot(filt, group.by = "Phase")
DimPlot(filt, group.by = "Phase", reduction = "pca")
```



