---
title: "rpca_velociraptor dynamical"
author: "Sarah Hazell Pickering"
date: "2024-08-28"
output: html_document
---

Use paired end mode 

```{r params}
module="progenitors"
dataname = "rpca"
library(here); i_am(file.path("analysis",paste(module,sep="_",
                                               paste0(dataname, "_integration.Rmd"))))
indir = here("data/cellranger/count_no_introns/day0")
odir = here("output", module, dataname) 
load_data = T #Loading initial seurat analysis
subset_data =  F
model="dynamical"

if (!dir.exists(odir)){dir.create(odir, recursive = T)}
```

```{r libs, warning=F, message=F, echo=F}
library(dplyr)
library(ggplot2)
library(Seurat)
library(BPCells)
library(LoomExperiment)
library(RColorBrewer)
library(velociraptor)
library(SingleCellExperiment)
library(scater) #for tsne/umaps
prog_cols = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0")
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
  #tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
```

```{r}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
  Idents(filt) = "seurat_clusters"
} else {
  filt = readRDS(here(odir, "complete_analysis.rds"))
  DefaultAssay(filt) = "RNA"
}
filt
variable.genes = VariableFeatures(filt)
```

Equalise cell numbers between subjects with random downsampling
```{r}
Idents(filt) = "subject"
table(Idents(filt))
#subject S4 has least number of cells after filtering
#downsample other subjects

sbs = subset(filt, downsample=sum(filt$subject =="S4"))
table(Idents(sbs))
rm(filt)
Idents(sbs) = "rpca_clusters"
```



Set dataname to save a separate result for the pairedend velocity analysis
```{r}
red_name = "INTEGRATED.RPCA"
#umap_name = "UMAP"
clu_name =  "rpca_clusters"
#tsne_name = "TSNE"
```

```{r}
DimPlot(sbs, group.by=clu_name, cols=clusters, reduction=tolower(red_name))
DimPlot(sbs, cols=clusters, reduction="umap")
```

### Convert to SingleCell Experiment

```{r}
sce = as.SingleCellExperiment(sbs, assay="RNA")
sce
```

```{r}
rowData(sce)
head(rownames(sce))
head(colData(sce))
```


```{r}
rm(sbs)
```

## Load our unspliced data

```{r}
samples = list.dirs(here(indir), full.names = F, recursive = F)
samples = samples[grep("p.chr18|beige", samples, invert = T)]

data_list = c()
for (i in 1:length(samples)){
  print(samples[i])
  #Finding file name if different from sample/dir name
  path = here(indir, samples[i], "velocyto")
  fn = list.files(path)
  fn = fn[grep("Sample.\\.loom", fn)]

  loom = import(here(path, fn), 
                type="SingleCellLoomExperiment")
  cell_id = paste0(gsub("_",".",samples[i]),"_")
  colnames(loom) = gsub(".*:",cell_id, gsub("x$","-1",colData(loom)$CellID))
  rownames(loom) = rowData(loom)$Gene
  rownames(loom)[duplicated(rownames(loom))] = paste0(rownames(loom)[duplicated(rownames(loom))], ".1")
  data_list = c(data_list,loom)
}
summary(data_list)

loom = do.call(cbind, data_list)
loom
```


```{r}
rowData(loom)
colData(loom)
assays(loom)[["unspliced"]]
loom$subject = gsub("_.*","", gsub("day0\\.subject", "",colnames(loom)))
table(loom$subject)
```


### Combine objects
the handling of duplicate gene names is different between the two methods.  The loom stores more gene information, so it doesn't matter about the duplicates.  seurat though doesn't load the  extra gene information

```{r}
summary(rownames(sce) %in% rownames(loom))
summary(colnames(sce) %in% colnames(loom))
head(colnames(sce)[!colnames(sce) %in% colnames(loom)] )

# one cell missing because cellranger filtering
ncol(sce)
sce = sce[,colnames(sce) %in% colnames(loom)]
ncol(sce)


# filter cells from seurat analysis
filt = loom[rownames(sce),colnames(sce)]

assays(sce)[["unspliced"]] = assays(filt)[["unspliced"]]
assays(sce)[["spliced"]] = assays(filt)[["spliced"]]
sce
table(sce$subject)
```


## Get some stats on the unspliced counts
Sum = total counts and 
detected = the number of detected features
calculating for the unspliced 

```{r}
sce <- addPerCellQC(sce, 
    subsets=list(Mito=grep("MT-", rownames(sce))),
    assay.type = "unspliced")
colData(sce)
plotColData(sce, y = "sum", x="detected", color_by="seurat_clusters", other_fields = "subject") + facet_wrap(~subject) 
singleCellTK::plotSCEViolin(sce, "colData", "sum", groupBy = "subject", ylab="nCounts")
singleCellTK::plotSCEViolin(sce, "colData", "detected", groupBy = "subject", ylab ="nFeatures")
```

Okay interesting, the number of detected features drops for BOTH day1_rep2 and day3, even though day3 has been quantified in paired-end mode.  


## Filter low expr genes

```{r find_filter}
nonzeroes = (assay(sce, "spliced") > 0 )* (assay(sce, "unspliced") > 0)
head(nonzeroes)

shared_counts = assay(sce, "spliced")*nonzeroes + assay(sce, "unspliced")*nonzeroes 
head(shared_counts)

genes_to_keep = rowSums(shared_counts) > 20
summary(genes_to_keep)
```

```{r}
filt = sce[genes_to_keep,]
filt

summary(variable.genes %in% rownames(filt))
variable.genes = variable.genes[variable.genes %in% rownames(filt)]
reducedDimNames(sce)
```

## Starting velocity analysis

Supplying variable genes and previously computed Integration reduction (acts as a PCA)



```{r}
velo.out = scvelo(sce, subset.row = variable.genes, use.dimred=red_name,
                  assay.X="counts", mode= model)
```

```{r}
velo.out
sce$velocity_pseudotime = velo.out$velocity_pseudotime
```

## pseuodtime
```{r}
plotTSNE(sce, color_by="velocity_pseudotime")
plotUMAP(sce, color_by="velocity_pseudotime")
```

## Arrows on PCA

```{r}
embedded_pca <- embedVelocity(reducedDim(sce, red_name), velo.out)
dim(embedded_pca)
```


```{r}
reducedDim(sce, "PCA.SIMPLE") = reducedDim(sce, red_name)[,1:2]
grid.df <- gridVectors(sce, embedded_pca[,1:2], use.dimred = "PCA.SIMPLE")
colnames(grid.df) = gsub(gsub("\\.","",red_name), "PC", colnames(grid.df), ignore.case = T)
head(grid.df)
dim(grid.df)
plotReducedDim(sce, red_name,colour_by=clu_name) +
    geom_segment(data=grid.df, mapping=aes(x=start.PC_1, y=start.PC_2, 
        xend=end.PC_1, yend=end.PC_2, colour=NULL), alpha=0.5,arrow=arrow(length=unit(0.05, "inches"))) &
  scale_colour_manual(values=clusters)

plotReducedDim(sce, red_name,colour_by="subject") +
    geom_segment(data=grid.df, mapping=aes(x=start.PC_1, y=start.PC_2, 
        xend=end.PC_1, yend=end.PC_2, colour=NULL), alpha=0.5,arrow=arrow(length=unit(0.05, "inches"))) &
  scale_colour_manual(values=pal)
```

## Arrows on UMAP

```{r} 
embedded <- embedVelocity(reducedDim(sce, "UMAP"), velo.out)
dim(embedded)
grid.df <- gridVectors(sce, embedded, use.dimred = "UMAP", resolution=40)
head(grid.df)
dim(grid.df)
if (subset_data){pt_size=1} else{pt_size=0.2}

plotReducedDim(sce, "UMAP",colour_by=clu_name, point_size=pt_size) +
    geom_segment(data=grid.df, mapping=aes(x=start.umap_1, y=start.umap_2, 
        xend=end.umap_1, yend=end.umap_2, colour=NULL), width=1,arrow=arrow(length=unit(0.05, "inches"))) +
  scale_colour_manual(values=clusters)

plotReducedDim(sce, "UMAP",colour_by="subject", point_size=pt_size) +
    geom_segment(data=grid.df, mapping=aes(x=start.umap_1, y=start.umap_2, 
        xend=end.umap_1, yend=end.umap_2, colour=NULL), width=1,arrow=arrow(length=unit(0.05, "inches"))) +
  scale_colour_manual(values=pal)
```



## Arrows on tSNE

```{r} 
embedded <- embedVelocity(reducedDim(sce, "TSNE"), velo.out)
dim(embedded)
grid.df <- gridVectors(sce, embedded, use.dimred = "TSNE", resolution=40)
head(grid.df)
dim(grid.df)
if (subset_data){pt_size=1} else{pt_size=0.2}

plotReducedDim(sce, "TSNE",colour_by=clu_name, point_size=pt_size) +
    geom_segment(data=grid.df, mapping=aes(x=start.tsne_1, y=start.tsne_2, 
        xend=end.tsne_1, yend=end.tsne_2, colour=NULL), width=1,arrow=arrow(length=unit(0.05, "inches"))) +
  scale_colour_manual(values=clusters)

plotReducedDim(sce,"TSNE",colour_by="subject", point_size=pt_size) +
    geom_segment(data=grid.df, mapping=aes(x=start.tsne_1, y=start.tsne_2, 
        xend=end.tsne_1, yend=end.tsne_2, colour=NULL), width=1,arrow=arrow(length=unit(0.05, "inches"))) +
  scale_colour_manual(values=pal)
```



## poking around the output object

```{r}
colnames(colData(velo.out))
names(rowData(velo.out))
```

I see a column here named root_cells.  How does scvelo select root cells?  
velocity confidence, cool, but confidence_transition?  

Then we have stuff calculated per gene; like alpha, beta & gamma; fit_steady, fit_likelihood

```{r}
summary(velo.out$root_cells)
summary(velo.out$end_points)
```

```{r}
str(metadata(velo.out))
```



## Velocity confidence and length
confidence = coherence of velocity directions  
length = speed of transcription between cells = roughly arrow length

```{r}
sce$velocity_confidence = velo.out$velocity_confidence
sce$velocity_length = velo.out$velocity_length
plotReducedDim(sce, "UMAP",color_by="velocity_confidence", )

plotReducedDim(sce, red_name,color_by="velocity_confidence", )

plotReducedDim(sce, "UMAP",color_by="velocity_length" )
```

## Velocity stream plots
Error in `plotVelocityStream()`:
! use.dimred %in% reducedDimNames(sce) is not TRUE
so not running this one
```{r}
colnames(reducedDim(sce, "UMAP")) 
str(embedded)
```


```{r}
colnames(reducedDim(sce, "UMAP")) = c("V1","V2")
str(embedded)

plotVelocityStream(sce, embedded, use.dimred="UMAP", 
                   color_by = clu_name,grid.resolution=20,
                   stream.L = 10,
                   arrow.angle = 10, arrow.length = 0.5)

plotVelocityStream(sce, embedded, use.dimred="UMAP", 
                   color_by = clu_name,grid.resolution=30,
                   stream.L = 8,
                   arrow.angle = 15, arrow.length = 0.5)

plotVelocityStream(sce, embedded, use.dimred="UMAP", 
                   color_by = clu_name,grid.resolution=40,
                   stream.L = 5,
                   arrow.angle = 15, arrow.length = 0.5)
```


```{r}
if (subset_data == F){
  saveRDS(sce, file = here(odir,paste0(model, "_velocity_scexperiment.rds")))
  saveRDS(velo.out, file = here(odir,paste0(model, "_velocity_output.rds")))
  
} else if (subset_data){
  saveRDS(sce, file = here(odir,paste0("10%_subset_", model, "_velocity_scexperiment.rds")))
  saveRDS(velo.out, file = here(odir, paste0("10%_subset_", model, "_velocity_output.rds")))
}
```

