---
title: "Progenitor seurat qc and analysis"
output: html_document
---

previous code: mito_on_mad_take2_full
now updating seuratv5

```{r params}
module = "progenitors"
dataname = "initial"
library(here); i_am(file.path("analysis",module, "seurat_qc_and_analysis.Rmd"))
indir = here("data/cellranger/count_no_introns/day0")
odir = here("output", module, dataname) 
dataname = "initial"

dir.create(odir, recursive = T)
load_data = T
subset_data = F #subset data to 10% total cells?
```


```{r libs}
library(dplyr, quietly = T)
library(ggplot2)
library(ggsci)

library(Seurat)
library(BPCells)
library(patchwork)
library(clusterProfiler, quietly = T)
library(enrichplot)
```


```{r colors}
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
library(RColorBrewer)
library(ggsci)
clusters = ggsci::pal_simpsons()(12)[c(1:7,9:12)]
spectral =  rev(brewer.pal(n = 11, name = "Spectral"))

clusters = c("#FED439FF", "#709AE1FF", "#8A9197FF", "#D2AF81FF",
             "#FD7446FF", "#D5E4A2FF", "#197EC0FF", "#46732EFF", "#71D0F5FF")
```

## Read data

This filtered matrix has been filtered for cells with low UMI counts already (EmptyDrops method).  
If first run, save bpcells objects, otherwise load data.  Can work on a subset if subset_data=T

```{r read_or_load}
samples = list.dirs(here(indir), full.names = F, recursive = F)
samples

if (load_data==FALSE){# load count matrices
  
  data_list = c()
  
  for (s in samples){
    #if this has been run before delete bp dir
    bc_matrix = Read10X(data.dir = here(indir, s, "outs/filtered_feature_bc_matrix"))
    so = CreateSeuratObject(counts=bc_matrix, project=s, names.field = 2, names.delim = "-")
    write_matrix_dir(mat=so[["RNA"]]$counts, dir = here(odir, "bpcells", s))
    gene_mat = open_matrix_dir(here(odir, "bpcells", s))
    so[["RNA"]]$counts = gene_mat
    
    data_list = c(data_list,so)
  }
  names(data_list) = samples
  aggr = merge(data_list[[1]], y= data_list[2:length(data_list)], 
               add.cell.ids = gsub("_",".",samples), project=dataname )
  
  saveRDS(aggr,here(odir,"unfiltered_object.rds"))

  
  rm(data_list, bc_matrix, so)
} else if(load_data){
  aggr = readRDS(here(odir, "unfiltered_object.rds"))
}

aggr
head(aggr@meta.data) #auto metadata
```


```{r}
aggr$subject = gsub("_.*","", gsub("day0.subject","",colnames(aggr)))
Idents(aggr) = "subject"
aggr$orig.ident = aggr$subject
table(Idents(aggr))
aggr <- JoinLayers(aggr)
aggr
```

## Quality control

Number of genes (features) detected per cell high, due to the depth of our sequencing and the stem-cell nature of the samples.  There is a wide spread of the number of genes (nFeature)and reads (nCount) per cell.  The percentage mitochondrial genes is super low for most cells, but some few have up to 80% mt.   


```{r before_filt1}
aggr[["percent.mt"]] = PercentageFeatureSet(aggr, pattern="^MT-")
VlnPlot(aggr, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0, cols=pal)
VlnPlot(subset(x = aggr, downsample = 1000), features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, cols=pal)
RidgePlot(aggr, feature= "nFeature_RNA", cols=pal) + geom_vline(aes(xintercept=2000))
```

```{r before_filter}
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "percent.mt", cols=pal)
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", cols=pal)
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal)
```
We set an adaptive Mt.percent filter that's the 3* mad maximum. 
```{r}
mt_info = group_by(aggr@meta.data, subject) %>% 
  summarise(mad = mad(percent.mt),
            median = median(percent.mt), 
            thresh = median(percent.mt) + mad(percent.mt)*3,
            th95 = quantile(percent.mt, 0.95))
mt_info
thresh = max(mt_info$thresh)
```
```{r}
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal) + 
  geom_hline(aes(yintercept=thresh), linetype="dashed") +
  geom_vline(aes(xintercept=2000), linetype="dashed")
```


```{r filter}
ncol(aggr)
filt <- subset(aggr,  percent.mt < thresh & nFeature_RNA > 2000)
ncol(filt)

FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size=0.5, cols=pal)
FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size=0.5, cols=pal)
FeatureScatter(filt, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal) 
```


## Normalise

```{r normalise}
filt= NormalizeData(filt, normalization.method = "LogNormalize", scale.factor = 10000)
```

## Identification of Variable Features

```{r variable_features}
filt <- FindVariableFeatures(filt, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(filt), 10)

# plot variable features with and without labels
plot1 = VariableFeaturePlot(subset(filt, downsample=1000))
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

## Scaling

Centres around zero and sets the variance to 1 - basically a z-score transformation.  Here we use all genes, but by default Seurat uses only identified variable features.

```{r scaling}
all.genes = rownames(filt)
filt = ScaleData(filt, features=all.genes)
```

## PCA

PC1 is separating a small number of cells from the main group.  MKI67 is very much a prolif/cell cycle gene.  

```{r pca}
filt = RunPCA(filt )
print(filt[["pca"]], dims = 1:5, nfeatures = 5)
FeaturePlot(filt, reduction = "pca", feature="MKI67", cols=c("lightgrey","red"))
```

## Determine the dimensionality

```{r elbow}
ElbowPlot(filt, ndims = 50)
```

## Clustering

```{r clust}
filt <- FindNeighbors(filt, dims = 1:20)
filt <- FindClusters(filt, resolution = 0.5)
```

```{r}
table(Idents(filt))
VlnPlot(filt, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0, cols=clusters)
```


### Cluster Composition
```{r}
composition = group_by(filt@meta.data, subject,seurat_clusters) %>% count()
composition = composition[order(composition$n, decreasing = T),]
colnames(composition) = c("donor_id","seurat_clusters","n_cells")
head(composition)
```


```{r}
ggplot(composition) + geom_bar(stat="identity",aes(y=donor_id, x=n_cells, fill=seurat_clusters)) + 
  scale_y_discrete(limits=rev)+ scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15) + scale_fill_manual(values=clusters)
ggplot(composition) + geom_bar(stat="identity",aes(y=donor_id, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15)  + scale_fill_manual(values=clusters)
```

```{r}
ggplot(composition) + geom_bar(stat="identity",aes(fill=donor_id, x=n_cells, y=seurat_clusters)) +
    scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)  + scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15)
ggplot(composition) + geom_bar(stat="identity",aes(fill=donor_id, x=n_cells, y=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)  + scale_x_continuous(expand = c(0.01, 0))+ 
  theme_classic(base_size=15)
```



## Non-linear dimensionality reduction

```{r umap}
filt = RunUMAP(filt, dims=1:20)
DimPlot(filt, reduction="umap", label=T, cols=clusters)
DimPlot(filt, reduction="umap", group.by = "subject", label=T, cols=pal, label.color = "white", label.box=T)
DimPlot(filt, reduction="umap", split.by = "subject",group.by = "seurat_clusters", ncol=3, cols = clusters)
```

## t-SNE

```{r}
filt = RunTSNE(filt, dims = 1:20)
DimPlot(filt, reduction = "tsne", label=T, cols=clusters)
DimPlot(filt, reduction = "tsne", group.by= "subject", label=T, cols=pal, label.color = "white", label.box = T)
DimPlot(filt, reduction = "tsne", split.by = "subject", ncol=3,cols = clusters)
```


## Find Markers
```{r densify}
cluster.markers = FindAllMarkers(filt, min.pct=0.25, logfc.threshold=0.5, 
                              densify=T, only.pos=T)
head(cluster.markers); dim(cluster.markers)
table(filter(cluster.markers,p_val_adj < 0.01) %>% pull(cluster))

top.markers = as.data.frame(cluster.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC))
top.markers

expr_markers = as.data.frame(cluster.markers) %>%
    group_by(cluster) %>% filter(pct.1 > 0.5) %>%
    slice_max(n=2, order_by = avg_log2FC)
expr_markers

multiply_markers = as.data.frame(cluster.markers) %>%
    group_by(cluster) %>% filter(p_val_adj < 0.01) %>%
    slice_max(n=2, order_by = avg_log2FC*pct.1)
multiply_markers
```

```{r}
library(ggrepel)
ggplot(filter(cluster.markers, cluster ==7), aes(x=avg_log2FC, y=pct.1, color=pct.1)) + geom_point() + 
  geom_text_repel(data = filter(cluster.markers, cluster == 7) %>% slice_max(n=20, order_by=avg_log2FC),
                  aes(label=gene))

ggplot(filter(cluster.markers, cluster ==7), aes(x=avg_log2FC, y=avg_log2FC*pct.1, color=pct.1)) + geom_point() + 
  geom_text_repel(data = filter(cluster.markers, cluster == 7) %>% slice_max(n=20, order_by=avg_log2FC),
                  aes(label=gene))
```

Multplication method = logfc * min.pct1 = MKI67(+) + CDC20(+)
Threshold method = logfc if min.pct.1 > 0.5 = MKI67(+) & DLGAP5 (+?)
LogFC only = SPCC25(+) & HMMR(?)

```{r}
FeaturePlot(filt, features=c("MKI67")) & scale_color_gradientn(colours=spectral) & ggtitle("MKI67 \nlogfc*min.pct1 And logfc (min.pct > 0.5)")
FeaturePlot(filt, features=c("CDC20")) & scale_color_gradientn(colours=spectral)& ggtitle("CDC20 \nlogfc*min.pct1")
FeaturePlot(filt, features=c("DLGAP5")) & scale_color_gradientn(colours=spectral)& ggtitle("DLGAP5\nlogfc (min.pct > 0.5)")
FeaturePlot(filt, features=c("SPC25")) & scale_color_gradientn(colours=spectral)& ggtitle("SPC25\nlogfc only")
FeaturePlot(filt, features="HMMR") & scale_color_gradientn(colours=spectral)& ggtitle("HMMR \nlogfc only")
```


### Marker Gene Dot Plots
```{r}
DotPlot(filt, features=rev(unique(top.markers$gene))) & 
  scale_colour_gradientn(colours = spectral) & coord_flip()

top.markers$gene = factor(top.markers$gene, levels=rev(unique(top.markers$gene)))
overlap.to.plot = filter(cluster.markers, gene %in% top.markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(top.markers$gene))

ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point() + 
  theme_bw(base_size=15)  + 
  theme(axis.text.x=element_text(angle=30, hjust=1))  + 
  scale_color_manual(values=clusters)
```


```{r}
DotPlot(filt, features=rev(unique(expr_markers$gene))) & 
  scale_colour_gradientn(colours = spectral) & coord_flip()

expr_markers$gene = factor(expr_markers$gene, levels=rev(unique(expr_markers$gene)))
overlap.to.plot = filter(cluster.markers, gene %in% expr_markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(expr_markers$gene))

ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point() + 
  theme_bw(base_size=15)  + 
  theme(axis.text.x=element_text(angle=30, hjust=1))  + 
  scale_color_manual(values=clusters)
```

```{r}
DotPlot(filt, features=rev(unique(multiply_markers$gene))) & 
  scale_colour_gradientn(colours = spectral) & coord_flip()
multiply_markers$gene = factor(multiply_markers$gene, levels=rev(unique(multiply_markers$gene)))
overlap.to.plot = filter(cluster.markers, gene %in% multiply_markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(multiply_markers$gene))

ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point() + 
  theme_bw(base_size=15)  + 
  theme(axis.text.x=element_text(angle=30, hjust=1))  + 
  scale_color_manual(values=clusters)
```


Visualisation methods

```{r marker_vis}
RidgePlot(filt,features = c("ACTA2","TAGLN"), slot="scale.data")

for (i in 1:(length(multiply_markers$gene)/4)){
  start = ((i-1)*4)+1; print(start)
  end = i*4; print(end)
  to_plot = multiply_markers$gene[start:end]
  to_plot = as.character(to_plot)
  p= FeaturePlot(subset(filt,downsample=1000), features = to_plot) & 
  scale_colour_gradientn(colours = spectral)
  print(p)
}

FeaturePlot(subset(filt,downsample=1000), features = c("percent.mt", "nCount_RNA","nFeature_RNA")) & 
  scale_colour_gradientn(colours = spectral)
FeaturePlot(subset(filt,downsample=1000), features = c("PC_1","PC_2","PC_3","PC_4"))& 
  scale_colour_gradientn(colours = spectral)
```


```{r}
cluster.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(subset(filt, downsample=600), features= top10$gene)
```

And then we would usually assign cell type identity.  

## Calculate donor markers
```{r}
Idents(filt) = "subject"
donor.markers = FindAllMarkers(filt, min.pct=0.25, logfc.threshold = 0.25, 
                               only.pos=T, densify=T)
Idents(filt)= "seurat_clusters"

head(donor.markers)

top.donor.markers = as.data.frame(donor.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC))
top.donor.markers

multiply_donor = as.data.frame(donor.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC*pct.1))
```

```{r}
DotPlot(filt, features=top.donor.markers$gene, group.by="subject") & 
  scale_colour_gradientn(colours = spectral) & coord_flip()
```

```{r}
DotPlot(filt, features=rev(unique(multiply_donor$gene)), group.by = "subject") & 
  scale_colour_gradientn(colours = spectral) & coord_flip()

multiply_donor$gene = factor(multiply_donor$gene, levels=rev(unique(multiply_donor$gene)))
overlap.to.plot = filter(donor.markers, gene %in% multiply_donor$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(multiply_donor$gene))

ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point() + 
  theme_bw(base_size=15)  + 
  theme(axis.text.x=element_text(angle=30, hjust=1))  + 
  scale_color_manual(values=pal)
```


### Save 

```{r save} 
if (subset_data == FALSE){
  saveRDS(filt, file = here(odir, "complete_analysis.rds"))
  write.table(cluster.markers, here (odir, "marker_genes.txt"),
            sep="\t", row.names = F)

  write.table(donor.markers, here(odir,"donor_marker_genes.txt"),
            sep="\t", row.names = F )
  write.table(composition, here(odir, "cluster_composition.txt"),
              sep="\t", row.names = F)
  Idents(filt) = "orig.ident"
  sub = subset(filt, downsample=1000)
  Idents(filt) = "seurat_clusters"
  saveRDS(sub,here(odir,"10%_complete_analysis.rds"))
} else if (subset_data){ #if subset, don't save complete analysis
  saveRDS(filt,here(odir, "10%_subset_analysis.rds"))
  write.table(cluster.markers, file.path(odir,dataname, "10%_subset_marker_genes.txt"),
            sep="\t", row.names = F)
}
```

