---
title: "bulk_day15_mitopathways"
author: "Sarah Hazell Pickering"
date: "2024-09-16"
output: html_document
---

```{r}
library(here); i_am("analysis/bulk/bulk_day15_mitopathways.Rmd")
knitr::opts_knit$set(dev=c("png","pdf"))
odir = here("output/bulk")
figdir = here(odir,"plots")
if (!dir.exists(figdir)){dir.create(figdir, recursive = T)}
```

```{r message=FALSE, warning=FALSE, echo=TRUE}
library(tidyr)
library(dplyr)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(stringr)
library(viridis)
```


```{r}
mitop = read.delim(here("data/other_papers/Human.MitoCarta3.0.csv"), sep=",", stringsAsFactors = F, blank.lines.skip = T, nrows = 149)
mitop = mitop[2:ncol(mitop)]
colnames(mitop) = c("term","hierarchy","gene")
mitop$gene = gsub(" ","", mitop$gene)
head(mitop); tail(mitop)
```

```{r}
milong = separate_rows(mitop, gene, sep=",")
head(milong)
table(table(milong$hierarchy))[1:20]
length(unique(milong$term))
```


```{r}
head(unique(milong$hierarchy))
# only include terms that are not top-level in the heirarchy
milong_simp = milong[str_count(milong$hierarchy,">") > 1,]
milong_simp = milong_simp[ c("term","gene")]
head(table(milong_simp$term))
table(table(milong_simp$term))
length(unique(milong_simp$term))
```

```{r}
sig = read.delim(file="https://raw.githubusercontent.com/sarahhp/splicing_thermogenesis/main/03limma/any_and_all_donor_DGE.tsv")
head(sig)
nrow(sig)
tail(sig)

summary(sig$all.donors.adj.P.Val < 0.01)
```

```{r}
make_gene_list = function(sig, donor){
  gl = sig[,paste0("logFC.",donor)]
  names(gl) = sig$gene_name
  gl = gl[order(gl, decreasing = T)]
  return (gl)
}


allgl = list()
for (donor in c("s1","s2","s3","s4","s5","s6")){
  print(donor)
  gl = make_gene_list(sig, donor)
  allgl[[donor]] = gl
}
str(allgl)
```

```{r}
allea = compareCluster(allgl, fun=GSEA, TERM2GENE=milong_simp, pvalueCutoff=1, minGSSize=3, eps=1e-20)
allea = arrange(allea, p.adjust)
head(allea)
summary(allea@compareClusterResult$p.adjust)
nrow(allea)
```

```{r}
write.table(as.data.frame(allea), here(odir, "mitopathways_GSEA.tsv"),
            sep="\t", quote=F, row.names = F)
```


Alpha matters here for the includeAll plots

## Alpha 0.01
```{r}
filt = filter(allea, p.adjust < 0.01) %>% arrange(p.adjust)
```

NES = Normalised enrichment score

```{r}
dotplot(filt, showCategory=5, includeAll=T)
dotplot(filt, showCategory=5, includeAll=F)
dotplot(filt, showCategory=3, includeAll=T, color="NES") + 
  #scale_colour_gradientn(colours = c("beige", "#FF7F00")) 
  scale_colour_distiller(palette="Oranges", direction=1) + theme_classic(base_size=15) + 
  theme(axis.title.y=element_blank(),axis.text.y = element_text(size=12),
        axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25)) + xlab("Subject")
ggsave(here(figdir,"MitoPathways_GSEA_top3filt.pdf"))
```

## logFC heatmaps

```{r}
logfc = sig[,c(2,grep("logFC\\.s", colnames(sig)))]
head(logfc)
rownames(logfc) = logfc$gene_name
logfc$gene_name = NULL
#logfc =t(data.matrix(logfc) )
logfc[1:6,1:6]
```


```{r}
oxphos_subunits = milong[grep("C.* subunits", milong$hierarchy),]
library(ComplexHeatmap)
oxphos_subunits = oxphos_subunits[oxphos_subunits$gene %in% rownames(logfc),]#expr only
head(oxphos_subunits)
table(oxphos_subunits$term)

oxphos_logfc = logfc[oxphos_subunits$gene,]
oxphos_logfc
```


```{r}
library(circlize)
col_fun = colorRamp2(c(-2,0,2),c("#1F78B4","beige", "#FF7F00"))


Heatmap(t(data.matrix(oxphos_logfc)), column_split = gsub(" subunits" ,"",oxphos_subunits$term), 
        name="logfc", col=col_fun)
```

Differentially expressed only
```{r}
all.donor.sig = filter(sig, all.donors.adj.P.Val < 0.01) 
oxphos_logfc = oxphos_logfc[rownames(oxphos_logfc) %in% all.donor.sig$gene_name,]
nrow(oxphos_logfc)
pdf(here(figdir, "OXPHOS_complexes_heatmap.pdf"), height=4)
h= Heatmap(t(data.matrix(oxphos_logfc)), name="logfc",col=col_fun,
        column_split = gsub(" subunits" ,"",oxphos_subunits$term[oxphos_subunits$gene %in% all.donor.sig$gene_name]),
        column_names_gp = gpar(fontsize = 6)
        )
h 
dev.off()
h
```

## Non-canonical thermogensis logFC 
FYI any duplicate genes between groups get a ".1" added to the end of the gene name in order not to have duplicate column names.  
```{r}
creatine_pathway = c("CKB","CKMT1B","CKMT2","GAMT","GATM","CKMT1A", "ALPL") #"CKM" not expr
noncan = milong[grep("Fatty acid oxidation|Branched-chain.* metabolism|TCA cycle", milong$term),]
table(noncan$term)
noncan = noncan[noncan$gene %in% rownames(logfc),] # remove non expr
table(noncan$term)
creatine_pathway %in% rownames(logfc)
```

```{r}
noncan_logfc = logfc[c(noncan$gene, creatine_pathway),]

Heatmap(t(data.matrix(noncan_logfc)), name="logfc",col=col_fun,
        column_split = gsub("chain","chain\n",c(noncan$term, rep("Creatine", length(creatine_pathway)))),
        column_title_rot = 15, column_title_gp = gpar(fontsize = 10),)
```

```{r}
noncan_logfc = noncan_logfc[gsub("\\.(1|2)","",rownames(noncan_logfc)) %in% all.donor.sig$gene_name,]
nrow(noncan_logfc)
noncan = noncan[noncan$gene %in% all.donor.sig$gene_name,]
nrow(noncan)
creatine_pathway = creatine_pathway[creatine_pathway %in% all.donor.sig$gene_name]
length(creatine_pathway)

noncan_labels = gsub("\\.(1|2)","",rownames(noncan_logfc)) 
dups = noncan_labels[duplicated(noncan_labels)]
noncan_labels[noncan_labels %in% dups] = paste0( "'",dups)

pdf(here(figdir,"NonCanonicalTherm_heatmap.pdf"), height=4)
h = Heatmap(t(data.matrix(noncan_logfc)), name="logfc", col=col_fun,
        column_split =  gsub("chain","chain\n",c(noncan$term, rep("Creatine", length(creatine_pathway)))),
        column_title_rot = 15, column_title_gp = gpar(fontsize = 12),
        column_labels = noncan_labels,
        column_names_gp = gpar(fontsize=6))
h 
dev.off()
h

```


### Average expression plots?  
average logFC across genes is a bit tricky, trends towards zero... but easier to show than two expression values.  Though really it makes sense to average fpkm then calculate logFC
