---
title: "bulk_day15_mitopathways"
author: "Sarah Hazell Pickering"
date: "2024-09-16"
output: html_document
---

```{r libs, message=FALSE, warning=FALSE}
library(ComplexHeatmap)
library(circlize)
library(clusterProfiler)
library(dplyr)
library(edgeR)
library(enrichplot)
library(ggplot2)
library(ggpubr)
library(here)
library(stringr)
library(tidyr)
library(viridis)
```

```{r params}
rmd_filename = "bulk_day15_mitopathways"
module = "bulk"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
dataname = "day15"
load_data = T
subset_data = F
odir = here('output',module, dataname)

```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#DA1819", "#00AED1", "#434D51", "#EBB400", "#691A93", "#3AAD00")
clusters = ggsci::pal_simpsons()(12)[c(1:7, 9:15)]
```


```{r}
mitop = read.delim(here("data/other_papers/Human.MitoCarta3.0.csv"), sep=",", stringsAsFactors = F, blank.lines.skip = T, nrows = 149)
mitop = mitop[2:ncol(mitop)]
colnames(mitop) = c("term","hierarchy","gene")
mitop$gene = gsub(" ","", mitop$gene)
head(mitop); tail(mitop)
```

```{r}
milong = separate_rows(mitop, gene, sep=",")
head(milong)
table(table(milong$hierarchy))[1:20]
length(unique(milong$term))
```


```{r}
head(unique(milong$hierarchy))
# only include terms that are not top-level in the heirarchy
milong_simp = milong[str_count(milong$hierarchy,">") > 1,]
milong_simp = milong_simp[ c("term","gene")]
head(table(milong_simp$term))
table(table(milong_simp$term))
length(unique(milong_simp$term))
```

```{r}
sig = read.delim(file="https://raw.githubusercontent.com/sarahhp/splicing_thermogenesis/main/03limma/any_and_all_donor_DGE.tsv")
head(sig)
nrow(sig)
tail(sig)

summary(sig$all.donors.adj.P.Val < 0.01)
```

```{r}
make_gene_list = function(sig, donor){
  gl = sig[,paste0("logFC.",donor)]
  names(gl) = sig$gene_name
  gl = gl[order(gl, decreasing = T)]
  return (gl)
}


allgl = list()
for (donor in c("s1","s2","s3","s4","s5","s6")){
  print(donor)
  gl = make_gene_list(sig, donor)
  allgl[[donor]] = gl
}
str(allgl)
```

```{r}
allea = compareCluster(allgl, fun=GSEA, TERM2GENE=milong_simp, pvalueCutoff=1, minGSSize=3, eps=1e-20)
allea = arrange(allea, p.adjust)
head(allea)
summary(allea@compareClusterResult$p.adjust)
nrow(allea)
```

```{r}
write.table(as.data.frame(allea), here(odir, "mitopathways_GSEA.tsv"),
            sep="\t", quote=F, row.names = F)
```




## GSEA plots 

IncludeAll, not just the highest ranked terms. 
```{r}
filt = filter(allea, p.adjust < 0.01) %>% arrange(p.adjust)
top3_gsea = slice_min(as_tibble(filt), p.adjust, n=3, by = Cluster)
```

NES = Normalised enrichment score

```{r MitoPathways_GSEA_top3filt, dev = c("png","pdf")}
dotplot(filt, showCategory=5, includeAll=T)
dotplot(filt, showCategory=5, includeAll=F)
dotplot(filt, showCategory=3, includeAll=T, color="NES") + 
  #scale_colour_gradientn(colours = c("beige", "#FF7F00")) 
  scale_colour_distiller(palette="Oranges", direction=1) + 
  theme(axis.title.y=element_blank(),axis.text.y = element_text(size=12),
        axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25)) + xlab("Subject")
```

## logFC heatmaps

```{r}
logfc = sig[,c(2,grep("logFC\\.s", colnames(sig)))]
head(logfc)
rownames(logfc) = logfc$gene_name
logfc$gene_name = NULL
#logfc =t(data.matrix(logfc) )
logfc[1:6,1:6]
```


```{r}
oxphos_subunits = milong[grep("C.* subunits", milong$hierarchy),]
oxphos_subunits = oxphos_subunits[oxphos_subunits$gene %in% rownames(logfc),]#expr only
head(oxphos_subunits)
table(oxphos_subunits$term)

oxphos_logfc = logfc[oxphos_subunits$gene,]
oxphos_logfc[1:10,1:6]
```


```{r}
library(circlize)
col_fun = colorRamp2(c(-2,0,2),c("#1F78B4","beige", "#FF7F00"))


Heatmap(t(data.matrix(oxphos_logfc)), column_split = gsub(" subunits" ,"",oxphos_subunits$term), 
        name="logfc", col=col_fun)
```

Differentially expressed only
```{r dev = c("png","pdf")}
all.donor.sig = filter(sig, all.donors.adj.P.Val < 0.01) 
oxphos_logfc = oxphos_logfc[rownames(oxphos_logfc) %in% all.donor.sig$gene_name,]
nrow(oxphos_logfc)
pdf(here(fig_dir, "OXPHOS_complexes_heatmap.pdf"), height=4)
h= Heatmap(t(data.matrix(oxphos_logfc)), name="logfc",col=col_fun,
        column_split = gsub(" subunits" ,"",oxphos_subunits$term[oxphos_subunits$gene %in% all.donor.sig$gene_name]),
        column_names_gp = gpar(fontsize = 6)
        )
h 
dev.off()
h
```

## Non-canonical thermogensis logFC 
FYI any duplicate genes between groups get a ".1" added to the end of the gene name in order not to have duplicate column names.  
```{r}
creatine_pathway = c("CKB","CKMT1B","CKMT2","GAMT","GATM","CKMT1A", "ALPL") #"CKM" not expr
noncan = milong[grep("Fatty acid oxidation|Branched-chain.* metabolism|TCA cycle", milong$term),]
table(noncan$term)
noncan = noncan[noncan$gene %in% rownames(logfc),] # remove non expr
table(noncan$term)
creatine_pathway %in% rownames(logfc)
```

```{r}
noncan_logfc = logfc[c(noncan$gene, creatine_pathway),]

Heatmap(t(data.matrix(noncan_logfc)), name="logfc",col=col_fun,
        column_split = gsub("chain","chain\n",c(noncan$term, rep("Creatine", length(creatine_pathway)))),
        column_title_rot = 15, column_title_gp = gpar(fontsize = 10),)
```

```{r}
noncan_logfc = noncan_logfc[gsub("\\.(1|2)","",rownames(noncan_logfc)) %in% all.donor.sig$gene_name,]
nrow(noncan_logfc)
noncan = noncan[noncan$gene %in% all.donor.sig$gene_name,]
nrow(noncan)
creatine_pathway = creatine_pathway[creatine_pathway %in% all.donor.sig$gene_name]
length(creatine_pathway)

noncan_labels = gsub("\\.(1|2)","",rownames(noncan_logfc)) 
dups = noncan_labels[duplicated(noncan_labels)]
noncan_labels[noncan_labels %in% dups] = paste0( "'",dups)

pdf(here(fig_dir,"NonCanonicalTherm_heatmap.pdf"), height=4)
h = Heatmap(t(data.matrix(noncan_logfc)), name="logfc", col=col_fun,
        column_split =  gsub("chain","chain\n",c(noncan$term, rep("Creatine", length(creatine_pathway)))),
        column_title_rot = 15, column_title_gp = gpar(fontsize = 12),
        column_labels = noncan_labels,
        column_names_gp = gpar(fontsize=6))
h 
dev.off()
h

```

## Test significance of GSEA differences

### Get logFC for each replicate

```{r}
#load preproceesed object called filt
load(url("https://github.com/sarahhp/splicing_thermogenesis/raw/refs/heads/main/03limma/DGElist_ob_limma_filt.RData"))
```

```{r}
make_rep_gene_list = function(filt, donor, rep, ps=0.5){
  expr = rpkm(filt, normalized.lib.sizes = T)
  w = expr[,filt$samples$donor == donor & filt$samples$rep == rep & filt$samples$condition == "white"]
  b = expr[,filt$samples$donor == donor & filt$samples$rep == rep & filt$samples$condition == "beige"]
  gl = log2((b+ps)/(w+ps))
  names(gl) = filt$genes$gene_name
  gl = gl[order(gl, decreasing = T)]
  return (gl)
}
head(make_rep_gene_list(filt, "subject2", "3",ps=0.5))

allgl = list()
for (donor in c("s1","s2","s3","s4","s5","s6")){
  for (rep in 1:3){
    gl = make_rep_gene_list(filt, gsub("s","subject",donor), rep)
    allgl[[paste(donor, rep, sep="_")]] = gl
  }
#print(donor)
}
str(allgl)
```

### GSEA per replicate

Differentiation replicates give an estimate of the variation in enrichment scores per subject. 
```{r}
repea = compareCluster(allgl, fun=GSEA, TERM2GENE=milong_simp, pvalueCutoff=1, minGSSize=3, eps=1e-20)
repea = arrange(repea, p.adjust)
head(repea, n=15)
summary(repea@compareClusterResult$p.adjust)
summary(repea@compareClusterResult$NES)
nrow(repea)
```


```{r}
dfrep = as.data.frame(repea) %>% mutate(subject = gsub("_.*","", Cluster), rep = gsub(".*_","", Cluster)) %>%
  filter(ID %in% top3_gsea$ID)
head(dfrep)
```

## Pairwise t-tests
This is necessary for seeing differences when 2 or more subjects have a similar pattern of enrichment; in the limma test below I set up an s v mean(s) which enriches for unique NES between subjects.  This satisifies me that there are differences in pathway enrichment between donors, in particular between oxphos components, creatine metabolism and BCAA metabolism.  
```{r}
enrich= compare_means(enrichmentScore ~ subject, method="t.test", data=dfrep, group="ID", p.adjust.method = "fdr") %>%
  arrange(p.adj)
enrich 
```


```{r}
summary(aov(NES~subject+ID, data=dfrep))
t.test(filter(dfrep, ID== "CV subunits" & subject == "s2") %>% pull(NES),
filter(dfrep, ID== "CV subunits" & subject != "s2") %>% pull(NES))

t.test(filter(dfrep, ID== "CV subunits" & subject == "s2") %>% pull(NES),
filter(dfrep, ID== "CV subunits" & subject == "s6") %>% pull(NES))

t.test(filter(dfrep, ID== "CV subunits" & subject == "s2") %>% pull(NES),
filter(dfrep, ID== "CV subunits" & subject == "s4") %>% pull(NES))

filter(dfrep, ID== "CV subunits")
```



```{r}
ggplot(filter(dfrep, ID== "CV subunits"), aes(x=subject, y=NES)) + geom_boxplot()
ggplot(filter(dfrep, ID== "CI subunits"), aes(x=subject, y=NES)) + geom_boxplot()
ggplot(filter(dfrep, ID== "Creatine metabolism"), aes(x=subject, y=NES)) + geom_boxplot()
ggplot(filter(dfrep, ID== "CIII subunits"), aes(x=subject, y=NES)) + geom_boxplot()
ggplot(filter(dfrep, ID== "Fatty acid oxidation"), aes(x=subject, y=NES)) + geom_boxplot()
ggplot(filter(dfrep, ID== "Branched-chain amino acid metabolism"), aes(x=subject, y=NES)) + geom_boxplot()
ggplot(filter(dfrep, ID== "Mitochondrial ribosome"), aes(x=subject, y=NES)) + geom_boxplot()
```

```{r}
head(dfrep)
```

## Limma test

```{r}
enrichmat = filter(as_tibble(repea), p.adjust < 0.05) %>% pivot_wider(names_from = Cluster, values_from = NES, id_cols=ID, values_fill = 0) %>% as.data.frame()
rownames(enrichmat) = enrichmat$ID
enrichmat = data.matrix(enrichmat[2:ncol(enrichmat)])
head(enrichmat)

sets =  repea@compareClusterResult[c("ID","setSize")] %>% distinct() %>% as.data.frame()
rownames(sets) = sets$ID
sets = sets[rownames(enrichmat),]
head(sets)

samples = select(dfrep, Cluster, subject, rep) %>% distinct()
rownames(samples) = samples$Cluster
samples$subject = factor(samples$subject)
samples
design = model.matrix(~0+subject, data = samples)
colnames(design) = levels(samples$subject)
design
```


```{r}
fit = lmFit(enrichmat, design)
explong = makeContrasts(s1 = s1 - (s2 + s3 + s4 + s5 + s6)/5 ,
                        s2 = s2 - (s1 + s3 + s4 + s5 + s6)/5 ,
                        s3 = s3 - (s1 + s2 + s4 + s5 + s6)/5 ,
                        s4 = s4 - (s1 + s2 + s3 + s5 + s6)/5 ,
                        s5 = s5 - (s1 + s2 + s3 + s4 + s6)/5 ,
                        s6 = s6 - (s1 + s2 + s3 + s4 + s5)/5 ,
                         levels = design)

next_fit = contrasts.fit(fit, explong)
#Becomes a weighted contrast table
next_fit$contrasts
plot(sets$setSize, fit$sigma)
fit = eBayes(next_fit, robust=T,  trend=sets$setSize)
results = decideTests(fit, p.value = 0.05)
summary(results)
```



```{r}
as.data.frame(results)
results[unique(top3_gsea$ID),]
```

