---
title: "bulk_day15_DGE_between_subjects"
author: "Sarah Hazell Pickering"
date: "2025-01-21"
output: html_document
---

Preprocessing as in: https://github.com/sarahhp/splicing_thermogenesis/blob/main/R/01_six_donor_DGE_limma.Rmd

```{r}
library(here); i_am("analysis/bulk_day15_mitopathways.Rmd")
knitr::opts_knit$set(dev=c("png","pdf"))
odir = here("output/bulk")
```

```{r message=FALSE, warning=FALSE, echo=TRUE}
library(tidyr)
library(readr)
library(dplyr)
library(edgeR)
library(limma)
library(RUVSeq)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(stringr)
library(viridis)
```


```{r themes}
theme_set(theme_classic(base_size=15) + 
  theme(axis.title.y=element_blank(),axis.text.y = element_text(size=12),
        axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25)))

pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
```


```{r}
#load preproceesed object called filt
load(url("https://github.com/sarahhp/splicing_thermogenesis/raw/refs/heads/main/03limma/DGElist_ob_limma_filt.RData"))
```

### voom normlisation

```{r}
norm = voom(filt, plot=T)
```

## Limma

```{r}
head(norm$design)
#set some factor levels
norm$targets$group = paste(norm$targets$condition, norm$targets$donor, sep=".") #explicit
#simplify names for intercepting
norm$targets$group = gsub("subject","s", norm$targets$group)
norm$targets$group = factor(gsub("eige|hite","", norm$targets$group))
norm$targets$group
design = model.matrix(~0 +frac_assigned_to_genes + group, data=norm$targets)
colnames(design) = c("frac_assigned_to_genes",levels(norm$targets$group))
head(design)
```

```{r}
explicit_fit = lmFit(norm, design=design)
# Explicitly writing out contrasts
explong = makeContrasts(s1_uniq = (b.s1 - w.s1) - (((b.s2 - w.s2)+(b.s3 - w.s3)+ (b.s4 - w.s4)+(b.s5 - w.s5)+ (b.s6 - w.s6))/5),
                        s2_uniq = (b.s2 - w.s2) - (((b.s1 - w.s1)+(b.s3 - w.s3)+ (b.s4 - w.s4)+(b.s5 - w.s5)+ (b.s6 - w.s6))/5),
                        s3_uniq = (b.s3 - w.s3) - (((b.s1 - w.s1)+(b.s2 - w.s2)+ (b.s4 - w.s4)+(b.s5 - w.s5)+ (b.s6 - w.s6))/5),
                        s4_uniq = (b.s4 - w.s4) - (((b.s1 - w.s1)+(b.s2 - w.s2)+ (b.s3 - w.s3)+(b.s5 - w.s5)+ (b.s6 - w.s6))/5),
                        s5_uniq = (b.s5 - w.s5) - (((b.s1 - w.s1)+(b.s2 - w.s2)+ (b.s3 - w.s3)+(b.s4 - w.s4)+ (b.s6 - w.s6))/5),
                        s6_uniq = (b.s6 - w.s6) - (((b.s1 - w.s1)+(b.s2 - w.s2)+ (b.s3 - w.s3)+(b.s4 - w.s4)+ (b.s5 - w.s5))/5),
                        background = frac_assigned_to_genes,
                      levels = design)
next_fit = contrasts.fit(explicit_fit, explong)
#Becomes a weighted contrast table
next_fit$contrasts
efit = eBayes(next_fit, robust=T)
results= decideTests(efit, method="separate", adjust.method = "BH", p.value = 0.05)
summary(results)
```

```{r}
vennDiagram(results[,c(1,2,4,6)])
vennDiagram(results[,c(1,2,4,6)])
vennDiagram(results[,c(2,6,7)])
vennDiagram(results[,c(2,6,7)], include="up")
vennDiagram(results[,c(2,6,7)], include="down")
vennDiagram(results[,c(1,4,7)])
vennDiagram(results[,c(1,2,4)], include = "up")
vennDiagram(results[,c(1,2,4)], include = "down")
vennDiagram(results[,c(1,2,6,7)])
```

### Subject tables

```{r}
coefs = colnames(efit$contrasts)
subjecttabs = list()

base_matrix = matrix(NA, dimnames = list(efit$genes$gene_name,coefs), 
                                     nrow = nrow(efit$genes), ncol=length(coefs))
subjectlists = list(adj.p.val = base_matrix, logFC = base_matrix,  
                  genes= efit$genes )
for ( co in coefs){
  onesubject = topTable(efit, number = Inf, adjust.method = "BH", 
                    coef = co)
  subjecttabs[[co]] = onesubject
  subjectlists$adj.p.val[,co] = onesubject$adj.P.Val[order(onesubject$Geneid)]
  subjectlists$logFC[,co] = onesubject$logFC[order(onesubject$Geneid)]
}
summary(subjecttabs)
```

```{r}
ann_cols = c("Geneid", "gene_name", "description")
alltab = efit$genes[,!colnames(efit$genes) %in%c("ensembl_gene_id_version", "ID","Length")]
for (d in coefs){
  formatted = subjecttabs[[d]]
  formatted = formatted[,c(ann_cols,"logFC","P.Value","adj.P.Val","AveExpr")]
  alltab = merge(alltab, formatted, by=ann_cols,
                 suffixes = c("",paste(".", d, sep="")))
  alltab = alltab[!duplicated(alltab$gene_name),] #remove duplicates because they tend to propogate on merges
}
head(alltab); nrow(alltab)
# add .s1_uniq to the first occurence
alltab = dplyr::rename(alltab, P.Value.s1_uniq = "P.Value", adj.P.Val.s1_uniq = "adj.P.Val", AveExpr.s1_uniq = "AveExpr", logFC.s1_uniq= logFC)
summary(rowSums(alltab[,grep("adj.P.Val\\.s", colnames(alltab))] < 0.05) == 6)
```




```{r}
arrange(alltab, adj.P.Val.s2_uniq) %>% select(ann_cols, contains("s2_uniq"))
arrange(alltab, adj.P.Val.s2_uniq) %>% filter(logFC.s2_uniq > 0) %>% select(ann_cols, contains("s2_uniq"))
```


Titin is a giant protein, the largest in the human body, primarily found in striated muscle (skeletal and cardiac). very low expressed so probably not a major constituent.  

```{r}
arrange(alltab, adj.P.Val.s6_uniq) %>% select(ann_cols, contains("s6_uniq"))
arrange(alltab, adj.P.Val.s6_uniq) %>% filter(logFC.s6_uniq > 0) %>% select(ann_cols, contains("s6_uniq"))
```

## overlap term
```{r}
summary(alltab$adj.P.Val.s2_uniq < 0.05)
sum_results = rowSums(results)
```

Think about how to represent/analyse further.  GO or GSEA per donor, probably also add in some overlap term.  


```{r}
write.table(alltab, here(odir, "DGE_between_subjects.tsv"),
            sep="\t", quote=F, row.names = F)
```



## GO setup
```{r}
go <- clusterProfiler::read.gmt("/projects/imb-pkbphil/sp/annotations/molsigdb/msigdb.v2024.1.Hs.symbols.gmt")
molsig = go[gsub("_.*","", go$term) %in% c("GOCC","GOBP","GOMF","KEGG","HALLMARK","WP","REACTOME"),]

head(molsig); nrow(molsig)
molsig$term = factor(molsig$term)
table(gsub("_.*","", molsig$term))
```


```{r}
make_gene_vector = function(sig, donor, direction="up"){
  sig = sig[sig[,paste0("adj.P.Val.",donor)] < 0.05,]
  if (direction == "up"){
    gv = sig$gene_name[sig[,paste0("logFC.",donor)] > 0 ]
  } else if (direction == "down"){
    gv = sig$gene_name[sig[,paste0("logFC.",donor)] < 0 ]
  }else {print("Error: direction option should be either \"up\" or \"down\"")}
  return (gv)
}

allgv = list()
for (donor in c(coefs)){
  print(donor)
  up = make_gene_vector(alltab, donor, direction="up")
  down = make_gene_vector(alltab, donor, direction="down")
  if (length(up) > 0 ){
    allgv[[paste0(donor, ".up")]] = up
  }
  if (length(down) >0){
    allgv[[paste0(donor, ".down")]] = down
  }
}
str(allgv)
```

## Run GO

```{r}
allgo = compareCluster(allgv, fun=enricher, TERM2GENE=molsig, minGSSize=3, universe = alltab$gene_name)
allgo = arrange(allgo, p.adjust, Count)
head(allgo)
summary(allgo@compareClusterResult$p.adjust)
nrow(allgo)
```


```{r}
filt = filter(allgo, p.adjust < 0.01) 
```

## GO  plot

```{r go5, dev=c("png","pdf")}
dotplot(filt, showCategory=6, includeAll=T) + theme(axis.text.x = element_text(angle=30,hjust=1) )
dotplot(filt, showCategory=2, includeAll=T)+ theme(axis.text.x = element_text(angle=30,hjust=1) )

dotplot(filt, showCategory=6, includeAll=F) + theme(axis.text.x = element_text(angle=30,hjust=1) )
```

```{r}
write.table(as.data.frame(allgo), here(odir, "DGE_between_subjects_GO.tsv"),
            sep="\t", quote=F, row.names = F)
```

## Top 200 genes per set

```{r}
make_gene_vector = function(sig, donor, direction="up"){
  sig = sig[sig[,paste0("adj.P.Val.",donor)] < 0.05,]
  sig = sig[order(sig[paste0("adj.P.Val.",donor)]),]
  if (direction == "up"){
    sig = sig[sig[,paste0("logFC.",donor)] > 0 ,]
    gv = slice_head(sig, n=200) %>% pull(gene_name)
  } else if (direction == "down"){
    sig = sig[sig[,paste0("logFC.",donor)] < 0 ,]
    gv = slice_head(sig, n=200) %>% pull(gene_name)
  }else {print("Error: direction option should be either \"up\" or \"down\"")}
  return (gv)
}

allgv = list()
for (donor in c(coefs)){
  print(donor)
  up = make_gene_vector(alltab, donor, direction="up")
  down = make_gene_vector(alltab, donor, direction="down")
  if (length(up) > 0 ){
    allgv[[paste0(donor, ".up")]] = up
  }
  if (length(down) >0){
    allgv[[paste0(donor, ".down")]] = down
  }
}
str(allgv)
```


## Run GO

```{r}
allgo = compareCluster(allgv, fun=enricher, TERM2GENE=molsig, minGSSize=3, universe = alltab$gene_name)
allgo = arrange(allgo, p.adjust, Count)
head(allgo)
summary(allgo@compareClusterResult$p.adjust)
nrow(allgo)
```


```{r}
filt = filter(allgo, p.adjust < 0.05) 
```

## GO  plot

```{r go5, dev=c("png","pdf")}
dotplot(filt, showCategory=6, includeAll=T) + theme(axis.text.x = element_text(angle=30,hjust=1) )
dotplot(filt, showCategory=2, includeAll=T)+ theme(axis.text.x = element_text(angle=30,hjust=1) )

dotplot(filt, showCategory=6, includeAll=F) + theme(axis.text.x = element_text(angle=30,hjust=1) )
```



## MitoPathways setup

```{r}
mitop = read.delim(here("data/other_papers/Human.MitoCarta3.0.csv"), sep=",", stringsAsFactors = F, blank.lines.skip = T, nrows = 149)
mitop = mitop[2:ncol(mitop)]
colnames(mitop) = c("term","hierarchy","gene")
mitop$gene = gsub(" ","", mitop$gene)
head(mitop); tail(mitop)
```

```{r}
milong = separate_rows(mitop, gene, sep=",")
head(milong)
table(table(milong$hierarchy))[1:20]
length(unique(milong$term))
```


```{r}
head(unique(milong$hierarchy))
# only include terms that are not top-level in the heirarchy
milong_simp = milong[str_count(milong$hierarchy,">") > 1,]
milong_simp = milong_simp[ c("term","gene")]
head(table(milong_simp$term))
table(table(milong_simp$term))
length(unique(milong_simp$term))
```

### mitP run 


```{r}
allmt = compareCluster(allgv, fun=enricher, TERM2GENE=milong_simp, minGSSize=3, pvalueCutoff=1, universe = alltab$gene_name)
allmt = arrange(allgo, p.adjust, Count)
head(allmt, n=20)
summary(allmt@compareClusterResult$p.adjust)
nrow(allmt)
```


## GSEA setup
```{r}
make_gene_list = function(sig, donor){
  gl = sig[,paste0("logFC.",donor)] * -log10(sig[,paste0("adj.P.Val.",donor)])
  names(gl) = sig$gene_name
  gl = gl[order(gl, decreasing = T)]
  return (gl)
}


allgl = list()
for (donor in c(coefs)){
  print(donor)
  gl = make_gene_list(alltab, donor)
  allgl[[donor]] = gl
}
str(allgl)
```

```{r}
allea = compareCluster(allgl, fun=GSEA, TERM2GENE=molsig, pvalueCutoff=1, minGSSize=3, eps=1e-40)
allea = arrange(allea, p.adjust, abs(NES))
head(allea)
summary(allea@compareClusterResult$p.adjust)
nrow(allea)
```

```{r}
filt = filter(allea, p.adjust < 0.01) 
```

NES = Normalised enrichment score
## GSEA  plot

```{r go, dev=c("png","pdf")}
dotplot(filt, showCategory=6, includeAll=T, color="NES") + scale_color_gradient2(high = "orange",low="cyan3") + theme(axis.text.x = element_text(angle=30,hjust=1) )
dotplot(filt, showCategory=2, includeAll=T, color="NES") + scale_color_gradient2(high = "orange",low="cyan3") + theme(axis.text.x = element_text(angle=30,hjust=1) )
```

