---
title: "initial_r2_only"
author: "Sarah Hazell Pickering"
date: "2024-08-26"
output: html_document
---

Quality control and Mt filtering for adipogenesis scRNAseq (gene-level). More specific analyses white_only and white_vs_beige come start from this seurat object.   

Cellranger was run on day1 and day3 samples in "read2 only" mode, to match the way the second replicate was sequenced.  I've just linked day0 from donor6 into the same directory, will replace if necessary, but since day0 cells are so different, it shouldn't make a difference that they were processed in PE mode.

Goal: to assess if R2 only mode removed the replicate differences at day 1 (it minimised them yes).  

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(here)
library(patchwork)
```

```{r params}
module = "adipogenesis"
dataname = "initial_r2_only"
load_data = F #load 10x counts on first run through
subset_data = F

rmd_filename = "adipogenesis_initial_r2_only"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
indir = here("data/cellranger/count_r2_only") #links atm; todo: link to GEO counts
odir = here("output", module, dataname) 
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4", "#B2DF8A","#85C2EA", "#B3B3B3","#F18486","#FDBF6F","#E31A1C")
clusters = tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
```




## Read data

```{r read_or_load}
samples = list.dirs(here(indir), full.names = F, recursive = F)

if (load_data==FALSE){# load count matrices
  
  data_list = c()
  
  for (s in samples){
    #if this has been run before delete bp dir
    bc_matrix = Read10X(data.dir = here(indir, s, "outs/filtered_feature_bc_matrix"))
    so = CreateSeuratObject(counts=bc_matrix, project=s, names.field = 2, names.delim = "-")
    write_matrix_dir(mat=so[["RNA"]]$counts, dir = here(odir, "bpcells", s))
    gene_mat = open_matrix_dir(here(odir, "bpcells", s))
    so[["RNA"]]$counts = gene_mat
    
    data_list = c(data_list,so)
  }
  names(data_list) = samples
  aggr = merge(data_list[[1]], y= data_list[2:length(data_list)], 
               add.cell.ids = gsub("_",".",samples), project=dataname )
  
  saveRDS(aggr,here(odir, "unfiltered_object.rds"))

  rm(data_list, bc_matrix, so)
} else if(load_data){
  aggr = readRDS(here(odir, "unfiltered_object.rds"))
}

aggr
head(aggr@meta.data) #auto metadata
```

```{r set_metadata}
aggr$orig.ident = gsub("\\.","_",gsub("_.*", "",rownames(aggr@meta.data)))
aggr$orig.ident = factor(aggr$orig.ident, levels = c("white_day3","white_day1_rep2","white_day1","asc_day0",
                                                     "beige_day1","beige_day1_rep2","beige_day3"))
Idents(aggr) = aggr$orig.ident
aggr$sample = aggr$orig.ident
table(aggr@meta.data$orig.ident)
aggr$condition = gsub("_.*","", aggr$orig.ident)
table(aggr$condition)

aggr$time = gsub(".*_", "", gsub("_rep2","",aggr$orig.ident))
table(aggr$time)
aggr$rep = "rep1"
aggr$rep[grep("_rep2", aggr$orig.ident)] = "rep2"
```

```{r}
aggr <- JoinLayers(aggr)
aggr
```

```{r}
if (subset_data){
  aggr = subset(aggr, downsample=1000)
  print(aggr)
}
```


```{r no_cells}
ggplot(aggr@meta.data) + geom_bar(aes(x=orig.ident, fill=orig.ident)) + scale_fill_manual(values=pal)
```
## Quality control and Filtering

Number of genes (features) detected per cell high, due to the depth of our sequencing and the stem-cell nature of the samples.  There is a wide spread of the number of genes (nFeature)and reads (nCount) per cell.  The percentage mitochondrial genes is super low for most cells, but some few have up to 80% mt.   

```{r before_filt1, dev=c("png","pdf")}
aggr[["percent.mt"]] = PercentageFeatureSet(aggr, pattern="^MT-")
VlnPlot(aggr, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0, cols=pal)
```


```{r before_filt2}
VlnPlot(subset(x = aggr, downsample = 1000), features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, cols=pal)
RidgePlot(aggr, feature= "nFeature_RNA", cols=pal) + geom_vline(aes(xintercept=2000))
```

```{r before_filt3}
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "percent.mt", cols=pal)
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", cols=pal)
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal)
```

We set an adaptive Mt.percent filter that's the 3* mad maximum. 

```{r}
mt_info = group_by(aggr@meta.data, sample) %>% 
  summarise(mad = mad(percent.mt),
            median = median(percent.mt), 
            thresh = median(percent.mt) + mad(percent.mt)*3,
            th95 = quantile(percent.mt, 0.95))
mt_info
thresh = max(mt_info$thresh)
```

```{r}
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal) + 
  geom_hline(aes(yintercept=thresh), linetype="dashed") +
  geom_vline(aes(xintercept=2000), linetype="dashed")
```


```{r filter}
ncol(aggr)
filt <- subset(aggr,  percent.mt < thresh & nFeature_RNA > 2000)
ncol(filt)

FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size=0.5, cols=pal)
FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size=0.5, cols=pal)
FeatureScatter(filt, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal) 
```


## Normalisation 

Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. This is then natural-log transformed using log1p
```{r lognorm}
filt= NormalizeData(filt, normalization.method = "LogNormalize", scale.factor = 10000)
filt#161Mb
```

```{r scale}
all.genes = rownames(filt)
filt = ScaleData(filt, features=all.genes)
```

```{r variable_features}
filt <- FindVariableFeatures(filt, selection.method = "vst", nfeatures = 2000)
filt #163Mb
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(filt), 10)

# plot variable features with and without labels
plot1 = VariableFeaturePlot(filt)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2

plot1 + coord_cartesian(y=c(0,5))
```

## PCA

```{r pca}
filt = RunPCA (filt, ndims.print=2)
DimPlot(filt, reduction = "pca", cols = pal)
FeaturePlot(filt, features= c("percent.mt","nCount_RNA","ADIPOQ","ACTA2"))
DimHeatmap(filt, dims = 1:6, cells = 500, balanced = TRUE)
```

### Determine the dimensionality

They encourage to complete downstream analyses with different numbers of PCs.  
```{r elbow}
ElbowPlot(filt, ndims = 50)
```

## Cluster
```{r cluster}
filt <- FindNeighbors(filt, dims = 1:20)
filt <- FindClusters(filt, resolution = 0.5)
table(filt@meta.data$seurat_clusters)
```

### Cluster composition

```{r cluster_composition}
composition = group_by(filt@meta.data, orig.ident,seurat_clusters) %>% count()
composition = composition[order(composition$n, decreasing = T),]
colnames(composition) = c("orig.ident","seurat_clusters","n_cells")
head(composition)

ggplot(composition) + geom_bar(stat="identity",aes(y=orig.ident, x=n_cells, fill=seurat_clusters)) + 
  scale_y_discrete(limits=rev) + scale_fill_manual(values=clusters)
ggplot(composition) + geom_bar(stat="identity",aes(y=orig.ident, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=clusters)
ggplot(composition) + geom_bar(stat="identity",aes(fill=orig.ident, x=n_cells, y=seurat_clusters)) +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)
ggplot(composition) + geom_bar(stat="identity",aes(fill=orig.ident, x=n_cells, y=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev) + scale_fill_manual(values=pal)
```

## UMAP
```{r umap}
filt = RunUMAP(filt, dims=1:20)
DimPlot(filt, reduction="umap", label=T, shuffle = T, cols=clusters)
DimPlot(filt, reduction="umap", split.by = "orig.ident", ncol=3, shuffle = T,  cols=clusters)
DimPlot(filt, group.by="orig.ident", cols=pal, shuffle = T)
```

## t-SNE
```{r tsne}
filt = RunTSNE(filt, reduction="pca", dims=1:30)
DimPlot(filt, reduction = "tsne",shuffle=T,  cols=clusters)
DimPlot(filt, reduction = "tsne", group.by= "orig.ident", shuffle = T, cols=pal)
```


## Find biomarkers
Min.pct sets the minimum percentage of cells for a marker to be found in

Does densify, which increases speed at the cost of memory help?  
```{r densify}
norm.markers = FindAllMarkers(filt, only.pos = T, min.pct=0.25, 
                              logfc.threshold = 0.5, densify=T)
head(norm.markers); dim(norm.markers)

top.markers = as.data.frame(norm.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC))
top.markers
```
 

Visualisation methods

```{r marker_vis}
VlnPlot(filt, features = c("ACTA2","MKI67"))
FeaturePlot(filt, features = c("ACTA2","MKI67","MT-ATP8","CHI3L1")) 

for (i in 1:(length(top.markers$gene)/4)){
  start = ((i-1)*4)+1; print(start)
  end = i*4; print(end)
  print(FeaturePlot(filt, features = top.markers$gene[start:end]))
}

FeaturePlot(filt, features = c("percent.mt", "nCount_RNA","nFeature_RNA"))
FeaturePlot(filt, features = c("PC_1","PC_2","PC_3","PC_4"))
```


```{r}
norm.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(subset(filt,downsample=100), features= top10$gene)
```

And then we would usually assign cell type identity.  

Be useful to add GO terms here/  Be a little more selective about the marker genes.  Some of the logFC choices seem to favour high cell-cell variability genes


## Cluster GO

### set up

```{r load_go}
molsig <- clusterProfiler::read.gmt(here("data/gene_sets/msigdb.v2024.1.Hs.symbols.gmt"))
head(molsig); nrow(molsig)
##Pick gene sets to look in bcos all is a lot
prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))
```

```{r}
shorten = function(ont) {
  abbreviate(gsub("_"," ", tolower(ont)),minlength=40, dot=T, named = F)
}
```

### Run 

```{r}
norm.markers$cluster = factor(norm.markers$cluster, levels =0:max(as.integer(norm.markers$cluster)))
all.genes = rownames(filt)
```

```{r run_go}
clgo = compareCluster(gene~cluster, data=norm.markers, 
                      fun="enricher", TERM2GENE = some.molsig, universe = all.genes)
clgo = mutate(clgo, cluster = factor(cluster, levels=0:max(as.integer(as.character(cluster)))),
              abbrev_ID = shorten(ID))
head(clgo)[,c(2,4:10)]
```

### GO Plots 
```{r go_plots}
p = dotplot(clgo, x="cluster", showCategory=2, includeAll=F) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))

p = dotplot(filter(clgo,!grepl("^(WP|KEGG|REACTOME)_",ID)), x="cluster", showCategory=2, includeAll=F) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))

p = dotplot(clgo, x="cluster", showCategory=2) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))
```


```{r save}
if (subset_data == F){
  saveRDS(filt, file = here(odir, "complete_analysis.rds"))
  write.table(norm.markers, here(odir, "marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(clgo, here(odir, "ORA_marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(composition, here(odir, "cluster_composition.txt"),
              sep="\t", row.names = F)
  Idents(filt) = "orig.ident"
  sbs = subset(filt, downsample=1000)
  saveRDS(sbs, file = here(odir, "10%_complete_analysis.rds"))
} else if (subset_data){
  saveRDS(filt, file = here(odir, "10%_subset_analysis.rds"))
}
```





