---
title: "beige_vs_white_downsample_annotate_w_miranda"
author: "Sarah Hazell Pickering"
date: "2025-08-26"
output: html_document
---

Using Azimuth to transfer cell identities found in miranda et al 2025 to the beige and white timecourse.  

To do: add the emont counts so we can integrate those as well


```{r message=FALSE, warning=FALSE}
library(Azimuth)
library(Seurat)
library(BPCells)
library(tidyr)
library(ggplot2)
```

```{r}
module="adipogenesis"
qdataname = "beige_vs_white_downsample"
library(here); i_am(file.path("analysis", module,paste0(qdataname,"_annotate_w_miranda.Rmd")))
odir = here("output", module,qdataname)

subset_data =  F
load_data = T #Load prebuilt Azimuth reference
miranda_data = "/home/sarahhp/home/sc_rnaseq/other_data/miranda2025"
```

```{r themes}
pal = c("#1F78B4", "#85C2EA","#B3B3B3", "#F18486","#E31A1C")
other= c(rev(paletteer::paletteer_d("MetBrewer::Redon")),  RColorBrewer::brewer.pal("Dark2", n=8))
clusters = tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
theme_set(theme_classic(base_size=15) + 
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25)))
```

## Set reference dir

```{r}
azdir = here(miranda_data, "azimuth_reference")
if (subset_data){
  azdir = here(azdir, "10%_subset")
}
```

## Load Query Datasets

```{r}
if (subset_data){
  qso = readRDS(here(odir, "10%_complete_analysis.rds"))
  k.weight=5
} else {
  qso = readRDS(here(odir, "complete_analysis.rds"))
  k.weight=50
}
qso
```

```{r}
DimPlot(qso, group.by = "seurat_clusters") + scale_colour_manual(values = clusters)
DimPlot(qso, group.by = "cond.time") + scale_colour_manual(values = pal)
```

## Map clusters

Not many anchors found for mapping :/ so we're adding a kweight limit to the Run Azimuth code aka. 
trace("RunAzimuth.Seurat", where=asNamespace("Azimuth"), edit=T)
adding kanchors=k.weight as a param to MappingScore on ln 107

```{r}
annot <- RunAzimuth(
    query = qso,
    reference = azdir,
    k.weight=k.weight)
```

```{r}
colnames(annot@meta.data)
table(annot$predicted.cell_state)
summary(annot$predicted.cell_state.score)
```


```{r}
names(other) = c(paste0("AD", 1:8), paste0("APC",1:5))
other[3] = "#F6768D"
other[8+1:5] = c("#809593","#6B3A4D","#1B4200","#4EC503","#3B5CFF")

DimPlot(annot, group.by = "predicted.cell_state", shuffle = T) + scale_colour_manual(values = other)
FeaturePlot(annot,"predicted.cell_state.score") + scale_colour_gradientn(colours =  spectral)
```


```{r}
DimPlot(annot,  group.by="predicted.BMI") + scale_colour_manual(values = spectral)
FeaturePlot(annot,  "predicted.BMI.score") + scale_colour_gradientn(colors=spectral)
```

```{r}
names(other) =NULL
DimPlot(annot, group.by = "predicted.condition", shuffle = T) + scale_colour_manual(values = other[c(1,3)])
FeaturePlot(annot,"predicted.condition.score") + scale_colour_gradientn(colours =  spectral)
```


```{r save}
if (subset_data == FALSE){
  saveRDS(annot, file = here(odir, "complete_analysis.annot.rds"))
  sub = subset(annot, downsample=1000)
  saveRDS(sub,here(odir,"10%_complete_analysis.annot.rds"))
} else if (subset_data){ #if subset, don't save complete analysis
  saveRDS(annot,here(odir, "10%_subset_analysis.annot.rds"))
}
```

