---
title: "white_only_downsample_downsample_lazarescu_adipo_only"
author: "Sarah Hazell Pickering"
date: "2025-08-26"
output: html_document
---


Using Azimuth to transfer Adipocyte and progenitor cell identities found in lazarescu et al 2025 to the beige and white timecourse.  

See publ data dir for reference generation.  

```{r libs, message=FALSE, warning=FALSE}
library(Azimuth)
library(BPCells)
library(Seurat)
library(ggplot2)
library(here)
library(tidyr)
```

```{r params}
module = "adipogenesis"
dataname = "white_only_downsample"
load_data = T #Load prebuilt Azimuth reference
subset_data = F

rmd_filename = "white_only_downsample_lazarescu_adipo"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))

qdataname = "white_only_downsample"
indir = here("output", module,dataname)
publ_data = here("data/other_papers/lazarescu2025")
odir = here(indir,"lazarescu")
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4", "#85C2EA","#B3B3B3", "#F18486","#E31A1C")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
other=ggsci::pal_npg()(7)
refclusters = c(rev(paletteer::paletteer_d("MetBrewer::Redon")),  RColorBrewer::brewer.pal("Dark2", n=8))
```

## Load Query Datasets

```{r}
if (subset_data){
  qso = readRDS(here(indir, "10%_complete_analysis.rds"))
  k.weight=5
} else {
  qso = readRDS(here(indir, "complete_analysis.rds"))
  k.weight=22
}
qso
```

```{r}
DimPlot(qso, group.by = "seurat_clusters", shuffle = T) + scale_colour_manual(values = clusters)
DimPlot(qso, group.by = "time", shuffle = T) + scale_colour_manual(values = pal)
```
## set reference dir
```{r}
azdir = here(publ_data, "azimuth_reference")
if (subset_data){
  azdir = here(azdir, "10%_subset")
}
```

## Map clusters

Not many anchors found for mapping :/ so we're adding a kweight limit to the Run Azimuth code aka. 
trace("RunAzimuth.Seurat", where=asNamespace("Azimuth"), edit=T)
adding kanchors=k.weight as a param to MappingScore on ln 107

```{r}
annot <- RunAzimuth(
    query = qso,
    reference = azdir,
    annotation.levels = "author_cell_type",
    k.weight=k.weight
    )
```

```{r}
colnames(annot@meta.data)
table(annot$predicted.author_cell_type)

summary(annot$predicted.author_cell_type.score)
```
## Plot Overlap

```{r}
DimPlot(annot, group.by = "predicted.author_cell_type", shuffle = T) + scale_colour_manual(values=refclusters)
FeaturePlot(annot,"predicted.author_cell_type.score") + scale_colour_gradientn(colours =  spectral)
```
## Add uncertain category

```{r filt_lazarescu_cell_type}
annot$filt.lazarescu.celltype = "uncertain"
annot$filt.lazarescu.celltype[annot$predicted.author_cell_type.score > 0.75] =  
  annot$predicted.author_cell_type[annot$predicted.author_cell_type.score > 0.75]

DimPlot(annot, group.by = "filt.lazarescu.celltype", shuffle = T) +
  scale_colour_manual(values = c("#00A087", "#E64B35","grey"))
```


```{r}
annot$filt.lazarescu.celltype = "uncertain"
annot$filt.lazarescu.celltype[annot$predicted.author_cell_type.score > 0.9] =  
  annot$predicted.author_cell_type[annot$predicted.author_cell_type.score > 0.9]

DimPlot(annot, group.by = "filt.lazarescu.celltype", shuffle = T) + 
  scale_colour_manual(values = c("#00A087", "#E64B35","grey"))
```


```{r filt_lazarescu_cell_type_90}
annot$filt.lazarescu.celltype = "uncertain"
annot$filt.lazarescu.celltype[annot$predicted.author_cell_type.score > 0.85] =  
  annot$predicted.author_cell_type[annot$predicted.author_cell_type.score > 0.85]

DimPlot(annot, group.by = "filt.lazarescu.celltype", shuffle = T) + 
  scale_colour_manual(values = c("#00A087", "#E64B35","grey"))
ggsave(here(fig_dir, "lazarescu_celltype_with_uncertainty.pdf"), device = "pdf")
ggsave(here(fig_dir, "lazarescu_celltype_with_uncertainty.png"), device = "png")
```

```{r}
VlnPlot(annot,"predicted.author_cell_type.score", group.by = "predicted.author_cell_type", pt.size = 0)
```



```{r save}
if (subset_data == FALSE){
  saveRDS(annot, file = here(odir, "complete_analysis.annot.rds"))
  sub = subset(annot, downsample=1000)
  saveRDS(sub,here(odir,"10%_complete_analysis.annot.rds"))
} else if (subset_data){ #if subset, don't save complete analysis
  saveRDS(annot,here(odir, "10%_subset_analysis.annot.rds"))
}
```

