---
title: "white_only_downsample"
author: "Sarah Hazell Pickering"
date: "2024-08-23"
output: html_document
---

Basic analysis of gene-level counts using seurat v5.  Using white samples only (day0, day1 (rep1+2) and day3) and downsampling so that we have the same number of cells per time point (half each from the two day1 replicates).  

Uses the rData files from initial_r2_only to save memory and shorten processing time.  

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(here)
library(patchwork)
```

```{r params}
module = "adipogenesis"
dataname = "white_only_downsample"
load_data = T
subset_data = F

rmd_filename = "white_only_downsample"
i_am(file.path("analysis", module, paste0(rmd_filename,".Rmd")))
odir = here("output", module, dataname)
indir = here("output", module, "initial_r2_only")
```


```{r setup}
set.seed(8)
fig_dir = here("analysis", module, "figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))

if (!dir.exists(odir)) dir.create(odir, recursive = TRUE)
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#B3B3B3","#85C2EA","#B2DF8A","#1F78B4")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```


## Read data

```{r}
if (subset_data){
  filt = readRDS(here(indir, "10%_complete_analysis.rds"))
  Idents(filt) = "seurat_clusters"
} else{
  filt = readRDS(here(indir, "complete_analysis.rds"))
}

Idents(filt) = "orig.ident"
table(Idents(filt))
filt = subset(filt, idents=c("asc_day0","white_day1", "white_day1_rep2","white_day3"))
filt$sample = factor(filt$sample, levels=c("asc_day0","white_day1", "white_day1_rep2","white_day3"))
filt$orig.ident = filt$sample
filt
```

```{r no_cells}
ggplot(filt@meta.data) + geom_bar(aes(x=orig.ident, fill=orig.ident)) + 
  scale_fill_manual(values=pal) + coord_cartesian(expand=F)
```

## Downsampling
so each timepoint has the same number of cells
This makes sure our clustering isn't biased towards timepoints with more cells. 
```{r}
sum(filt$sample == "asc_day0")
wd3 = subset(filt, idents=c("asc_day0","white_day3"), downsample =  sum(filt$sample == "asc_day0"))
wd3
d1 =  subset(filt, idents=c("white_day1","white_day1_rep2"), downsample =  sum(filt$sample == "asc_day0")/2)
d1

sbs = merge(x=wd3,y=d1)
sbs = JoinLayers(sbs)
sbs
table(sbs$time)
```

```{r}
ggplot(sbs@meta.data) + geom_bar(aes(x=orig.ident, fill=orig.ident)) + 
  scale_fill_manual(values=pal)+ coord_cartesian(expand=F)
```


## Normalisation 

Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. This is then natural-log transformed using log1p
```{r lognorm}
sbs= NormalizeData(sbs, normalization.method = "LogNormalize", scale.factor = 10000)
sbs#161Mb
```


```{r variable_features}
sbs <- FindVariableFeatures(sbs, selection.method = "vst", nfeatures = 2000)
sbs #163Mb
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(sbs), 10)

# plot variable features with and without labels
plot1 = VariableFeaturePlot(sbs)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```


```{r}
all.genes = rownames(sbs)
sbs = ScaleData(sbs, features=all.genes)
```

## PCA

```{r pca}
sbs = RunPCA (sbs, ndims.print=2)
DimPlot(sbs, reduction = "pca", cols = pal, shuffle = T) + theme_hp
DimPlot(sbs, reduction = "pca", cols = pal[c(1,2,4)], shuffle = T, group.by = "time")
FeaturePlot(sbs, features= c("percent.mt","nCount_RNA","CHI3L1","ACTA2"))
DimHeatmap(sbs, dims = 1:6, cells = 500, balanced = TRUE)
```

### Determine the dimensionality

```{r elbow}
ElbowPlot(sbs, ndims = 50)
```

## Cluster
```{r cluster}
sbs <- FindNeighbors(sbs, dims = 1:20)
sbs <- FindClusters(sbs, resolution = 0.5)
table(sbs@meta.data$seurat_clusters)
```

## Cluster composition

```{r cluster_composition}
composition = group_by(sbs@meta.data, orig.ident,seurat_clusters) %>% count()
composition = composition[order(composition$n, decreasing = T),]
colnames(composition) = c("orig.ident","seurat_clusters","n_cells")
#calculate proportions directly, then plot percentage just for the three cyc;ling cells
composition = group_by(composition, orig.ident) %>% mutate(cells_per_sample = n_cells/sum(n_cells),
                                                           time =gsub("(white|asc)_","", gsub("_rep2","",orig.ident)))
composition = group_by(composition, seurat_clusters) %>% mutate(cells_per_cluster = n_cells/sum(n_cells))
head(composition)
```


```{r plot_composition}
ggplot(composition) + geom_bar(stat="identity",aes(y=orig.ident, x=n_cells, fill=seurat_clusters)) + 
  scale_y_discrete(limits=rev) + scale_fill_manual(values=clusters)+
  coord_cartesian(expand = F)
ggplot(composition) + geom_bar(stat="identity",aes(y=orig.ident, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=clusters)+
  coord_cartesian(expand = F)
ggplot(composition) + geom_bar(stat="identity",aes(fill=orig.ident, x=n_cells, y=seurat_clusters)) +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)+
  coord_cartesian(expand = F)
ggplot(composition) + geom_bar(stat="identity",aes(fill=orig.ident, x=n_cells, y=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev) + scale_fill_manual(values=pal)+
  coord_cartesian(expand = F)
```

```{r composition_per_time, dev=c('png','pdf')}
ggplot(composition) + geom_bar(stat="identity",aes(y=time, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ 
  scale_fill_manual(values=clusters, labels= function (x){gsub("^","a",x)}, name="Adipogenesis\nclusters") +
  coord_cartesian(expand = F)
```


## UMAP
```{r umap}
sbs = RunUMAP(sbs, dims=1:20, n.neighbors = 20L)
DimPlot(sbs, reduction="umap", label=T, shuffle = T, cols=clusters) 
DimPlot(sbs, reduction="umap", split.by = "orig.ident", ncol=3, shuffle = T,  cols=clusters)
DimPlot(sbs, group.by="orig.ident", cols=pal, shuffle = T)
DimPlot(sbs, cols=pal[c(1,2,4)], shuffle = T, group.by="time")
```

## t-SNE
```{r tsne}
sbs = RunTSNE(sbs, reduction="pca", dims=1:20)
DimPlot(sbs, reduction = "tsne",shuffle=T,  cols=clusters) + theme_hp
DimPlot(sbs, reduction = "tsne", group.by= "orig.ident", shuffle = T, cols=pal)
DimPlot(sbs,  reduction = "tsne", cols=pal[c(1,2,4)], shuffle = T, group.by="time")
```


## Find biomarkers
Min.pct sets the minimum percentage of cells for a marker to be found in

Does densify, which increases speed at the cost of memory help?  
```{r densify}
norm.markers = FindAllMarkers(sbs, only.pos = T, min.pct=0.25, 
                              logfc.threshold = 0.5, densify=T)
head(norm.markers); dim(norm.markers)
norm.markers = filter(norm.markers, p_val_adj < 0.01)
table(norm.markers$cluster)

top.markers = as.data.frame(norm.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC))
top.markers
```
 
```{r top_markers, dev=c('png','pdf')}
DotPlot(sbs, features=rev(unique(top.markers$gene)), group.by = "seurat_clusters") & 
  scale_colour_gradientn(colours = spectral) & coord_flip()

top.markers$gene = factor(top.markers$gene, levels=rev(unique(top.markers$gene)))
overlap.to.plot = filter(top.markers, gene %in% top.markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(top.markers$gene))

ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point() + 
  theme(axis.text.x=element_text(angle=30, hjust=1))  + 
  scale_color_manual(values=clusters)
```

```{r}
include_all_plot = filter(norm.markers, gene %in% overlap.to.plot$gene & p_val_adj< 0.01)
include_all_plot$gene = factor(include_all_plot$gene, levels=levels(top.markers$gene))

a = ggplot(include_all_plot, aes(x=cluster, y=gene, colour=cluster, size=avg_log2FC)) + 
  geom_point()+  
  theme(axis.text.x=element_text(angle=30, hjust=1)) +scale_color_manual(values=clusters)
a


a = ggplot(include_all_plot, aes(x=cluster, y=gene, colour=cluster, size=pct.1)) + 
  geom_point()+  
  theme(axis.text.x=element_text(angle=30, hjust=1)) +scale_color_manual(values=clusters)
a
```



Visualisation methods

```{r marker_vis}
VlnPlot(sbs, features = c("MGP","ADIPOQ"))
FeaturePlot(sbs, features = c("MGP","ADIPOQ","MT-ATP8","ACTA2")) 
top.markers$gene = as.character(top.markers$gene)

for (i in 1:(length(top.markers$gene)/4)){
  start = ((i-1)*4)+1; print(start)
  end = i*4; print(end)
  print(FeaturePlot(sbs, features = top.markers$gene[start:end]))
}

FeaturePlot(sbs, features = c("percent.mt", "nCount_RNA","nFeature_RNA"))
FeaturePlot(sbs, features = c("PC_1","PC_2","PC_3","PC_4"))
```


```{r}
norm.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(subset(sbs,downsample=100), features= top10$gene)
```


## Cluster GO

### set up

```{r load_go}
molsig <- clusterProfiler::read.gmt(here("data/gene_sets/msigdb.v2024.1.Hs.symbols.gmt"))
head(molsig); nrow(molsig)
##Pick gene sets to look in bcos all is a lot
prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))
```

```{r}
shorten = function(ont) {
  abbreviate(gsub("_"," ", tolower(ont)),minlength=40, dot=T, named = F)
}
```

### Run 

```{r}
norm.markers$cluster = factor(norm.markers$cluster, levels =0:max(as.integer(norm.markers$cluster)))
all.genes = rownames(sbs)
```

```{r run_go}
clgo = compareCluster(gene~cluster, data=norm.markers, 
                      fun="enricher", TERM2GENE = some.molsig, universe = all.genes)
clgo = mutate(clgo, cluster = factor(cluster, levels=0:max(as.integer(as.character(cluster)))),
              abbrev_ID = shorten(ID))
head(clgo)[,c(2,4:10)]
```

### GO Plots 
```{r go_plot}
p = dotplot(clgo, x="cluster", showCategory=2, includeAll=F) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))

p = dotplot(filter(clgo,!grepl("^(WP|KEGG|REACTOME)_",ID)), x="cluster", showCategory=2, includeAll=F) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))

p = dotplot(clgo, x="cluster", showCategory=2) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))
```


## save

```{r save}
if (subset_data == F){
  saveRDS(sbs, file = here(odir, "complete_analysis.rds"))
  write.table(norm.markers, here(odir,"marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(clgo, here(odir, "ORA_marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(composition, here(odir, "cluster_composition.txt"),
              sep="\t", row.names = F)
  Idents(sbs) = "orig.ident"
  sbs = subset(sbs, downsample=1000)
  saveRDS(sbs, file = here(odir, "10%_complete_analysis.rds"))
} else if (subset_data){
  saveRDS(sbs, file = here(odir, "10%_subset_analysis.rds"))
}
```


