---
title: "Figure 5: Beige_vs_white"
author: "Sarah Hazell Pickering"
date: "2026-01-14"
output: html_document
---

# Figure 5
## Beige vs white

## Setup


```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(ggrepel)
library(here)
library(tidyr)
```

```{r global_params}
module = "beige_vs_white"

load_data = T #loading data
subset_data = F

rmd_filename = "figure5_beige_vs_white"
i_am(file.path("figures", paste0(rmd_filename,".Rmd")))
```

```{r setup}
set.seed(8)
fig_dir = here("figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank(),
        #Add facet boxes
        panel.border=element_rect(colour="lightgrey", fill=NA),
        #Add black border to colourbar legends
        legend.ticks = element_line(colour="black", linewidth=0.25),
        legend.frame = element_rect(colour="black", linewidth = 0.25),
        legend.title = element_text(size=13,hjust=0.5))
theme_set(theme_hp)
pal = c("#1F78B4","#85C2EA","#B3B3B3", "#F18486", "#E31A1C")
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = 'Spectral'))
```

## Gene-level 
### panels B & C

```{r gene_params}
dataname = "downsample"
odir = here("output", module, dataname)
clusters =   MetBrewer::met.brewer("Redon", 12)
paired_clusters = c(rbind(colorspace::darken(clusters, amount=0.3),
                    colorspace::lighten(clusters, amount=0.3)))
cluster_levels = c("b9","b6","b0","b7","b10","b5","b1","b2","b4")
```

### Read tables
```{r}
cluster_test_gene = read.delim(here(odir,"cluster_composition_test.tsv"))
cluster_test_gene$cluster = factor(cluster_test_gene$cluster, levels=cluster_levels)
head(cluster_test_gene)
```


```{r}
bvw_deg = read.delim(here(odir, "DE_per_cluster_white_vs_beige.tsv"))
bvw_deg$cluster = factor(bvw_deg$cluster, levels=paste0("b",0:max(as.integer(gsub("b","", bvw_deg$cluster)))))
head(bvw_deg)
```

## B: Cluster comparison (column plots)

```{r}
to_plot = dplyr::rename(cluster_test_gene, prop_beige=prop, conf.int.1_beige=conf.int.1, conf.int.2_beige=conf.int.2)
to_plot$prop_white = 1-to_plot$prop_beige
to_plot$conf.int.1_white = to_plot$conf.int.1_beige + (to_plot$prop_white - to_plot$prop_beige)
to_plot$conf.int.2_white = to_plot$conf.int.2_beige + (to_plot$prop_white - to_plot$prop_beige)
to_plot = pivot_longer(to_plot, matches("white|beige"), 
                       names_to = c(".value","condition"),
                       names_pattern = "(.*)_(.*)",
                     )
to_plot = mutate(to_plot, 
                 condition = factor(condition, levels=c("white","beige")),
                 clu_pval_star = if_else(p.adjust < 0.01, paste(cluster, "**"),
                                 if_else(p.adjust < 0.05, paste(cluster, "*"),
                                         cluster)))
to_plot$cond.time = paste(to_plot$condition, sep=".", to_plot$time)
to_plot$cond.time = factor(to_plot$cond.time, levels=c("white.day1","white.day3","beige.day1","beige.day3"))
to_plot$clu.cond.time = paste(to_plot$cluster, sep=".", to_plot$cond.time)
to_plot 
```



To do: Stars on b5 and b2
```{r B_wvb_proportion_test}
ggplot(to_plot, aes(x=condition, fill=clu.cond.time, y=prop))  +
  geom_hline(yintercept = 0.5, linewidth=0.4, colour="grey55") + 
  geom_bar(position="dodge", stat='identity', show.legend = F) + 
  geom_errorbar( aes(ymin=conf.int.1, ymax=conf.int.2), 
                 position=position_dodge(0.9), 
                 width=0.4, linewidth=0.3, colour="black") +
  facet_wrap(time~clu_pval_star,nrow=2, scales = "free_x")  +
  scale_y_continuous(expand=c(0,0)) + scale_fill_manual(values=paired_clusters) +
    labs(x=NULL, y="Proportion of cells") 
```


## C: DEG per cluster

```{r}
nonmt = filter(bvw_deg, !grepl("MT-", gene_name))
top_nonmt = group_by(nonmt, cluster) %>% arrange(avg_log2FC) 
beige = filter(top_nonmt, avg_log2FC > 0) %>% slice_max(n=3, order_by = avg_log2FC)
white = filter(top_nonmt, avg_log2FC < 0) %>% slice_min( n=3, order_by = avg_log2FC)
top_total = union(beige, white) # removes duplicates
head(top_total)
```

```{r}
#needed to sort gene names by cluster
top_total = arrange(top_total, time, cluster, avg_log2FC)
top_total$gene_name = factor(top_total$gene_name, levels=unique(top_total$gene_name))
```

FYI cluster b10 had no significant genes after multiple testing adjustment (1 before)
```{r C_nomt_de_per_cluster}
ggplot(top_total, aes(y=gene_name, colour=avg_log2FC, 
                    x=cluster)) + 
  geom_point(size=5) + 
  scale_colour_gradient2(low="#1F78B4",mid="beige",high="#FF7F00") +
    facet_wrap(~time, scales="free") +
    labs(y="Gene",x="Beige vs white cluster", colour="Mean\nlog_2FC") +
    theme(legend.title = element_text(size=13,hjust=0.5, vjust=4))
```


## TSS-level
### panels D-H

```{r tss_params}
dataname = "tss_bvw_initial"
odir = here("output", module, dataname)
clusters =  tvthemes::gravityFalls_pal()(12)
cluster_levels = c("t7","t10","t1","t11","t3","t8","t9","t5","t0","t2","t4","t6")
names(clusters) = paste0("t", 0:(length(cluster_levels)-1))
clusters
paired_clusters = c(rbind(colorspace::lighten(clusters, amount=0.3),
                    colorspace::darken(clusters, amount=0.3)))
clu.cond_levels = paste(rep(names(clusters), each=2), sep=".", rep(c("white","beige"), times=12))
names(paired_clusters) = clu.cond_levels
paired_clusters
```

### Read data

```{r}
if (subset_data){
  sbs = readRDS(here(odir, "10%_complete_analysis.rds"))
} else {
  sbs = readRDS(here(odir, "complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
Idents(sbs) = "seurat_clusters"
sbs
all.genes = row.names(sbs)
table(Idents(sbs))
```

### Read tables

#### Cluster composition tables
```{r}
tss_cluster_test = read.delim(here(odir,"cluster_composition_test.tsv"))
tss_cluster_test$cluster = factor(tss_cluster_test$cluster, levels=cluster_levels)
head(tss_cluster_test)

tcomp = read.delim(here(odir,"cluster_composition_condition_time.tsv"))
time_per_cluster = group_by(tcomp, seurat_clusters, time) %>% summarise(n_cells = sum(n_cells)) %>% 
  slice_max(n=1, order_by = n_cells) %>% mutate(cluster = paste0("t", seurat_clusters)) %>%
    select(seurat_clusters, cluster, time)
time_per_cluster
```
#### Marker genes

```{r}
tss.markers = read.delim(here(odir, "marker_genes.txt"))
tss.markers$cluster = factor(paste0("t", tss.markers$cluster), levels = cluster_levels)
tss.markers = merge(tss.markers, time_per_cluster, by="cluster")
head(tss.markers)

marker_go = read.delim(here(odir, "ORA_marker_genes.txt"))
marker_go$cluster = factor(paste0("t", marker_go$cluster), levels = cluster_levels)
marker_go =merge(marker_go, time_per_cluster[c("cluster","time")],by= "cluster")
head(marker_go)
```


#### DE per cluster
```{r}
deg_clusters = read.delim(here(odir,"DE_per_cluster_wvb.tsv"))
deg_clusters$cluster = factor(deg_clusters$cluster, levels=cluster_levels)
head(deg_clusters)

go_degclus = read.delim(here(odir,"DE_per_cluster_wvb_GO_terms.tsv"))
go_degclus = go_degclus %>% mutate(cluster = factor(cluster, levels=cluster_levels),
                                   condition = factor(condition, levels=c("white","beige")))
head(go_degclus)
```


## D: UMAPs

```{r D_umap_time}
sbs$cond.time = factor(sbs$cond.time, levels=c("white.day3","white.day1","asc.day0","beige.day1","beige.day3"))
DimPlot(sbs, group.by="cond.time", cols=pal, shuffle=T, label=F) + 
    geom_text(data=data.frame(), 
              aes(label="day0", x=-10, y=4), colour=pal[3],
              size=5) +
    geom_text(data=data.frame(), 
              aes(label="day1", x=-2, y=-8), colour=pal[4],
              size=5) +
    geom_text(data=data.frame(), 
              aes(label="day3", x=-2, y=2), colour=pal[5],
              size=5) +
    labs(title=NULL)
```


```{r D_umap_cluster}
DimPlot(sbs, cols= clusters, shuffle=T, label=F) + 
    scale_color_manual(values=unname(clusters), labels=function(x)paste0("t",x), name="TSS cluster")

DimPlot(sbs, cols= clusters, shuffle=T, label=T) + 
    scale_color_manual(values=unname(clusters), labels=function(x)paste0("t",x), name="TSS cluster")
```

## E: Cluster comparison (column plots)

```{r}
to_plot = dplyr::rename(tss_cluster_test, prop_beige=prop, conf.int.1_beige=conf.int.1, conf.int.2_beige=conf.int.2)
to_plot$prop_white = 1-to_plot$prop_beige
to_plot$conf.int.1_white = to_plot$conf.int.1_beige + (to_plot$prop_white - to_plot$prop_beige)
to_plot$conf.int.2_white = to_plot$conf.int.2_beige + (to_plot$prop_white - to_plot$prop_beige)
to_plot = pivot_longer(to_plot, matches("white|beige"), 
                       names_to = c(".value","condition"),
                       names_pattern = "(.*)_(.*)",
                     )
to_plot = mutate(to_plot, 
                 condition = factor(condition, levels=c("white","beige")),
                 clu_pval_star = if_else(p.adjust < 0.01, paste(cluster, "**"),
                                 if_else(p.adjust < 0.05, paste(cluster, "*"),
                                         cluster)))
to_plot$cond.time = paste(to_plot$condition, sep=".", to_plot$time)
to_plot$cond.time = factor(to_plot$cond.time,
                           levels=c("white.day1","white.day3","beige.day1","beige.day3"))
to_plot$clu.cond = factor(paste(to_plot$cluster, sep=".", to_plot$condition),
                          levels = names(paired_clusters))
to_plot 
```



To do: Stars
```{r E_wvb_proportion_test}
ggplot(to_plot, aes(x=condition, fill=clu.cond, y=prop))  +
    geom_hline(yintercept = 0.5, linewidth=0.4, colour="grey55") + 
    geom_bar(position="dodge", stat='identity', show.legend = F) + 
    geom_errorbar( aes(ymin=conf.int.1, ymax=conf.int.2), 
                 position=position_dodge(0.9), width=0.4, linewidth=0.3, colour="black") +
    facet_wrap(time~clu_pval_star,nrow=2, scales = "free_x")  +
    scale_y_continuous(expand=c(0,0)) + 
    scale_fill_manual(values=paired_clusters) +
    labs(x=NULL, y="Proportion of cells") 
```

## F: Marker genes

```{r}
top.markers = tss.markers %>%
            filter(!grepl("ADDG", gene) & time != "day0")  %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)
top.markers

```

```{r}
top.markers$gene = factor(top.markers$gene, levels=rev(unique(top.markers$gene)))
overlap.to.plot = filter(top.markers, gene %in% top.markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(top.markers$gene))

ggplot(overlap.to.plot, aes(y=gene, size=avg_log2FC, color=cluster,
                    x=cluster)) + 
    geom_point(colour="black",aes(size = avg_log2FC +(avg_log2FC*0.25))) +
    geom_point() + 
    theme(axis.text.x=element_text(angle=30, hjust=1))  + 
    scale_color_manual(values=clusters, guide="none") +
    scale_size_continuous(breaks=c(2,4,6))+
    facet_grid(~time, scales="free_x", space="free_x")+
    labs(y="Top TSSs", x="TSS cluster", size="Mean\nlog_2FC")

```

```{r  F_top_markers_include_all}
include_all_plot = filter(tss.markers, gene %in% overlap.to.plot$gene & 
                              p_val_adj< 0.01 & time !="day0")
include_all_plot$gene = factor(include_all_plot$gene, levels=levels(top.markers$gene))

ggplot(include_all_plot, aes(x=cluster, y=gene, colour=cluster, size=avg_log2FC)) + 
   geom_point(colour="black",aes(size = avg_log2FC +(avg_log2FC*0.25))) + 
    geom_point() +
    scale_color_manual(values=clusters, guide = "none") +
  facet_grid(~time, scales="free_x", space="free_x")+
    labs(y="Top TSSs", x="TSS cluster", size="Mean\nlog_2FC")
```

## G: GO terms 

```{r go_replace}
phrases= read.delim(here("data/gene_sets/abbreviations.txt"))
recap= read.delim(here("data/gene_sets/capitalisations.txt"))
```

```{r abrev_go}
drama_shorten = function(ont) {
  for (x in 1:nrow(phrases)){
    ont = gsub( phrases$phrase[x],phrases$abbreviation[x], ont, ignore.case = T)
  }
  a = abbreviate(gsub("_"," ", gsub("^[[:alpha:]]*_","",tolower(ont))),minlength=41, dot=T, named = F)
  a = stringr::str_to_sentence(a)
  for (i in 1:nrow(recap))
    a = gsub(sprintf("\\b%s\\b", recap$lowercase[i]),recap$capital[i],a, ignore.case = T)
  return(a)
}
```


```{r go_wo_wpmf}
go_to_plot = filter(marker_go, p.adjust < 0.01 & 
                                !grepl("^MF", ID) & 
                                time !="day0") %>% 
    mutate(short_desc = drama_shorten(Description)) %>%
    group_by(cluster) %>%
    slice_min(n = 3, order_by = data.frame(p.adjust,pvalue))
go_to_plot$short_desc = factor(go_to_plot$short_desc, levels=rev(unique(go_to_plot$short_desc)))

ggplot(go_to_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
    geom_point(colour="black",aes(size = Count +(Count*0.25)))  +
    geom_point()+   
    theme(axis.title.y = element_blank()) +
    scale_color_manual(values=clusters, guide="none") + 
    facet_grid(~time, scales="free_x", space = "free_x") 
```

## H: DE TSSs per cluster


#### Filter out MT and unannotated

```{r}
nonmt = filter(deg_clusters, !grepl("^MT-|^ADDG", gene_names)) %>% group_by(cluster) %>% 
  arrange(desc(abs(avg_log2FC)))
table(nonmt$sig_cond, nonmt$cluster)
beige = filter(nonmt, avg_log2FC > 0) %>% slice_max(n=3, order_by = avg_log2FC)
white = filter(nonmt, avg_log2FC < 0) %>% slice_min( n=3, order_by = avg_log2FC)
top_total = union(beige, white) # removes duplicates
head(top_total)
```



```{r}
#needed to sort gene names by cluster
top_total = arrange(top_total, time, cluster, avg_log2FC)
top_total$best_feature_name= factor(top_total$best_feature_name, levels=unique(rev(top_total$best_feature_name)))
```

### Plot BvW DE TSS
```{r H_toptss_bvw}
ggplot(top_total, aes(y=best_feature_name, colour=avg_log2FC, 
                    x=cluster)) + 
  geom_point(size=5) + 
  scale_colour_gradient2(low="#1F78B4",mid="beige",high="#FF7F00"
                        ) +  
    theme( legend.title = element_text(size=13,hjust=0.5, vjust=4),
        legend.key.height = rel(1.5)) +
  facet_wrap(~time, scales = "free") + labs(y="DE TSSs",x="TSS cluster", colour="Mean\nlog_2FC") 
```


## I: GO per cluster

```{r}
go_to_plot = as.data.frame(go_degclus) %>% filter(p.adjust < 0.05 & Count > 3) %>% 
              group_by(time,Cluster) %>%
              slice_min(n = 2, order_by = data.frame(p.adjust, pvalue))
go_to_plot$short_desc = factor(go_to_plot$short_desc, levels=rev(unique(go_to_plot$short_desc)))

ggplot(go_to_plot, aes(x=cluster, y=short_desc, colour=condition, size=Count,shape=p.adjust < 0.01)) + 
  geom_point()+   
  scale_color_manual(values=c("#1F78B4","#FF7F00")) + 
    scale_size(name="Genes") +
  scale_x_discrete(name="TSS clusters") +
  facet_grid(condition~time, scales="free",space="free", axis.labels = ) +
  theme( axis.title.y=element_blank(), strip.text.y = element_blank()) +
  scale_shape_manual(name="Adj. p < ", labels=c("0.05","0.01"), values=c(1,19))
```

```{r go_bvw_coloured_all}
go_to_plot$desc.cond = paste(go_to_plot$Description ,sep=".", go_to_plot$condition)
include_all_plot = mutate(go_degclus,desc.cond = paste(Description, sep=".", condition)) %>%
    filter(desc.cond %in% go_to_plot$desc.cond & 
           Count >= 3 & p.adjust < 0.05)
include_all_plot$short_desc = factor(include_all_plot$short_desc, levels=levels(go_to_plot$short_desc))

ggplot(include_all_plot, 
       aes(x=cluster, y=short_desc, colour=condition, 
           size=Count,shape=p.adjust < 0.01)) + 
  geom_point()+   
  scale_color_manual(values=c("#1F78B4","#FF7F00")) + 
    scale_size(name="Genes") +
  scale_x_discrete(name="TSS clusters") +
  facet_grid(condition~time, scales="free",space="free", axis.labels = ) +
  theme(axis.title.y=element_blank(), strip.text.y = element_blank()) +
  scale_shape_manual(name="Adj. p < ", labels=c("0.05","0.01"), values=c(1,19))
```
