---
title: "figure6_early_adipogenic_trajectories"
author: "Sarah Hazell Pickering"
date: "2026-01-19"
output: html_document
---

# Figure 6

New figure 6 plan
A & B Large bvw cluster monocle trajectory and RNA velocity with arrows
C Alignment match between beige and white pseudotimes -> show 3 genes for each step of the trajectory
    1-> add velocity_pseudotime and prog cols to bvw_fastmnn/complete_analysis.rds 
    2-> save as _seurat.rds file (snakemake will convert to h5ad)
    3-> add velocity_column param to the g2g alignment rule & script.  When we set of pseudotime column for g2g, which has to be called "time", we'll use this velocity column param, rather than using "monocle_pseudotime" (which is hard-coded atm)
D Violin plot, split by bvw cluster, with paired white and beige?
    -> use output of 1 above :) 
E Violin plot with progenitor colours

## Setup

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(ggrepel)
library(here)
library(tidyr)
```

```{r global_params}
load_data = T #loading data
subset_data = F

rmd_filename = "figure6_early_adipogenic_trajectories"
i_am(file.path("figures", paste0(rmd_filename,".Rmd")))
```

```{r setup}
set.seed(8)
fig_dir = here("figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4","#85C2EA","#B3B3B3", "#F18486", "#E31A1C")
prog_cols = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00")
prog_levels = c("p5","p4","p6","p3","p0","p1","p2")
names(prog_cols) = paste0("p", 0:(length(prog_levels)-1))
ps_cols = c(prog_cols, pal[2:1])
names(ps_cols) = c(names(prog_cols),"day1","day3")
```

## TSS level trajectories
```{r tss_setup}
module = "beige_vs_white"
dataname="tss_fastmnn"
odir = here("output", module, dataname)
clusters =   tvthemes::gravityFalls_pal()(13)
```

### read data
```{r}
beige = readRDS(here(odir,"beige_monocle_complete_analysis_umap.rds"))
beige$progen_clusters = paste0("p", beige$progen_clusters)
table(beige$progen_clusters)
```


## A: UMAP with trajectory

```{r Alarge_umap}
p=plot_cells(beige, color_cells_by = "initial_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F,  cell_size=0.5,  alpha = 0.5,
             graph_label_size = 5)+ theme_hp +
    scale_colour_manual(values=clusters, name="TSS clusters", labels=function(x) paste0("t",x))  +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=5)))
p$layers[[1]]$aes_params$colour = 'transparent' 
p
```


```{r Alarge_umap_save}
for (d in c("png","pdf")){
    ggsave(here(fig_dir,paste0("Aggsave_monocle_UMAP.",d)), dpi=600)
}
```
## B Progenitor trajectories
```{r Csplit_umap}
beige_progen = beige[,subset(colData(beige), grepl("^p[0-7]",time.progen_clusters)) %>%row.names]
w=plot_cells(beige_progen, color_cells_by = "progen_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F, cell_size=0.2)
w$layers[[1]]$aes_params$colour = 'transparent'
w + facet_wrap(~ progen_clusters) +
  scale_colour_manual(values=prog_cols,name = "cluster") +
    theme_hp + coord_cartesian(x=c(-11,11), y=c(-6, 6))

for (d in c("png","pdf")){
    ggsave(here(fig_dir,paste0("Bggsave_MonocleUMAPs.",d)), dpi=600)
}
```

## C: Violins

```{r}
meta= read.delim(here(odir, "monocle_pseudotime.tsv"))
meta$time.progen_clusters = factor(meta$time.progen_clusters )
meta$cond.time = factor(meta$cond.time, levels=c("white.day3","white.day1","asc.day0","beige.day1","beige.day3"))
```



```{r Cmonocleviolin_stacked}
names(ps_cols) = c(names(prog_cols), "day1","day3")
ggplot(filter(meta, sample_or_progen_cluster != "day0"), 
       aes(x=monocle_pseudotime, y=time.progen_clusters, fill=time.progen_clusters)) + 
    geom_violin(scale="width", show.legend = F)+
    coord_cartesian(expand=F) +
  facet_grid(pscondition~.) + scale_fill_manual(values=ps_cols, name="identity") + 
  scale_y_discrete(limits=c(paste0("day",c(3,1)), rev(prog_levels))) +
    labs(y="Progenitor cluster or time") +theme_hp
```

```{r Cmonocleviolin_side_by_side}

ggplot(filter(meta, sample_or_progen_cluster != "day0"), 
       aes(x=monocle_pseudotime, y=time.progen_clusters, fill=time.progen_clusters)) + 
    geom_violin(scale="width", show.legend = F)+
    coord_cartesian(expand=F) +
  facet_grid(.~pscondition) + scale_fill_manual(values=ps_cols, name="identity") + 
  scale_y_discrete(limits=c(paste0("day",c(3,1)), rev(prog_levels))) +
    labs(y="Progenitor cluster or time") +theme_hp
```

## Density maps

```{r}
ggplot(meta, aes(x=monocle_pseudotime, fill=condition)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(3,5,1)])
ggplot(meta, aes(x=monocle_pseudotime, fill=pscondition)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(5,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=pscondition)) + 
    geom_histogram(alpha=0.5, bins=15, position="identity") +
    scale_fill_manual(values=pal[c(3,5,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=pscondition)) + 
    geom_histogram( alpha=0.5, bins=30, position = "identity") +
    scale_fill_manual(values=pal[c(5,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(3,2,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=cond.time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal)
```


```{r}
ggplot(meta, aes(x=monocle_pseudotime, fill=cond.time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=c(pal ,"lavender","orange")) +  
    geom_vline(xintercept = seq(from=0, to=1, length.out=16))

```



# Beige vs white

```{r bvw_setup}
module = "beige_vs_white"
dataname="bvw_fastmnn"
odir = here("output", module, dataname)
clusters =  MetBrewer::met.brewer("Redon", 12)
```

### read monocle
```{r}
monoclew = readRDS(here(odir,"white_monocle_complete_analysis_umap.rds"))
monoclew$progen_clusters = paste0("p", monoclew$progen_clusters)
table(monoclew$time.progen_clusters)

monocleb = readRDS(here(odir,"beige_monocle_complete_analysis_umap.rds"))
monocleb$progen_clusters = paste0("p", monocleb$progen_clusters)
table(monocleb$time.progen_clusters)
```


## D: UMAPs

```{r}
p=plot_cells(monoclew, color_cells_by = "time.progen_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F,  cell_size=0.75,  alpha = 0.5,
             graph_label_size = 5)+ theme_hp +
    scale_colour_manual(values=ps_cols, name="Progenitor\ncluster\nor time", limits = c(names(prog_cols), "day1", "day3"))  +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=5)))
p$layers[[1]]$aes_params$colour = 'transparent' 
p
```

```{r}
p=plot_cells(monocleb, color_cells_by = "time.progen_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F,  cell_size=0.75,  alpha = 0.5,
             graph_label_size = 5)+ theme_hp +
    scale_colour_manual(values=ps_cols, name="Progenitor\ncluster\nor time", limits = c(names(prog_cols), "day1", "day3"))  +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=5)))
p$layers[[1]]$aes_params$colour = 'transparent' 
p
```

## PLay with denisty maps

```{r}
bvwmeta = read.delim(here(odir, "monocle_pseudotime.tsv"))
head(bvwmeta)

ggplot(bvwmeta, aes(x=monocle_pseudotime, fill=pscondition)) + geom_density(alpha=0.5)
ggplot(bvwmeta, aes(x=monocle_pseudotime, fill=cond.time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal)

ggplot(bvwmeta, aes(x=monocle_pseudotime, fill=time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(3,2,1)])
```



