---
title: "figure6_early_adipogenic_trajectories"
author: "Sarah Hazell Pickering"
date: "2026-01-19"
output: html_document
---

# Figure 6

    a. Violin plot of monocle3 pseudotime calculated for separate beige and white trajectories, separated by common clusters (see also Supp S7). 
    b. Alignment of beige vs white scRNAseq trajectories on key genes (method: g2g alignment on monocle3 pseuodtime)]
    c. Monocle trajectory with prog cols - white and beige trajectory overlayed
    d. RNA velocity arrows with prog cols - total analysis. 
    e. Violin prog colours beige vs white monocle/ RNA velocity / harmony (RNA velocity)
    f. TSS prog colours
    g. Supp - Harmony map with prog colours. 



## Setup

```{r libs, message=FALSE, warning=FALSE}
library(BPCells)
library(Seurat)
library(clusterProfiler)
library(dplyr)
library(enrichplot)
library(ggplot2)
library(ggrepel)
library(here)
library(tidyr)
```

```{r global_params}
load_data = T #loading data
subset_data = F

rmd_filename = "figure6_early_adipogenic_trajectories"
i_am(file.path("figures", paste0(rmd_filename,".Rmd")))
```

```{r setup}
set.seed(8)
fig_dir = here("figures", rmd_filename)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
knitr::opts_chunk$set(fig.path = paste0(fig_dir, "/"),
                      dev = c("png","pdf"))
```


```{r themes}
theme_hp = theme_classic(base_size = 15) +
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth = 0.25),
        strip.background = element_blank())
theme_set(theme_hp)
pal = c("#1F78B4","#85C2EA","#B3B3B3", "#F18486", "#E31A1C")
names(pal) =  c("white.day3","white.day1","day0", "beige.day1","beige.day3")
prog_cols = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00")
prog_levels = c("p5","p4","p6","p3","p0","p1","p2")
names(prog_cols) = paste0("p", 0:(length(prog_levels)-1))
ps_cols = c(prog_cols, pal[2:1])
names(ps_cols) = c(names(prog_cols),"day1","day3")
```

# TSS level trajectories
```{r tss_setup}
module = "beige_vs_white"
dataname="tss_fastmnn"
odir = here("output", module, dataname)
clusters =   tvthemes::gravityFalls_pal()(13)
cluster_levels = as.character(c(7,11,1,10,3,8,9,0,5,6,2,4))
```
## Load 
### Master pseudotime
```{r}
meta= read.delim(here(odir, "monocle_pseudotime.tsv"))
meta$time.progen_clusters = factor(meta$time.progen_clusters )
meta$initial_clusters = factor(meta$initial_clusters, levels = c(0:11))
meta$cond.time = factor(meta$cond.time, levels=c("white.day3","white.day1","asc.day0","beige.day1","beige.day3"))
table(meta$initial_clusters)
table(meta$cond.time)
```
### Subset summary
Ran monocle trajectory analysis 12x on random subsamples of the data (3300 cells or 50% of cells per condition, whichever was smaller)

```{r}
dns_meta = read.delim(here(odir, "subset_summary_monocle_pseudotime.tsv"))
mean_per_cluster = group_by(dns_meta, subset_no, initial_clusters, pscondition) %>% 
    summarise(mean_pseudotime= mean(monocle_pseudotime)) %>% 
    mutate(initial_clusters = factor(initial_clusters, levels=cluster_levels))
head(mean_per_cluster)
table(mean_per_cluster$subset_no)
```


##A: White vs beige violin
```{r}
mean_per_cluster$cluster.pscond = paste(mean_per_cluster$initial_clusters, sep=".",mean_per_cluster$pscondition)
ggplot(mean_per_cluster, aes(y=initial_clusters, fill=pscondition, x=mean_pseudotime))+
    geom_violin(data=meta, aes(x=monocle_pseudotime),scale="width") + 
    geom_boxplot(fill="transparent", position=position_dodge(0.9),
                 aes(group=cluster.pscond, colour="n=12 subsets")) +
     scale_fill_manual(values=c("#FF7F00","#1F78B4"), name="Pseudotime\ncalculated on",
                       guide=guide_legend(reverse=T)) +
    scale_colour_manual(values="black", name="Average\npseudotime") +
    coord_cartesian(expand=F) + scale_y_discrete(limits = rev(cluster_levels),
                                                 label=function(x)paste0("b",x)) +
    labs(x="Monocle pseudotime", title=NULL, y="TSS clusters") 
ggsave(here(fig_dir, "wvb_violin_with_variance.png"), width = 9, height = 8)
```


# Beige vs white

```{r bvw_setup}
module = "beige_vs_white"
dataname="bvw_fastmnn"
odir = here("output", module, dataname)
clusters =  MetBrewer::met.brewer("Redon", 12)
cluster_levels = as.character(c(8,3,11,9,6,0,7,10,1,2,5,4))
```

## Read seurat objects 
```{r}
sbs_white= readRDS(here(odir, "white_combined_pseudotime_seurat.rds"))
sbs_beige=   readRDS(here(odir, "beige_combined_pseudotime_seurat.rds"))
merged_obj <- merge(x = sbs_white, y = sbs_beige, add.cell.ids = c("white","beige"), merge.dr = F)
merged_obj
merged_obj$progen_clusters = factor(merged_obj$progen_clusters, levels=c(paste0("p",0:7),paste0("day",c(0,1,3))))
head(merged_obj@meta.data)

meta = read.delim( here(odir,"combined_pseudotime.tsv"))
meta$initial_clusters = factor(meta$initial_clusters, levels=cluster_levels)
meta$sample_or_progen_cluster = factor(meta$sample_or_progen_cluster, 
                                       levels=c(prog_levels, c("day0","white.day1", "beige.day1","white.day3","beige.day3")))
head(meta)
```

### read monocle object
```{r}
monoclew = readRDS(here(odir,"white_monocle_complete_analysis_umap.rds"))
monoclew$progen_clusters = paste0("p", monoclew$progen_clusters)
table(monoclew$time.progen_clusters)

monocleb = readRDS(here(odir,"beige_monocle_complete_analysis_umap.rds"))
monocleb$progen_clusters = paste0("p", monocleb$progen_clusters)
table(monocleb$time.progen_clusters)
```

## Bvw Violin Plots

```{r paired_ridge_plot}
VlnPlot(merged_obj, features = "monocle_pseudotime", group.by = "initial_clusters", split.by="pscondition",
        cols= c("#FF7F00",  "#1F78B4"), pt.size = 0, split.plot = T) + coord_flip(expand=F) +
  scale_x_discrete(limits = rev(cluster_levels) )+
  labs(x="Time // Progenitor cluster", y="Monocle pseudotime", title=NULL) + guides(fill=guide_legend(reverse=T)) +
    theme_hp
```


```{r}
ggplot(meta, 
       aes(x=monocle_pseudotime, y=initial_clusters, fill=initial_clusters)) + 
    geom_boxplot(position="dodge")+
  facet_grid(pscondition~.) + scale_fill_manual(values=clusters, name="Bvw clusters", labels = function(x)paste0("b",x)) +  
    coord_cartesian(expand=F) + scale_y_discrete(labels = function(x)paste0("b",x),limits=rev(cluster_levels)) +
    labs(y="Bvw clusters", x="Monocle pseudotime", title=NULL)
```


```{r}
ggplot(meta, 
       aes(x=monocle_pseudotime, y=initial_clusters, fill=initial_clusters)) + 
    geom_violin(scale="width", position="dodge")+
  facet_grid(pscondition~.) + scale_fill_manual(values=clusters, name="Bvw clusters", labels = function(x)paste0("b",x)) +  
    coord_cartesian(expand=F) + scale_y_discrete(labels = function(x)paste0("b",x),limits=rev(cluster_levels)) +
    labs(y="Bvw clusters", x="Monocle pseudotime", title=NULL)

ggplot(meta, 
       aes(x=monocle_pseudotime, y=initial_clusters, fill=pscondition)) + 
    geom_violin(scale="width", position="dodge")+
    scale_fill_manual(values=c("#FF7F00","#1F78B4"), name="Pseudotime\ncalculated on", guide=guide_legend(reverse=T)) +  
    coord_cartesian(expand=F) + 
    scale_y_discrete(labels = function(x)paste0("b",x),limits=rev(cluster_levels)) +
    labs(y="Bvw clusters", x="Monocle pseudotime", title=NULL)
```

### Progenitor colours
```{r paired_ridge_plot}
VlnPlot(merged_obj, features = "monocle_pseudotime", group.by = "time.progen_clusters", split.by="pscondition",
        cols= c("#FF7F00",  "#1F78B4"), pt.size = 0, split.plot = T) + coord_flip(expand=F) +
  scale_x_discrete(limits = rev(c(prog_levels, paste0("day",c(1,3))))) +
  labs(x="Time // Progenitor cluster", y="Monocle pseudotime", title=NULL) + guides(fill=guide_legend(reverse=T)) +
    theme_hp
```


```{r}
ps_cols = c(prog_cols, pal[1:2])
names(ps_cols) = c(names(prog_cols), "day1","day3")
ggplot(filter(meta, sample_or_progen_cluster != "day0"), 
       aes(x=monocle_pseudotime, y=time.progen_clusters, fill=time.progen_clusters)) + 
    geom_violin(scale="width")+
  facet_grid(pscondition~.) + scale_fill_manual(values=ps_cols, name="identity", limits = names(ps_cols)) + 
  scale_y_discrete(limits=c(paste0("day",c(3,1)), rev(prog_levels))) + coord_cartesian(expand=F) +
    labs(y="Time // Progenitor cluster", x="Monocle pseudotime", title=NULL)
```

## B: g2g Alignment

![day0]()

## D: UMAPs

```{r}
p=plot_cells(monoclew, color_cells_by = "time.progen_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F,  cell_size=0.75,  alpha = 0.5,
             graph_label_size = 5)+ theme_hp +
    scale_colour_manual(values=ps_cols, name="Progenitor\ncluster\nor time", limits = c(names(prog_cols), "day1", "day3"))  +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=5)))
p$layers[[1]]$aes_params$colour = 'transparent' 
p
```

```{r}
p=plot_cells(monocleb, color_cells_by = "time.progen_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F,  cell_size=0.75,  alpha = 0.5,
             graph_label_size = 5)+ theme_hp +
    scale_colour_manual(values=ps_cols, name="Progenitor\ncluster\nor time", limits = c(names(prog_cols), "day1", "day3"))  +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=5)))
p$layers[[1]]$aes_params$colour = 'transparent' 
p
```

## Play with denisty maps

```{r}
bvwmeta = read.delim(here(odir, "monocle_pseudotime.tsv"))
head(bvwmeta)

ggplot(bvwmeta, aes(x=monocle_pseudotime, fill=pscondition)) + geom_density(alpha=0.5)
ggplot(bvwmeta, aes(x=monocle_pseudotime, fill=cond.time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal)

ggplot(bvwmeta, aes(x=monocle_pseudotime, fill=time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(3,2,1)])
```


## TSS level trajectories
```{r tss_setup}
module = "beige_vs_white"
dataname="tss_fastmnn"
odir = here("output", module, dataname)
clusters =   tvthemes::gravityFalls_pal()(13)
cluster_levels = as.character(c(7,11,1,10,3,8,9,0,5,6,2,4))
```
## A: Violins

```{r}
meta= read.delim(here(odir, "monocle_pseudotime.tsv"))
meta$time.progen_clusters = factor(meta$time.progen_clusters )
meta$initial_clusters = factor(meta$initial_clusters, levels = c(0:11))
meta$cond.time = factor(meta$cond.time, levels=c("white.day3","white.day1","asc.day0","beige.day1","beige.day3"))
```

```{r A_tss_beige_vs_white_violin}
ggplot(meta, 
       aes(x=monocle_pseudotime, y=initial_clusters, fill=pscondition)) + 
    geom_violin(scale="width", position="dodge")+
    scale_fill_manual(values=c("#FF7F00","#1F78B4"), name="Pseudotime\ncalculated on", guide=guide_legend(reverse=T)) +  
    coord_cartesian(expand=F) + 
    scale_y_discrete(labels = function(x)paste0("t",x),limits=rev(cluster_levels)) +
    labs(y="TSS clusters", x="Monocle pseudotime", title=NULL)

ggplot(meta, 
       aes(x=monocle_pseudotime, y=initial_clusters, colour=pscondition, fill=initial_clusters)) + 
    geom_violin(scale="width", position="dodge")+
    scale_fill_manual(values=clusters, guide="none") +
    scale_colour_manual(values=c("black","grey"), name="Pseudotime\ncalculated on", guide=guide_legend(reverse=T)) +  
    coord_cartesian(expand=F) + 
    scale_y_discrete(labels = function(x)paste0("t",x),limits=rev(cluster_levels)) +
    labs(y="TSS clusters", x="Monocle pseudotime", title=NULL)
```


```{r Cmonocleviolin_stacked}
names(ps_cols) = c(names(prog_cols), "day1","day3")
ggplot(filter(meta, sample_or_progen_cluster != "day0"), 
       aes(x=monocle_pseudotime, y=time.progen_clusters, fill=time.progen_clusters)) + 
    geom_violin(scale="width", show.legend = F)+
    coord_cartesian(expand=F) +
  facet_grid(pscondition~.) + scale_fill_manual(values=ps_cols, name="identity") + 
  scale_y_discrete(limits=c(paste0("day",c(3,1)), rev(prog_levels))) +
    labs(y="Progenitor cluster or time") +theme_hp
```

```{r Cmonocleviolin_side_by_side}

ggplot(filter(meta, sample_or_progen_cluster != "day0"), 
       aes(x=monocle_pseudotime, y=time.progen_clusters, fill=time.progen_clusters)) + 
    geom_violin(scale="width", show.legend = F)+
    coord_cartesian(expand=F) +
  facet_grid(.~pscondition) + scale_fill_manual(values=ps_cols, name="identity") + 
  scale_y_discrete(limits=c(paste0("day",c(3,1)), rev(prog_levels))) +
    labs(y="Progenitor cluster or time") +theme_hp
```

### read data
```{r}
beige = readRDS(here(odir,"beige_monocle_complete_analysis_umap.rds"))
beige$progen_clusters = paste0("p", beige$progen_clusters)
table(beige$progen_clusters)
```


## A: UMAP with trajectory

```{r Alarge_umap}
p=plot_cells(beige, color_cells_by = "initial_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F,  cell_size=0.5,  alpha = 0.5,
             graph_label_size = 5)+ theme_hp +
    scale_colour_manual(values=clusters, name="TSS clusters", labels=function(x) paste0("t",x))  +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=5)))
p$layers[[1]]$aes_params$colour = 'transparent' 
p
```


```{r Alarge_umap_save}
for (d in c("png","pdf")){
    ggsave(here(fig_dir,paste0("Aggsave_monocle_UMAP.",d)), dpi=600)
}
```
## B Progenitor trajectories
```{r Csplit_umap}
beige_progen = beige[,subset(colData(beige), grepl("^p[0-7]",time.progen_clusters)) %>%row.names]
w=plot_cells(beige_progen, color_cells_by = "progen_clusters", label_leaves = F, label_branch_points = F, 
                trajectory_graph_color="black", label_cell_groups = F, cell_size=0.2)
w$layers[[1]]$aes_params$colour = 'transparent'
w + facet_wrap(~ progen_clusters) +
  scale_colour_manual(values=prog_cols,name = "cluster") +
    theme_hp + coord_cartesian(x=c(-11,11), y=c(-6, 6))

for (d in c("png","pdf")){
    ggsave(here(fig_dir,paste0("Bggsave_MonocleUMAPs.",d)), dpi=600)
}
```


## Density maps

```{r}
ggplot(meta, aes(x=monocle_pseudotime, fill=condition)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(3,5,1)])
ggplot(meta, aes(x=monocle_pseudotime, fill=pscondition)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(5,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=pscondition)) + 
    geom_histogram(alpha=0.5, bins=15, position="identity") +
    scale_fill_manual(values=pal[c(3,5,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=pscondition)) + 
    geom_histogram( alpha=0.5, bins=30, position = "identity") +
    scale_fill_manual(values=pal[c(5,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal[c(3,2,1)])

ggplot(meta, aes(x=monocle_pseudotime, fill=cond.time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=pal)
```


```{r}
ggplot(meta, aes(x=monocle_pseudotime, fill=cond.time)) + geom_density(alpha=0.5) +
    scale_fill_manual(values=c(pal ,"lavender","orange")) +  
    geom_vline(xintercept = seq(from=0, to=1, length.out=16))

```




