---
title: "Cluster_composition_tests"
author: "Sarah Hazell Pickering"
date: "2025-02-03"
output: html_document
---

- Step 1: Cluster composition plot
    - save cluster counts and proportions by condition & time (no replicate information) "cluster_composition_condition_time.tsv"
- Step 2: Test if the proportion of white and beige cells in each cluster is different; aka is it different from a 0.5 white/beige proportion.  To simplify this I'm only going to check cells from the majority time point for each cluster.
    - save test result "cluster_composition_test.tsv"

```{r params}
module="adipogenesis"
dataname = "beige_vs_white_downsample"
library(here); i_am(file.path("analysis", paste0(dataname,".Rmd")))

odir = here("output", module, dataname) 
load_data = T #loading data 
subset_data = F
```

```{r libs, warning=F, message=F, echo=F}
library(dplyr)
library(tidyr)
library(ggplot2)
library(Seurat)
library(BPCells)
library(patchwork)
```


```{r themes}
pal = c("#1F78B4", 
        #"#B2DF8A",
        "#85C2EA","#B3B3B3", "#F18486",
        #"#FDBF6F",
        "#E31A1C")
clusters = #ggsci::pal_simpsons()(13)[c(1:7,9:13)]
  tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))

#define new theme
theme_hp = theme_classic(base_size=15) + 
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25),
        strip.background = element_blank(), # remove boxes around facet labels
        )
#set for all ggplots
theme_set(theme_hp)

knitr::opts_chunk$set(dev=c("png","pdf"))
```

## Read data

```{r}
if (subset_data){
  sbs = readRDS(here(odir, "10%_complete_analysis.rds"))
} else {
  sbs = readRDS(here(odir, "complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
Idents(sbs) = "seurat_clusters"
Idents(sbs) = factor(Idents(sbs), levels=0:(length(unique(Idents(sbs))))-1)
sbs
table(Idents(sbs))
```

```{r cluster.umap}
DimPlot(sbs, cols= clusters, label = T) + theme_hp
```


```{r condition.time.umap}
DimPlot(sbs, group.by="cond.time",cols=pal, shuffle=T) + theme_hp
```

## Composition plots (per condition + time)
(not per sample)
```{r}
tcomp = group_by(sbs@meta.data, cond.time,seurat_clusters, time, condition) %>% dplyr::count(name="n_cells")
tcomp= tcomp[order(tcomp$n_cells, decreasing = T),]
#calculate proportions directly, then plot percentage just for the three cyc;ling cells
tcomp= group_by(tcomp, cond.time) %>% mutate(cells_per_sample = n_cells/sum(n_cells))
tcomp= group_by(tcomp, seurat_clusters) %>% mutate(prop.cells_per_cluster = n_cells/sum(n_cells), n.cells_per_cluster = sum(n_cells))
head(tcomp)
str(tcomp)
```


```{r condition.time_composition.all}
ggplot(tcomp, aes(x=seurat_clusters, fill=cond.time, y=prop.cells_per_cluster)) + 
  geom_bar(position="dodge", stat='identity') + 
  scale_fill_manual(values=pal) + facet_grid(~time, scales="free_x") +
  geom_hline(yintercept = 0.5, linetype="dashed", linewidth=0.25) +
  scale_y_continuous(expand=c(0,0))

ggplot(tcomp, aes(x=seurat_clusters, fill=cond.time, y=n_cells)) + 
  geom_bar(position="dodge", stat='identity') + 
  scale_fill_manual(values=pal) + facet_grid(~time, scales="free_x")+
  scale_y_continuous(expand=c(0,0))
```

```{r}
write.table(tcomp, here(odir, "cluster_composition_condition_time.tsv"),
            sep="\t", quote = F, row.names = F)
```


## Best matches
For plotting and testing we want to focus of the main timepoint for each cluster; this selects the majority timepoint per cluster and then subsets our frequency table.  
```{r}
time_per_cluster = group_by(tcomp, seurat_clusters, time) %>% summarise(n_cells = sum(n_cells)) %>% 
  slice_max(n=1, order_by = n_cells) %>% select(seurat_clusters, time)
time_per_cluster

bestmatches = merge(tcomp, time_per_cluster, by=c("seurat_clusters","time"), all=F)
bestmatches$condition = gsub("\\..*","",bestmatches$cond.time)
bestmatches
```


```{r condition.time_composition.best}
ggplot(bestmatches, aes(x=seurat_clusters, fill=cond.time, y=prop.cells_per_cluster)) + 
  geom_bar(position="dodge", stat='identity') + 
  scale_fill_manual(values=pal) + facet_wrap(~time, scales="free_x") +
  geom_hline(yintercept = 0.5, linetype="dashed", linewidth=0.25)+ facet_grid(~time, scales="free_x")+
  scale_y_continuous(expand=c(0,0))

ggplot(bestmatches, aes(x=seurat_clusters, fill=cond.time, y=n_cells)) + 
  geom_bar(position="dodge", stat='identity') + 
  scale_fill_manual(values=pal) + facet_wrap(~time, scales="free_x")+ facet_grid(~time, scales="free_x")+
  scale_y_continuous(expand=c(0,0.05))
```

^ the proportions plotted above also include cells from a different time point that get included in the count.  E.g. cluster 2 here has some day1 cells in it which is why day3 white and beige don't add up to 100%.  So to be totally covering our bases we should include them in the proportion test - then we have to do individual tests to figure out where the change is coming from.  


## White vs beige proportion test
Test if the proportion of white and beige cells in each cluster is different; aka is it different from a 0.5 white/beige proportion.  To simplify this I'm only going to check cells from the majority time point for each cluster.

```{r}
clusters_to_test = filter(bestmatches, condition == "beige") %>% pull(seurat_clusters) 
clusters_to_test

wvb_df = filter(bestmatches, seurat_clusters %in% clusters_to_test) %>% 
  pivot_wider(names_from=condition, values_from = n_cells, id_cols = seurat_clusters) %>% 
  as.data.frame()
rownames(wvb_df) = paste0("b",wvb_df$seurat_clusters)
wvb = data.matrix(wvb_df)
wvb = wvb[,2:3]
wvb

## quick test are any different than 50%?
prop.test(wvb) #Yes some are different
```


```{r}
clu_tests  = list()
tests = data.frame(cluster = paste0("b",clusters_to_test), prop = wvb[,"beige"]/(wvb[,"white"]+wvb[,"beige"]),p.value=NA, conf.int.1=NA, conf.int.2=NA)

for (clu in tests$cluster){
  res = prop.test(matrix(wvb[clu,], ncol = 2), conf.level=0.99)
  clu_tests[[clu]] = res
  tests[tests$cluster == clu,]$p.value = res$p.value
  tests[tests$cluster == clu,]$conf.int.1 = res$conf.int[[1]]
  tests[tests$cluster == clu,]$conf.int.2 = res$conf.int[[2]]
}
clu_tests[[1]]$p.value
clu_tests[[1]]
tests$p.adjust = p.adjust(tests$p.value, method="fdr")
tests
```
The most significant differences are for cluster b1 and cluster b6, but only cluster b6 has a p-value robust to adjustment.  

## Plot X-squared CIs 

```{r}
to_plot = dplyr::rename(tests, prop_beige=prop, conf.int.1_beige=conf.int.1, conf.int.2_beige=conf.int.2)
to_plot$prop_white = 1-to_plot$prop_beige
to_plot$conf.int.1_white = to_plot$conf.int.1_beige + (to_plot$prop_white - to_plot$prop_beige)
to_plot$conf.int.2_white = to_plot$conf.int.2_beige + (to_plot$prop_white - to_plot$prop_beige)
to_plot = pivot_longer(to_plot, matches("white|beige"), 
                       names_to = c(".value","condition"),
                       names_pattern = "(.*)_(.*)",
                     )
bestmatches$cluster = paste0("b",bestmatches$seurat_clusters)
to_plot = merge(to_plot, bestmatches, by=c("cluster","condition"))
```


```{r wvb_proportion_test}
ggplot(to_plot, aes(y=cluster, fill=cond.time, x=prop)) + 
  geom_bar(stat='identity')     + 
  geom_errorbar(data=filter(to_plot, condition =="beige"), aes(xmin=conf.int.1, xmax=conf.int.2), 
                 width=0.7, linewidth=0.5, colour="white") +
  geom_vline(xintercept = 0.5, linetype="dashed",linewidth=0.25) +
  scale_fill_manual(values=pal[c(1,2,4,5)]) + facet_grid(time~., scales="free_y") +
  scale_x_continuous(expand=c(0,0))
```


## Column version

```{r wvb_proportion_test_columns}
ggplot(to_plot, aes(x=cluster, fill=cond.time, y=prop))  +
  geom_hline(yintercept = 0.5, linewidth=0.4, colour="grey55") + 
  geom_bar(position="dodge", stat='identity') + 
  geom_errorbar( aes(ymin=conf.int.1, ymax=conf.int.2), 
                 position=position_dodge(0.9), width=0.4, linewidth=0.3) +
  scale_fill_manual(values=pal[c(1,2,4,5)]) + facet_wrap(~time, scales="free_x")  +
  scale_y_continuous(expand=c(0,0))
```



```{r}
write.table(tests, here(odir, "cluster_composition_test.tsv"),
            sep="\t", quote=F, row.names=F)
```





