---
title: "bulk_day15_GO"
author: "Sarah Hazell Pickering"
date: "2025-01-21"
output: html_document
---

FYI: figure in paper was from running only cc and hallmark. runnning all gives the same result; but slightly different pvalue adjustment. Can fix I guess. 

```{r}
library(here); i_am("analysis/bulk_day15_mitopathways.Rmd")
knitr::opts_knit$set(dev=c("png","pdf"))
odir = here("output/bulk")
```

```{r message=FALSE, warning=FALSE, echo=TRUE}
library(tidyr)
library(dplyr)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
theme_set(theme_classic(base_size=15) + 
  theme(axis.title.y=element_blank(),axis.text.y = element_text(size=12),
        axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25)))

library(stringr)
library(viridis)
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
```

## GSEA setup
```{r}
go <- clusterProfiler::read.gmt("/projects/imb-pkbphil/sp/annotations/molsigdb/msigdb.v2024.1.Hs.symbols.gmt")
molsig = go[gsub("_.*","", go$term) %in% c("GOCC","GOBP","GOMF","KEGG","HALLMARK","WP","REACTOME"),]

head(molsig); nrow(molsig)
molsig$term = factor(molsig$term)
table(gsub("_.*","", molsig$term))
rm(go)
```
```{r go_replace}
phrases = c("protein_containing_complex" = "complex",
    "structure_organization" = "structure",
"transforming_growth_factor_beta"="TGFb",
    "_(WITH|INCLUDES_BOTH)_BLOCH_AND_.*"=":_BLOCH_&_KANDut.", #simplify to Cholesterol metabolism... 
      "_CELLULAR_" = "_CELL_", 
      "response_to_topologically_incorrect_protein" = "topologically_incorrect_protein_response", #Synonym 
     "_DRIVEN_ACTIVE_TRANSMEMBRANE"= "",#OXIDOREDUCTION...TRANSPORTER_ACTIVITY
      "_AND_RESPIRATORY_ELECTRON_TRANSPORT" ="_&_electron_transport",#AEORBIC RESPIRATION...
      "ENDOPLASMIC_RETICULUM"="ER",
    "regulating_ANTI_OXIDANT"="&_ANTIOXIDANT",
    "_ENZYMES$"="",
    "the_citric_acid"="")
#phrases

recap <- c(
    "rna" = "RNA",
    "mrna" = "mRNA",
    "rrna" = "rRNA",
    "ncrna"="ncRNA",
    "tnfa" = "TFNa",
    "nfkb" = "NFKb",
    "mtorc1" = "mTORC1",
    "bmp" = "BMP",
    "e2f" = "E2F",
    "uv" = "UV",
    "er"="ER",
    "upr"="UPR",
    "tca"="TCA",
	"ecm"="ECM")
#recap
```


Function to intelligently abbreviate GO terms and correct the capitalisation of initialisations based on the above lists.  To do: save these lists to file to make reproducible and concise.   

```{r abrev_go}

drama_shorten = function(ont) {
  for (x in 1:length(phrases)){
    ont = gsub( names(phrases)[x],phrases[x], ont, ignore.case = T)
  }
  a = abbreviate(gsub("_"," ", gsub("^[[:alpha:]]*_","",tolower(ont))),minlength=41, dot=T, named = F)
  a = stringr::str_to_sentence(a)
  for (i in 1:length(recap))
    a = gsub(sprintf("\\b%s\\b", names(recap)[i]),recap[i],a, ignore.case = T)
  return(a)
}
```

## Load data
```{r}
sig = read.delim(file="https://raw.githubusercontent.com/sarahhp/splicing_thermogenesis/main/03limma/any_and_all_donor_DGE.tsv")
head(sig)
nrow(sig)
tail(sig)

summary(sig$all.donors.adj.P.Val < 0.01)
```

```{r}
make_gene_list = function(sig, donor){
  gl = sig[,paste0("logFC.",donor)]
  names(gl) = sig$gene_name
  gl = gl[order(gl, decreasing = T)]
  return (gl)
}


allgl = list()
for (donor in c("s1","s2","s3","s4","s5","s6")){
  print(donor)
  gl = make_gene_list(sig, donor)
  allgl[[donor]] = gl
}
str(allgl)
```

## Run GSEA 

```{r}
allea = compareCluster(allgl, fun=GSEA, TERM2GENE=molsig, pvalueCutoff=1, minGSSize=3, eps=1e-40)
allea = arrange(allea, p.adjust, abs(NES))
head(allea)
summary(allea@compareClusterResult$p.adjust)
summary(allea@compareClusterResult$NES)
nrow(allea)
```


Alpha matters here for the includeAll plots

## Filter adj. pvalue
```{r}
filt = filter(allea, p.adjust < 0.01 & gsub("_.*","", ID) %in% c("GOCC","HALLMARK")) 
```


```{r}
write.table(as.data.frame(allea), here(odir, "bulk_day15_GSEA.tsv"),
            sep="\t", quote=F, row.names = F)
```

## Include all plots
NES = Normalised enrichment score

```{r go5, dev=c("png","pdf")}
dotplot(filt, showCategory=6, includeAll=T, color="NES") + scale_colour_viridis(option="mako", direction=-1)
dotplot(filt, showCategory=2, includeAll=T, color="NES") + scale_colour_viridis(option="mako", direction=-1)
```


```{r}
top2 = filter(as_tibble(filt), NES > 0 & p.adjust < 1e-10 & ID != "GOCC_MITOCHONDRIAL_PROTEIN_CONTAINING_COMPLEX" ) %>% 
  slice_min(tibble(p.adjust,plyr::desc(NES)), n=4, by=Cluster)
w2 = filter(as_tibble(filt), NES < 0 & p.adjust < 1e-8) %>% slice_min(tibble(p.adjust,NES), n=4, by=Cluster)

top2 = rbind(top2, w2) %>% mutate(condition=if_else(NES>0,"beige","white"))
top2[,c(1:2,4:11)]
top2$ID = factor(top2$ID, levels = unique(c(w2$ID, top2$ID)))
```


```{r}
ggplot(top2,aes(x=Cluster, y=ID, color=condition, size=-log10(p.adjust))) + geom_point() +
  scale_color_manual(values = c("orange","cyan3")) 
```

```{r day15_GSEA_top2pfilt, dev=c("png","pdf")}
include_all_plot = filter(as_tibble(filt), Description %in% top2$Description & p.adjust < 0.05) %>%
  mutate(condition=if_else(NES>0,"beige","white"))
include_all_plot$ID = factor(include_all_plot$ID, levels=unique(top2$ID))

ggplot(include_all_plot,aes(x=Cluster, y=ID, color=condition, alpha=abs(NES))) + geom_point(size=10) +
  scale_color_manual(values = c("#FF7F00","#1F78B4")) + xlab("Subject") + 
  scale_y_discrete(labels= function(ID) drama_shorten(ID)) 

```

