---
title: "RPCA GO analysis"
author: "Sarah Hazell Pickering"
date: "2024-08-09"
output: html_document
---

Integrated data using RPCA, plotting and marker gene detection.  

```{r params}
module="progenitors"
dataname = "rpca"
library(here);i_am(file.path("analysis",paste(module,sep="_",dataname, "clusters_and_markers.Rmd")))
odir = here("output", module, dataname) 

load_data = T #Loading initial seurat analysis
subset_data =  F
```

```{r libs, warning=F, message=F, echo=T}
library(dplyr)
library(ggplot2)
library(ggsci)

library(Seurat)
library(BPCells)

library(clusterProfiler)
library(enrichplot)
```


```{r themes, warning=F, message=F, echo=T}
pal = c("#DA1819", "#00AED1", "#434D51","#EBB400","#691A93","#3AAD00")
clusters =  c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0")
spectral =  rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))

#define new theme
theme_hp = theme_classic(base_size=15) + 
  theme(axis.line = element_line(linewidth = 0.25),
        axis.ticks = element_line(linewidth=0.25),
        strip.background = element_blank(), # remove boxes around facet labels
        )
		#set for all ggplots
theme_set(theme_hp)
knitr::opts_chunk$set(dev=c("png","pdf"))
```

## Read data

```{r load_data}
if (subset_data){
  filt = readRDS(here(odir, "10%_complete_analysis.rds"))
} else{
  filt = readRDS(here(odir, "complete_analysis.rds"))
}
filt
colnames(filt@meta.data) 
table(filt$rpca_clusters)
```

```{r umap_basic}
DimPlot(filt, reduction = "umap",  group.by="rpca_clusters", cols = clusters) + theme_hp +
    scale_color_manual(values=clusters, name = "Progenitor \ncluster", labels =function(x) paste0("p",x))
DimPlot(filt, reduction = "umap",  group.by="rpca_clusters", cols = clusters, 
        split.by = "subject", ncol=3) + theme_hp  + 
  scale_color_manual(values=clusters, name = "Progenitor \ncluster", labels =function(x) paste0("p",x))
```

## Top dimenssions from RPCA embeddings
dominated by cell cycle

```{r rpca_dimensions}
table(filt$rpca_clusters)
DimPlot(filt, reduction="integrated.rpca", group.by = "rpca_clusters", cols=clusters)
DimPlot(filt, reduction="integrated.rpca", group.by = "rpca_clusters", cols=clusters, dims = c(3,4))
DimPlot(filt, reduction="integrated.rpca", group.by = "rpca_clusters", cols=clusters, dims = c(5,6))
DimPlot(filt, reduction="integrated.rpca", group.by = "rpca_clusters", cols=clusters, dims = c(7,8))
```
## Quality control metrics

Subject S2 has a low gene/ count population

```{r qc}
FeaturePlot(filt, reduction = "umap", 
            features=c("nFeature_RNA","nCount_RNA","percent.mt","RPL27A"))  & 
  scale_color_gradientn(colours=spectral) &theme_hp
VlnPlot(filt, group.by= "rpca_clusters", pt.size=0, cols=clusters,
            features=c("nFeature_RNA","nCount_RNA","percent.mt")) &theme_hp
```

```{r qc2}
FeatureScatter(filt,feature1="nFeature_RNA",feature2 = "percent.mt", 
               group.by="subject", cols=pal, 
               split.by = "rpca_clusters", ncol=4, pt.size=0.2,
               shuffle=T)
```

## Read marker genes
```{r}
cluster.markers = read.delim(here(odir,"marker_genes.txt"))
cluster.markers$cluster = factor(cluster.markers$cluster, levels=0:max(as.integer(cluster.markers$cluster)))
head(cluster.markers)
table(cluster.markers$cluster)
```
## Marker genes

```{r marker_genes}
top.markers =  cluster.markers %>%
    group_by(cluster) %>% filter(p_val_adj < 0.01) %>%
    slice_max(n=2, order_by = avg_log2FC*pct.1)
top.markers

top.markers$gene = factor(top.markers$gene, levels=rev(unique(top.markers$gene)))
overlap.to.plot = filter(cluster.markers, gene %in% top.markers$gene)
overlap.to.plot$gene = factor(overlap.to.plot$gene, levels=levels(top.markers$gene))


ggplot(overlap.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point() + 
  theme(axis.text.x=element_text(angle=30, hjust=1))  + 
  scale_color_manual(values=clusters, guide="none") +scale_size_continuous(name="|log2FC|")+
  scale_x_discrete(name = "Progenitor cluster", labels =function(x) paste0("p",x))
```



## GO setup
```{r}
molsig <- clusterProfiler::read.gmt("/projects/imb-pkbphil/sp/annotations/molsigdb/msigdb.v2024.1.Hs.symbols.gmt")
head(molsig); nrow(molsig)
##Pick gene sets to look in bcos all is a lot
prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))

shorten = function(ont) {
  abbreviate(gsub("_"," ", tolower(ont)),minlength=40, dot=T, named = F)
}
```

```{r go_replace}
phrases = c("protein_containing_complex" = "complex",
    "structure_organization" = "structure",
"transforming_growth_factor_beta"="TGFb",
    "_(WITH|INCLUDES_BOTH)_BLOCH_AND_.*"=":_BLOCH_&_KANDut.", #simplify to Cholesterol metabolism... 
      "_CELLULAR_" = "_CELL_", 
      "response_to_topologically_incorrect_protein" = "topologically_incorrect_protein_response", #Synonym 
     "_DRIVEN_ACTIVE_TRANSMEMBRANE"= "",#OXIDOREDUCTION...TRANSPORTER_ACTIVITY
      "_AND_RESPIRATORY_ELECTRON_TRANSPORT" ="_&_electron_transport",#AEORBIC RESPIRATION...
      "ENDOPLASMIC_RETICULUM"="ER",
    "regulating_ANTI_OXIDANT"="&_ANTIOXIDANT",
    "_ENZYMES$"="",
    "the_citric_acid"="")
#phrases

recap <- c(
    "rna" = "RNA",
    "mrna" = "mRNA",
    "rrna" = "rRNA",
    "ncrna"="ncRNA",
    "tnfa" = "TFNa",
    "nfkb" = "NFKb",
    "mtorc1" = "mTORC1",
    "bmp" = "BMP",
    "e2f" = "E2F",
    "uv" = "UV",
    "er"="ER",
    "upr"="UPR",
    "tca"="TCA")
#recap
```


Function to intelligently abbreviate GO terms and correct the capitalisation of initialisations based on the above lists.  To do: save these lists to file to make reproducible and concise.   

```{r abrev_go}
drama_shorten = function(ont) {
  for (x in 1:length(phrases)){
    ont = gsub( names(phrases)[x],phrases[x], ont, ignore.case = T)
  }
  a = abbreviate(gsub("_"," ", gsub("^[[:alpha:]]*_","",tolower(ont))),minlength=41, dot=T, named = F)
  a = stringr::str_to_sentence(a)
  for (i in 1:length(recap))
    a = gsub(sprintf("\\b%s\\b", names(recap)[i]),recap[i],a, ignore.case = T)
  return(a)
}
```



## Cluster GO

```{r}
all.genes=rownames(filt)
clgo = compareCluster(gene~cluster, data=filter(cluster.markers,p_val_adj < 0.01), 
                      fun="enricher", TERM2GENE = some.molsig, universe = all.genes)

clgo = mutate(clgo, cluster = factor(cluster, levels=c(0,1:max(as.integer(cluster)))))
table(clgo@compareClusterResult$cluster)
head(clgo)[,c(2,4:9)]
nrow(filter(clgo, p.adjust < 0.01))
summary(clgo@compareClusterResult$p.adjust)

dotplot(filter(clgo,p.adjust < 0.01), x="cluster", showCategory=2)
dotplot(clgo, x="cluster", showCategory=2, includeAll=F)
dotplot(clgo, x="cluster", showCategory=5, includeAll=F) + theme(axis.text.y=element_text(size=8))
```

```{r cluster_go_unique, dev=c("png","pdf")}
dfgo = as.data.frame(clgo)
dfgo$short_desc =drama_shorten(dfgo$Description)
go_to_plot = filter(dfgo, p.adjust < 0.01) %>% group_by(cluster) %>%
              slice_min(n = 2, order_by = p.adjust)
go_to_plot$short_desc = factor(go_to_plot$short_desc, levels=rev(unique(go_to_plot$short_desc)))

p = ggplot(go_to_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
  geom_point()+ 
  theme(axis.text.x=element_text(angle=30, hjust=1), axis.title.y = element_blank()) +  
  scale_color_manual(values=clusters, guide="none") + scale_size(name="Genes", breaks=c(25,50,100,200))+
  scale_x_discrete(name = "Progenitor cluster", labels =function(x) paste0("p",x))
p
```

```{r go_wo_e2m}
go_to_plot = filter(dfgo, p.adjust < 0.01 & !grepl("Epithel", short_desc)) %>% group_by(cluster) %>%
              slice_min(n = 2, order_by = p.adjust)
go_to_plot$short_desc = factor(go_to_plot$short_desc, levels=rev(unique(go_to_plot$short_desc)))

p = ggplot(go_to_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
  geom_point()+ 
  theme(axis.text.x=element_text(angle=30, hjust=1), axis.title.y = element_blank()) +
  scale_color_manual(values=clusters, guide="none") + scale_size(name="Genes", breaks=c(25,50,100,200))+
  scale_x_discrete(name = "Progenitor cluster", labels =function(x) paste0("p",x))
p
```


```{r cluster_go_overlapping, dev=c("png","pdf")}
include_all_plot = filter(dfgo, Description %in% go_to_plot$Description & p.adjust < 0.01)
include_all_plot$short_desc = factor(include_all_plot$short_desc, levels=levels(go_to_plot$short_desc))
a = ggplot(include_all_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
  geom_point()+ 
  theme(axis.text.x=element_text(angle=30, hjust=1), axis.title.y = element_blank()) +
 scale_color_manual(values=clusters, guide="none") + scale_size(name="Genes", breaks=c(25,50,100,200))+
  scale_x_discrete(name = "Progenitor cluster", labels =function(x) paste0("p",x))
a
```

```{r save}
if (subset_data == FALSE){
  write.table(dfgo, here (odir, "ORA_marker_genes.txt"),
            sep="\t", row.names = F)
} else if (subset_data){ #if subset, don't save complete analysis
  #don't write
}
```
