---
title: "TSS_analysis_bvw_initial"
author: "Sarah Hazell Pickering"
date: "2024-04-11"
output: html_document
---

```{r params}
module="adipogenesis"
dataname = "tss_bvw_initial"
library(here); i_am(file.path("analysis", paste0(dataname,".Rmd")))

indir = here("data/scafe/count/day0_to_day3_rep1")
odir = here("output", module, dataname) 

load_data = T #loading data 
subset_data = F

if (!dir.exists(here(odir))){dir.create(here(odir), recursive = T)}
```


```{r libs, warning=F, message=F, echo=F}
library(dplyr)
library(ggplot2)
library(ggsci)
library(Seurat)
library(BPCells)
library(patchwork)
library(clusterProfiler)
library(enrichplot)
library(RColorBrewer)
```


```{r}
pal = c("#1F78B4", "#B2DF8A","#85C2EA","#B3B3B3", "#F18486","#FDBF6F","#E31A1C")
clusters = #ggsci::pal_simpsons()(13)[c(1:7,9:13)]
  tvthemes::gravityFalls_pal()(13)
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))
theme_hp = theme_classic(base_size=15) + 
		  theme(axis.line = element_line(linewidth = 0.25),
		        axis.ticks = element_line(linewidth=0.25),
		        strip.background = element_blank(), # remove boxes around facet labels
		        )
		#set for all ggplots
theme_set(theme_hp)
```


## Read data

Can start from counts if load_data=FALSE, or work on a subset if subset_data=T

```{r read_or_load}
samples = list.dirs(here(indir), full.names = F, recursive = F)

if (load_data==FALSE){# load count matrices
  
  data_list = c()
  
  for (s in samples){
    #if this has been run before delete bp dir
    bc_matrix = Read10X(data.dir = here(indir, s, "matrix "))
    so = CreateSeuratObject(counts=bc_matrix, project=s, names.field = 2, names.delim = "-")
    write_matrix_dir(mat=so[["RNA"]]$counts, dir = here(odir, "bpcells", s))
    gene_mat = open_matrix_dir(here(odir, "bpcells", s))
    so[["RNA"]]$counts = gene_mat
    
    data_list = c(data_list,so)
  }
  names(data_list) = samples
  aggr = merge(data_list[[1]], y= data_list[2:length(data_list)], 
               add.cell.ids = gsub("_",".",samples), project="day0_to_day3" )
  
  saveRDS(aggr,here(odir, "unfiltered_object.rds"))

  
  rm(data_list, bc_matrix, so)
} else if(load_data){
  aggr = readRDS(here(odir, "unfiltered_object.rds"))
}

aggr
head(aggr@meta.data) #auto metadata
```

```{r set_metadata}
aggr$orig.ident = gsub("\\.","_",gsub("_.*", "",rownames(aggr@meta.data)))
aggr$orig.ident = factor(aggr$orig.ident, levels = c("white_day3","white_day1","asc_day0","beige_day1","beige_day3"))
Idents(aggr) = aggr$orig.ident
table(aggr@meta.data$orig.ident)
aggr$condition = gsub("_.*","", aggr$orig.ident)
table(aggr$condition)

aggr$time = gsub(".*_", "", gsub("asc_donor13","day0",aggr$orig.ident))
table(aggr$time)
```

```{r}
pal = c("#1F78B4", "#85C2EA","#B3B3B3", "#F18486","#E31A1C")
levels(aggr$orig.ident)
```

```{r no_cells}
ggplot(aggr@meta.data) + geom_bar(aes(x=orig.ident, fill=orig.ident)) + 
  scale_fill_manual(values=pal) + scale_y_continuous(expand=c(0,0), name="Cells")
```

### Merge layers for straight-forward analysis

```{r}
aggr <- JoinLayers(aggr)
aggr
```

```{r}
if (subset_data){
  aggr = subset(aggr, downsample=1000)
  print(aggr)
}
```



## CRE format
for annotated gene_promoters the gene name is listed after the @ (~ one third (25k) of CREs).  
```{r}
head(Features(aggr)) 
```

Simplify gene names on CREs - if there's a gene name, ignore cluster-specific ids

```{r}
feature_metadata = data.frame(full_id = row.names(aggr[["RNA"]]), row.names = row.names(aggr[["RNA"]]),
                              simple_id = gsub("r[0-9]*@ADDG[0-9]*\\.[A-Z],","",
                                              gsub(",r[0-9]*@ADDG[0-9]*\\.[A-Z]$","",rownames(aggr[["RNA"]]))))
feature_metadata$gene_name = gsub("[DP]:r[0-9]*@", "", feature_metadata$simple_id)
feature_metadata$proximity = gsub(":r[0-9]*@.*", "", feature_metadata$simple_id)

aggr[["RNA"]] = AddMetaData(aggr[["RNA"]], feature_metadata)
row.names(aggr[["RNA"]]) = aggr[["RNA"]][["simple_id"]]$simple_id
head(aggr[["RNA"]])
tail(aggr[["RNA"]])
aggr[["RNA"]][[]]["P:r1@SAMD11",]
```


## Mito QC

```{r before_filt1}
aggr[["percent.mt"]] = PercentageFeatureSet(aggr, pattern="@MT-")
VlnPlot(aggr, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0, cols=pal) 
```

  
```{r before_filter}
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "percent.mt", cols=pal, shuffle=T)
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", cols=pal, shuffle=T)
FeatureScatter(aggr, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", cols=pal, shuffle=T) + 
  coord_cartesian(y=c(0,2500), x=c(0,10000)) + geom_hline(yintercept=1000)
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal, shuffle=T) + 
  geom_hline(aes(yintercept=10), linetype="dashed")
```

### Final filter

Use the maximum 3*mad threshold 

```{r show_filt}
mt_info = group_by(aggr@meta.data, orig.ident) %>% 
  summarise(mad = mad(percent.mt),
            median = median(percent.mt), 
            thresh = median(percent.mt) + mad(percent.mt)*3,
            th95 = quantile(percent.mt, 0.95))
mt_info

thresh = max(mt_info$thresh)
thresh

FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", cols=pal, shuffle=T, pt.size = 0.5) + 
  geom_hline(aes(yintercept=thresh), linetype="dashed")
FeatureScatter(aggr, feature1 = "nFeature_RNA", feature2 = "percent.mt", cols=pal, shuffle=T, pt.size = 0.5) + 
  geom_hline(aes(yintercept=max(mt_info$thresh)), linetype="dashed") +coord_cartesian(ylim=c(0,10))

filt = subset(aggr, subset = percent.mt < thresh & nFeature_RNA > 1000 )
filt #82Mb
```


```{r after_filter}
FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal, shuffle = T)
FeatureScatter(filt, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", pt.size = 0.5, cols=pal, shuffle=T)
FeatureScatter(filt, feature1 = "nFeature_RNA", feature2 = "percent.mt", pt.size = 0.5, cols=pal, shuffle=T)
rm(aggr)
```

## Equalise cell numbers
Downsample to equalise cell numbers across conditions

```{r}
table(Idents(filt))
filt$sample = filt$orig.ident
head(filt@meta.data)
ident_tab = group_by(filt@meta.data, time, condition,sample) %>% count()
ident_tab
min_cells = group_by(ident_tab, time) %>% summarise(min=min(n))
min_cells
```

```{r dwnsmpl}
temp_sbs = list()
for (time in min_cells$time){
  print(time)
  idents = ident_tab$sample[ident_tab$time == time]
  dwns = min_cells$min[min_cells$time == time]
  temp_sbs[[time]] = subset(filt, idents=idents, downsample =dwns)
} 

sbs = merge(x=temp_sbs[[1]],y=temp_sbs[2:length(temp_sbs)])
sbs = JoinLayers(sbs)
sbs$sample = factor(sbs$sample, levels=levels(filt$sample))
sbs$orig.ident = sbs$sample
Idents(sbs) = "sample"
sbs$cond.time = paste0(sbs$condition,".", sbs$time )
sbs
```

```{r}
rm(filt)
```


```{r}
ggplot(sbs@meta.data) + geom_bar(aes(x=sample, fill=sample)) + scale_fill_manual(values=pal) +
  theme(axis.text.x = element_text(angle=15, hjust=1))
```




## Basic Normalisation Scale the read depth

Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. This is then natural-log transformed using log1p
```{r lognorm}
sbs= NormalizeData(sbs, normalization.method = "LogNormalize", scale.factor = 10000)
sbs#161Mb
```


```{r variable_features}
sbs <- FindVariableFeatures(sbs, selection.method = "vst", nfeatures = 2000)
sbs #163Mb
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(sbs), 10)

# plot variable features with and without labels
plot1 = VariableFeaturePlot(sbs)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2

plot1 + coord_cartesian(y=c(0,5))
```


```{r}
all.genes = rownames(sbs)
sbs = ScaleData(sbs, features=all.genes)
```

## PCA

 If features=NULL, PCA will be run using the variable features for the Assay. Note that the features must be present in the scaled data. Any requested features that are not scaled or have 0 variance will be dropped, and the PCA will be run using the remaining features.

```{r pre-pca}
var.feat = VariableFeatures(sbs)
for_pca = subset(sbs, features=var.feat)
for_pca = FindVariableFeatures(for_pca)
head(for_pca[["RNA"]][[]])
for_pca
```



```{r pca}
for_pca = RunPCA(for_pca)
DimPlot(for_pca,shuffle = T, cols = pal)
DimHeatmap(for_pca, dims = 1:6, cells = 500, balanced = TRUE)
```

```{r}
for_pca[["pca"]]
sbs[["pca"]] = for_pca[["pca"]]
```

```{r merged_pca}
DimPlot(sbs, reduction = "pca", cols = pal)
FeaturePlot(sbs, features= c("percent.mt","nCount_RNA","CHI3L1","ACTA2"))
DimHeatmap(sbs, dims = 1:6, cells = 500, balanced = TRUE)
```

## Determine the dimensionality

They encourage to complete downstream analyses with different numbers of PCs.  
```{r elbow}
ElbowPlot(sbs, ndims = 20)
ElbowPlot(sbs, ndims = 50)
```
## Cluster
```{r cluster}
sbs <- FindNeighbors(sbs, dims = 1:20)
sbs <- FindClusters(sbs, resolution = 0.5)
table(sbs@meta.data$seurat_clusters)
```

## Cluster composition

```{r cluster_composition}
composition = group_by(sbs@meta.data, orig.ident,seurat_clusters) %>% count()
composition = composition[order(composition$n, decreasing = T),]
colnames(composition) = c("orig.ident","seurat_clusters","n_cells")
head(composition)

ggplot(composition) + geom_bar(stat="identity",aes(y=orig.ident, x=n_cells, fill=seurat_clusters)) + 
  scale_y_discrete(limits=rev) + scale_fill_manual(values=clusters)
ggplot(composition) + geom_bar(stat="identity",aes(y=orig.ident, x=n_cells, fill=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=clusters)
ggplot(composition) + geom_bar(stat="identity",aes(fill=orig.ident, x=n_cells, y=seurat_clusters)) +
  scale_y_discrete(limits=rev)+ scale_fill_manual(values=pal)
ggplot(composition) + geom_bar(stat="identity",aes(fill=orig.ident, x=n_cells, y=seurat_clusters), position = "fill") +
  scale_y_discrete(limits=rev) + scale_fill_manual(values=pal)
```
## UMAP

```{r umap}
sbs = RunUMAP(sbs, dims=1:20)
DimPlot(sbs, reduction="umap", label=T, shuffle = T, cols=clusters)
DimPlot(sbs, reduction="umap", label=F, shuffle = T, cols=clusters)
DimPlot(sbs, reduction="umap", split.by = "orig.ident", ncol=3, shuffle = T, cols=clusters)
DimPlot(sbs, group.by="orig.ident", cols=pal, shuffle = T)
```



## t-SNE
```{r tsne}
sbs = RunTSNE(sbs, reduction="pca", dims=1:30)
DimPlot(sbs, reduction = "tsne",shuffle=T, cols=clusters)
DimPlot(sbs, reduction = "tsne", group.by= "orig.ident", shuffle = T, cols=pal)
```


## Find biomarkers
Min.pct sets the minimum percentage of cells for a marker to be found in

Does densify, which increases speed at the cost of memory help?  
```{r densify}
norm.markers = FindAllMarkers(sbs, only.pos = T, min.pct=0.25, 
                              logfc.threshold = 0.5, densify=T)
norm.markers = merge(norm.markers,
                     sbs[["RNA"]][[c("simple_id","gene_name","proximity")]],
                     by.x="gene", by.y="simple_id")
norm.markers = arrange(norm.markers, p_val_adj, avg_log2FC)
head(norm.markers); dim(norm.markers)


top.markers = as.data.frame(norm.markers %>%
    group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC))
top.markers
```


Visualisation methods

```{r marker_vis}
to_plot = grep("ACTA2|MKI67|MT-ATP8|CHI3L1", value=T, Features(sbs))

FeaturePlot(sbs, features =to_plot) 

for (i in 1:(length(top.markers$gene)/4)){
  start = ((i-1)*4)+1; print(start)
  end = i*4; print(end)
  print(FeaturePlot(sbs, features = top.markers$gene[start:end]))
}

FeaturePlot(sbs, features = c("percent.mt", "nCount_RNA","nFeature_RNA"))
FeaturePlot(sbs, features = c("PC_1","PC_2","PC_3","PC_4"))
```


```{r}
Idents(sbs) = "seurat_clusters"
norm.markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = avg_log2FC) -> top10
DoHeatmap(subset(sbs,n=100), features= top10$gene)
```

And then we would usually assign cell type identity.  

Be useful to add GO terms here/  Be a little more selective about the marker genes.  Some of the logFC choices seem to favour high cell-cell variability genes

## Cluster GO

```{r load_go}
molsig <- clusterProfiler::read.gmt("/projects/imb-pkbphil/sp/annotations/molsigdb/msigdb.v2024.1.Hs.symbols.gmt")
head(molsig); nrow(molsig)
##Pick gene sets to look in bcos all is a lot
prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))
shorten = function(ont) {
  abbreviate(gsub("_"," ", tolower(ont)),minlength=40, dot=T, named = F)
}
```

```{r}
norm.markers = rename(norm.markers, simple_id=gene) #switch column names
norm.markers = rename(norm.markers, gene=gene_name)
norm.markers$cluster = factor(norm.markers$cluster, levels =0:11)
all.genes = sbs[["RNA"]][["gene_name"]]
```


```{r basic_go}
go = compareCluster(gene~cluster, data=norm.markers, 
                      fun="enricher", TERM2GENE = some.molsig, universe = all.genes)
head(go)[,c(2,4:9)]

p = dotplot(go, x="cluster", showCategory=2) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))
p = dotplot(go, x="cluster", showCategory=2, includeAll=F) 
p+ scale_y_discrete(labels=shorten(levels(p$data$Description)))
```


```{r save}
if (subset_data == F){
  saveRDS(sbs, file = here(odir,"complete_analysis.rds"))
    write.table(composition, here(odir,"cluster_composition.tsv"),
            sep="\t", row.names = F, quote=F)
  write.table(norm.markers, here(odir,"marker_genes.txt"),
            sep="\t", row.names = F)
  write.table(go, here(odir,"ORA_marker_genes.txt"),
            sep="\t", row.names = F)
  Idents(sbs) = "sample"
  dwns = subset(sbs, downsample=1000)
  saveRDS(dwns, file = here(odir,"10%_complete_analysis.rds"))
} else if (subset_data){
  saveRDS(sbs, file = here(odir,"10%_subset_analysis.rds"))
}
```


