---
title: "markers_and_go"
author: "Sarah Hazell Pickering"
date: "2024-08-29"
output: html_document
---

```{r params}
module="adipogenesis"
dataname = "white_only_downsample"
library(here); i_am(file.path("analysis",paste0(dataname, ".Rmd")))
odir = here("output", module, dataname) 

load_data = T #loading data 
subset_data = F
```


```{r libs, warning=F, message=F, echo=F}
library(dplyr)
library(ggplot2)
library(ggsci)
library(patchwork)

library(Seurat)
library(BPCells)

library(clusterProfiler)
library(enrichplot)

knitr::opts_chunk$set(dev=c('png','pdf'))
```

## themes

```{r themes}
pal = c("#B3B3B3","#85C2EA","#B2DF8A","#1F78B4")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
names(clusters) = 0:11
figure_order = as.character(c(8,3,1,11,7,0,9,10,6,4,2,5))
spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))

theme_hp = theme_classic(base_size=15) + 
		  theme(axis.line = element_line(linewidth = 0.25),
		        axis.ticks = element_line(linewidth=0.25),
		        strip.background = element_blank(), # remove boxes around facet labels
		        )
		#set for all ggplots
theme_set(theme_hp)
```
## Load seurat obj
```{r}
if (subset_data){
  sbs = readRDS(here(odir, "10%_subset_analysis.rds"))
} else {
  sbs = readRDS(here(odir, "complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
sbs
#format clusters
sbs$fclusters = factor(sbs$seurat_clusters, levels=figure_order)
```
### UMAP
```{r umap}
DimPlot(sbs, reduction="umap", group.by="time", cols=pal[c(1,2,4)]) + theme_hp + 
  theme(plot.title=element_blank())
DimPlot(sbs, reduction="umap", cols=clusters, pt.size = 0.4) + theme_hp +
  scale_colour_manual(labels=function (x){gsub("^","a",x)}, name="Adipogenesis\nclusters",
                        values = clusters)
```


```{r}
clgo = read.delim(here(odir, "ORA_marker_genes.txt"))
raw_go_terms = clgo %>% slice_min(n=2, order_by = p.adjust, by=cluster) %>% pull(ID)
upr_go_terms = clgo %>%  filter(cluster %in% c(6,8)) %>% slice_min(n=2, order_by = p.adjust, by=cluster) %>% pull(ID)
head(raw_go_terms)
head(upr_go_terms)
```


```{r go_replace}
phrases = c("protein_containing_complex" = "complex",
    "structure_organization" = "structure",
    "_(WITH|INCLUDES_BOTH)_BLOCH_AND_.*"=":_BLOCH_&_KANDut.", #simplify to Cholesterol metabolism... 
      "CELLULAR" = "CELL", 
      "response_to_topologically_incorrect_protein" = "topologically_incorrect_protein_response", #Synonym 
     "_DRIVEN_ACTIVE_TRANSMEMBRANE"= "",#OXIDOREDUCTION...TRANSPORTER_ACTIVITY
      "_AND_RESPIRATORY_ELECTRON_TRANSPORT" ="_&_electron_transport",#AEORBIC RESPIRATION...
      "ENDOPLASMIC_RETICULUM"="ER",
    "the_citric_acid"="")
phrases

recap <- c(
    "rna" = "RNA",
    "mrna" = "mRNA",
    "rrna" = "rRNA",
    "ncrna"="ncRNA",
    "tnfa" = "TFNa",
    "nfkb" = "NFKb",
    "mtorc1" = "mTORC1",
    "bmp" = "BMP",
    "e2f" = "E2F",
    "uv" = "UV",
    "er"="ER",
    "upr"="UPR",
    "tca"="TCA")
recap
```


Function to intelligently abbreviate GO terms and correct the capitalisation of initialisations based on the above lists.  To do: save these lists to file to make reproducible and concise.   

```{r abrev_go}
drama_shorten = function(ont) {
  for (x in 1:length(phrases)){
    ont = gsub( names(phrases)[x],phrases[x], ont, ignore.case = T)
  }
  a = abbreviate(gsub("_"," ", gsub("^[[:alpha:]]*_","",tolower(ont))),minlength=41, dot=T, named = F)
  a = stringr::str_to_sentence(a)
  for (i in 1:length(recap))
    a = gsub(sprintf("\\b%s\\b", names(recap)[i]),recap[i],a, ignore.case = T)
  return(a)
}

drama_shorten(raw_go_terms)
#upr_go_terms
drama_shorten(upr_go_terms)
```

## Marker genes

```{r}
all.genes = rownames(sbs)
```

```{r}
norm.markers = read.delim(here(odir, "marker_genes.txt"))
norm.markers = filter(norm.markers, p_val_adj < 0.01)
table(norm.markers$cluster)
norm.markers$cluster = factor(norm.markers$cluster, levels=0:11)
uniq.norm.markers = norm.markers[!norm.markers$gene %in% names(which(table(norm.markers$gene) > 1)),]
head(norm.markers)
```

### Genes sorted logFC * pct.1
```{r}
top.markers = as.data.frame(norm.markers %>%
    group_by(cluster) %>% filter(pct.1 > 0.7) %>% #also adding a minimum filter to show CD36 instead of LINC01197 for cluster a11
    slice_max(n = 2, order_by = avg_log2FC*pct.1))
top.markers
```
 
```{r dotplot}
DotPlot(sbs, features=rev(unique(top.markers$gene)), group.by = "seurat_clusters") & 
  scale_colour_gradientn(colours = spectral) & coord_flip()
```


```{r top_markers_logfc_pct1}
top.markers$gene = factor(top.markers$gene, levels=rev(unique(top.markers$gene)))
markers.to.plot = filter(top.markers, gene %in% top.markers$gene)
markers.to.plot$gene = factor(markers.to.plot$gene, levels=levels(top.markers$gene))

ggplot(markers.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point()  +
  scale_color_manual(values=clusters, guide="none") +
  scale_x_discrete(labels=function (x){gsub("^","a",x)}, name="Adipogenesis clusters") +
  ylab("") + scale_size_continuous(name="log2FC")

markers.to.plot = mutate(markers.to.plot, cluster = factor(cluster, levels=figure_order)) %>%
  arrange(cluster)
markers.to.plot$gene = factor(markers.to.plot$gene, levels=unique(rev(markers.to.plot
                                                                  $gene)))

ggplot(markers.to.plot, aes(y=gene, size=abs(avg_log2FC), color=cluster,
                    x=cluster)) + 
  geom_point()  +
  scale_color_manual(values=clusters, guide="none") +
  scale_x_discrete(labels=function (x){gsub("^","a",x)}, name="Adipogenesis clusters") +
  ylab("") + scale_size_continuous(name="log2FC")
```

### Gene Overlaps plots

```{r top_markers_overlap}
include_all_plot = filter(norm.markers, gene %in% markers.to.plot$gene & p_val_adj< 0.01)
include_all_plot$gene = factor(include_all_plot$gene, levels=levels(top.markers$gene))

a = ggplot(include_all_plot, aes(x=cluster, y=gene, colour=cluster, size=avg_log2FC)) + 
  geom_point()+  
    scale_color_manual(values=clusters, guide="none") +
  scale_x_discrete(labels=function (x){gsub("^","a",x)}, name="Adipogenesis clusters") +
  ylab("") + scale_size_continuous(name="log2FC")
a
```


```{r top_markers_overlap_time_order}
include_all_plot = mutate(include_all_plot, cluster=factor(cluster, levels=figure_order)) %>% arrange()
include_all_plot$gene = factor(include_all_plot$gene, levels=levels(markers.to.plot$gene))
a = ggplot(include_all_plot, aes(x=cluster, y=gene, colour=cluster, size=avg_log2FC)) + 
  geom_point()+  
    scale_color_manual(values=clusters, guide="none") +
  scale_x_discrete(labels=function (x){gsub("^","a",x)}, name="Adipogenesis clusters") +
  ylab("") + scale_size_continuous(name="log2FC")
a
```


## GO plots

```{r}
dfgo = read.delim(here(odir,"ORA_marker_genes.txt"))
```



```{r load_go}
molsig <- clusterProfiler::read.gmt(here("data/gene_sets/msigdb.v2024.1.Hs.symbols.gmt"))
head(molsig); nrow(molsig)
##Pick gene sets to look in bcos all is a lot
prefixes = c("HALLMARK", "KEGG","REACTOME", "WP", "GOBP", "GOCC","GOMF")
colnames(molsig) = c("term","gene")
some.molsig = molsig[gsub("_.*","", molsig$term) %in% prefixes,]
some.molsig$term = factor(some.molsig$term)
table(gsub("_.*","", some.molsig$term))
```

### coloured plot

```{r go_all}
dfgo$cluster = factor(dfgo$cluster, levels=0:max(as.integer(dfgo$cluster)))
dfgo$short_desc =drama_shorten(dfgo$Description)
go_to_plot = filter(dfgo, p.adjust < 0.01) %>% group_by(cluster) %>%
              slice_min(n = 2, order_by = data.frame(p.adjust, pvalue))
go_to_plot$short_desc = factor(go_to_plot$short_desc, levels=rev(unique(go_to_plot$short_desc)))

p = ggplot(go_to_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
  geom_point()+   
  theme(axis.text.x=element_text(angle=30, hjust=1)) +scale_color_manual(values=clusters)
p
```

```{r go_wo_wp}
go_to_plot = filter(dfgo, p.adjust < 0.01 & !grepl("^wp", short_desc)) %>% group_by(cluster) %>%
              slice_min(n = 3, order_by = data.frame(p.adjust,pvalue))
go_to_plot$short_desc = factor(go_to_plot$short_desc, levels=rev(unique(go_to_plot$short_desc)))

p = ggplot(go_to_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
  geom_point()+
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  scale_color_manual(values=clusters, guide="none") +
  scale_x_discrete(labels=function (x){gsub("^","a",x)}, name="Adipogenesis clusters") +
  ylab("") + scale_size_continuous(name="Genes")
p
```


```{r cluster_go_overlapping}
include_all_plot = filter(dfgo, Description %in% go_to_plot$Description & p.adjust < 0.01)
include_all_plot$short_desc = factor(include_all_plot$short_desc, levels=levels(go_to_plot$short_desc))
a = ggplot(include_all_plot, aes(x=cluster, y=short_desc, colour=cluster, size=Count)) + 
  geom_point()+
  theme(axis.text.x=element_text(angle=30, hjust=1)) +scale_color_manual(values=clusters)
a
ggsave(here(odir, "cluster_go_overlapping.pdf"), height = 7, width=9)
```


## GO terms Heatmap

### Setup
step1: convert df of top 2 Go terms per cluster to a list of Go terms and genes
```{r}
top2_terms = go_to_plot[["ID"]]
top2_genelist = list()
for (n in unique(top2_terms)){
  top2_genelist[[n]] = some.molsig$gene[some.molsig$term == n]
}
str(top2_genelist)
```


```{r}
sbs = AddModuleScore(sbs, features = top2_genelist, name="top2GO")
colnames(sbs@meta.data)[grep("top2GO",colnames(sbs@meta.data))] = names(top2_genelist)
DotPlot(sbs, features=names(top2_genelist)) + coord_flip() + 
  scale_colour_gradientn(colours=spectral) + 
  scale_x_discrete(labels=drama_shorten(names(top2_genelist)))
```
^^ dots are not necesarily helpful because they're mostly large.  

### matrix formatting
```{r}
library(ComplexHeatmap)
head(sbs@meta.data)
filt = subset(sbs, downsample=100)
go_matrix= filt@meta.data[names(top2_genelist)]
cell_ids = filt@meta.data[c("time","fclusters")]
go_matrix = t(scale(data.matrix(go_matrix)))
go_matrix[1:3,1:3]
```


This is much better than the dotplot :)  But it also really shows how inadequate the cluster and GO analysis is.  Which is maybe fundamental to discretising a dynamic process.   

### Heatmap

```{r go_heatmap}
rownames(go_matrix) = drama_shorten(rownames(go_matrix))
ha = HeatmapAnnotation(cluster = cell_ids$fclusters, col = list(cluster=clusters), show_legend = F,
                       show_annotation_name = F)
h =Heatmap(go_matrix, show_column_names = F, column_split = cell_ids$fclusters, 
        cluster_columns = F, show_heatmap_legend = F, top_annotation = ha, row_names_max_width = unit(9,"cm"))
h

```


```{r heatmap_legend}
h =Heatmap(go_matrix, show_column_names = F, column_split = cell_ids$fclusters, 
        cluster_columns = F, top_annotation = ha, row_names_max_width = unit(9,"cm"), name="scaled expression")
h

```


## Adipogenesis markers

```{r}
FeaturePlot(sbs, reduction="tsne", features=c("ADIPOQ","LPL","PPARG","CEBPA"))
```

```{r}
FeaturePlot(sbs, reduction="tsne", features=c("MGP","DCN","CEBPD","CEBPB"))
```



## HALLMARK gene sets

```{r}
sbs = AddModuleScore(sbs, features = list(upr = some.molsig$gene[some.molsig$term == "HALLMARK_UNFOLDED_PROTEIN_RESPONSE"],
                                    adipo = some.molsig$gene[some.molsig$term == "HALLMARK_ADIPOGENESIS"],
                                    erstress = some.molsig$gene[some.molsig$term == "GOBP_RESPONSE_TO_ENDOPLASMIC_RETICULUM_STRESS"]),
               name="hallmark")
colnames(sbs@meta.data)[grep("hallmark",colnames(sbs@meta.data))] = c("UPR","ADIPO","ERSTRESS")
```

```{r adipo_er_featureplots}
FeaturePlot(sbs,  features = c("UPR","ADIPO","ERSTRESS")) & scale_color_gradientn(colours=spectral)
FeaturePlot(sbs,features = c("UPR"), pt.size=0.4) & scale_color_gradientn(colours=spectral)
FeaturePlot(sbs, features = c("ERSTRESS"), pt.size=0.4) & scale_color_gradientn(colours=spectral)
FeaturePlot(sbs,  features = c("ADIPO"), pt.size=0.4) & scale_color_gradientn(colours=spectral)
FeaturePlot(sbs,  features = c("ADIPO","ERSTRESS"), blend=T,cols=c("lightgrey","red","blue"), order=T, pt.size=0.4 ) 
FeaturePlot(sbs,  features = c("ADIPO","UPR"), blend=T,cols=c("lightgrey","red","blue"), order=T, pt.size=0.4 ) 
```


###UPR vs Adipogenesis
```{r upr_violin}
Idents(sbs) = "fclusters"
VlnPlot(sbs, features=c("UPR","ADIPO","ERSTRESS"), stack=T, flip=T, fill.by = "ident",  cols=clusters)
```

```{r upr_scatter}
FeatureScatter(sbs, feature1 = "ADIPO", feature2 = "ERSTRESS", ncol=3, cols=clusters, shuffle=T, pt.size = 0.5) &
  geom_hline(aes(yintercept=0.2))&
  geom_vline(aes(xintercept=0.2))

FeatureScatter(sbs, feature1 = "ADIPO", feature2 = "ERSTRESS",split.by = "seurat_clusters", ncol=4,  cols=clusters, pt.size = 0.2) &
  geom_hline(aes(yintercept=0.2))&
  geom_vline(aes(xintercept=0.2)) & geom_smooth(method="lm", se=F, colour="red", linetype="dashed", linewidth=0.5)

FeatureScatter(sbs, feature1 = "ADIPO", feature2 = "ERSTRESS", split.by = "time", ncol=2, cols=clusters, shuffle = T, pt.size = 0.5) &
  geom_hline(aes(yintercept=0.2))&
  geom_vline(aes(xintercept=0.2))

FeatureScatter(sbs, feature1 = "ADIPO", feature2 = "ERSTRESS", ncol=2, shuffle = T, pt.size = 0.5,
               cells= colnames(sbs)[sbs$seurat_clusters %in%  c(5,6,8)], cols=clusters[c(6,7,9)]) &
  geom_hline(aes(yintercept=0.2))&
  geom_vline(aes(xintercept=0.2))

FeatureScatter(sbs, feature1 = "ADIPO", feature2 = "ERSTRESS", ncol=2,  shuffle = T, pt.size = 0.5,
               cells= colnames(sbs)[sbs$seurat_clusters %in%  c(6)], cols=clusters[7]) &
  geom_hline(aes(yintercept=0.2))&
  geom_vline(aes(xintercept=0.2)) & geom_smooth(method="lm", se=F, colour="red", linetype="dashed", linewidth=0.5)
```


## UPR arms

```{r upr_3_arms}
sbs = AddModuleScore(sbs, features = list(perk = some.molsig$gene[some.molsig$term == "GOBP_PERK_MEDIATED_UNFOLDED_PROTEIN_RESPONSE"],
                                    atf6 = some.molsig$gene[some.molsig$term == "GOBP_ATF6_MEDIATED_UNFOLDED_PROTEIN_RESPONSE"],
                                    ire1 = some.molsig$gene[some.molsig$term == "GOBP_IRE1_MEDIATED_UNFOLDED_PROTEIN_RESPONSE"]),
               name="upr_pathways")
colnames(sbs@meta.data)[grep("upr_pathways",colnames(sbs@meta.data))] = paste0("upr_", c("perk","atf6","ire1"))
```

```{r}
VlnPlot(sbs, features=c("upr_perk","upr_atf6","upr_ire1"), stack=T, flip=T, fill.by = "ident",  cols=clusters)
```

```{r, dev=c('png')}
filt = subset(sbs, downsample=100)

DoHeatmap(filt, features= some.molsig$gene[some.molsig$term == "GOBP_PERK_MEDIATED_UNFOLDED_PROTEIN_RESPONSE"],
          group.colors = clusters)

DoHeatmap(filt, features= some.molsig$gene[some.molsig$term == "GOBP_ATF6_MEDIATED_UNFOLDED_PROTEIN_RESPONSE"],
          group.colors = clusters)

DoHeatmap(filt, features= some.molsig$gene[some.molsig$term == "GOBP_IRE1_MEDIATED_UNFOLDED_PROTEIN_RESPONSE"],
          group.colors = clusters)
```
```{r}
table(grep("ATF6",some.molsig$term, value=T))
```

## UPR marker genes
###SLC3A2


```{r scl3a2}
FeaturePlot(sbs, features = "SLC3A2") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "SLC3A2", cols=clusters, pt.size = 0) + 
    scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```

### HSPA5 aka BiP

```{r hspa5}
FeaturePlot(sbs, features = "HSPA5") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "HSPA5", cols=clusters, pt.size = 0)+ 
    scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```

### XBP1

```{r xbp1}
FeaturePlot(sbs, features = "XBP1") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "XBP1", cols=clusters, pt.size = 0)+ 
    scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```

## Adipogenesis marker genes
### PPARG

```{r pparg}
FeaturePlot(sbs, features = "PPARG") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "PPARG", cols=clusters, pt.size = 0)+ 
   scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```

### ADIPOQ

```{r adipoq}
FeaturePlot(sbs, features = "ADIPOQ") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "ADIPOQ", cols=clusters, pt.size = 0)+ 
    scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```


### FABP4

```{r fabp4}
FeaturePlot(sbs, features = "FABP4") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "FABP4", cols=clusters, pt.size = 0)+ 
    scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```


### LPL

```{r lpl}
FeaturePlot(sbs, features = "LPL") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "LPL", cols=clusters, pt.size = 0)+ 
   scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```

### FASN

```{r fasn}
FeaturePlot(sbs, features = "LPL") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "LPL", cols=clusters, pt.size = 0)+ 
    scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```


## Progenitor markers (for comparison)

### CD36

```{r cd36}
FeaturePlot(sbs, features = "CD36") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "CD36", cols=clusters, pt.size = 0)+ 
    scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```


### ACTA2

```{r acta2}
FeaturePlot(sbs, features = "ACTA2") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "ACTA2", cols=clusters, pt.size = 0)+ 
  scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```



## SWAT markers

### MGP

```{r mgp}
FeaturePlot(sbs, features = "MGP") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "MGP", cols=clusters, pt.size = 0)+ 
  scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```


### DCN

```{r dcn}
FeaturePlot(sbs, features = "DCN") & scale_color_gradientn(colours=spectral)
VlnPlot(sbs, features = "DCN", cols=clusters, pt.size = 0)+ 
   scale_x_discrete(limits =figure_order, labels=function(x){gsub("^","a",x)}, 
                   name="Adipogenesis clusters")  + theme_hp
```








