---
title: "fastmnn_progenitor_trajectory_markers"
author: "Sarah Hazell Pickering"
date: "2025-01-29"
output: html_document
---

## Setup
```{r params}
module="adipogenesis"
dataname = "white_only_downsample_res0.8"

library(here); i_am(file.path("analysis",paste0(dataname,"_cluster_composition.Rmd")))
load_data = T #loading data 
subset_data = F
odir = here("output", module, dataname)
```

```{r libs, warning=F, message=F, echo=F}
library(dplyr)
library(ggplot2)
library(Seurat)
library(BPCells)
library(RColorBrewer)
library(viridis)
```


```{r themes}
prog_cols = c("#DA1819","#691A93","#EBB400" , "#434D51","#00AED1","#3AAD00","#FF7F00","#AC8AD0")
pal = c("#B3B3B3","#85C2EA","#1F78B4")
clusters = ggsci::pal_simpsons()(13)[c(1:7,9:13)]
clusters = c(clusters, "purple4",  "#ff1493","#474747", "#9248A7" )
  #tvthemes::gravityFalls_pal()(16)
names(clusters) = 0:(length(clusters)-1)
figure_order = as.character(c(13,12,10,4,7,1,15,9,0,2,14,11,8,3,5,6))

spectral = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral"))

theme_hp = theme_classic(base_size=15) + 
		  theme(axis.line = element_line(linewidth = 0.25),
		        axis.ticks = element_line(linewidth=0.25),
		        strip.background = element_blank(), # remove boxes around facet labels
		        )
		#set for all ggplots
theme_set(theme_hp)
knitr::opts_chunk$set(dev=c("png","pdf"))
```


## Load seurat object
```{r}
if (subset_data){
  sbs = readRDS(here(odir,"10%_complete_analysis.rds"))
  Idents(sbs) = "seurat_clusters"
} else {
  sbs = readRDS(here(odir,"complete_analysis.rds"))
  DefaultAssay(sbs) = "RNA"
}
sbs
variable.genes = VariableFeatures(sbs)
```


### Add progenitor clusters

```{r}
pro_clusters = here("output/progenitors/integration_trial_res0.3")
if (subset_data){
  pro = readRDS(here(pro_clusters, "10%_complete_analysis.rds"))
} else{
  pro = readRDS(here(pro_clusters, "complete_analysis.rds"))
}
pro
table(pro$rpca_clusters)
head(pro$rpca_clusters)
tail(pro$rpca_clusters)
```


```{r}
pro_clusters = pro$rpca_clusters
names(pro_clusters) = gsub("day0.subjectS6","asc.day0",names(pro_clusters))
head(pro_clusters[rownames(sbs@meta.data)])
sbs@meta.data$progen_clusters = pro_clusters[rownames(sbs@meta.data)]
table(sbs$progen_clusters, useNA="ifany")
```


```{r}
sbs$if_progen_cluster = paste0("a",sbs$seurat_clusters)
sbs$if_progen_cluster[!is.na(sbs$progen_clusters)] = paste0("p",sbs$progen_clusters[!is.na(sbs$progen_clusters)])
un.pc = unique(sbs$if_progen_cluster)
sbs$if_progen_cluster = factor(sbs$if_progen_cluster , 
                               levels=un.pc[order(gsub("[[:digit:]]*$","",un.pc),
                                                  as.integer(gsub("p|w","", un.pc)))] )
table(sbs$if_progen_cluster, useNA="ifany")
```

```{r}
sbs$time.progen_clusters = sbs$time
sbs$time.progen_clusters[!is.na(sbs$progen_clusters)] = paste0("p",sbs$progen_clusters[!is.na(sbs$progen_clusters)])
sbs$time.progen_clusters = factor(sbs$time.progen_clusters,
                                  levels=c(paste0("p",0:7), "day0","day1","day3")) #approx pseuodtime ordering
table(sbs$time.progen_clusters)
rm(pro)
```

## Composition

```{r cluster_composition}
composition = group_by(sbs@meta.data, orig.ident,seurat_clusters, time.progen_clusters) %>% dplyr::count()
composition = composition[order(composition$n, decreasing = T),]
colnames(composition) = c("orig.ident","seurat_clusters","progen_clusters","n_cells")
#calculate proportions directly, then plot percentage just for the three cyc;ling cells
composition = group_by(composition, orig.ident) %>% mutate(cells_per_sample = n_cells/sum(n_cells),
                                                           time =gsub("(white|asc)_","", gsub("_rep2","",orig.ident)))
composition = group_by(composition, seurat_clusters) %>% mutate(cells_per_cluster = n_cells/sum(n_cells))
head(composition)
```

```{r}
filter(composition, time== "day0") %>%
  ggplot( aes(x=n_cells, y=seurat_clusters, fill=progen_clusters)) + 
  geom_bar(stat="identity") + scale_fill_manual(values=c(prog_cols,"grey")) +
  scale_y_discrete(name = "res0.8_adipogenesis")
```

## Basic TSNEs

```{r basic_tsne}
DimPlot(sbs, group.by="time", cols=pal, reduction ="tsne")
DimPlot(sbs, group.by="initial_clusters", cols=clusters, reduction ="tsne")
```



## Best P1/P2 markers
```{r best_p1.p2_markers}
VlnPlot(sbs,  features=c("MGP","CDKN1C","CCN5","ICAM1","CXCL8","TNFAIP6"), stack = T, flip = T, 
        group.by = "time.progen_clusters", split.by="time.progen_clusters", cols=c(prog_cols,pal[1:3]))  +
  scale_x_discrete(limits=c(c(paste0("p",c(0,4,3,2,1)), "day1","day3"))) #remove day0 for example
```



## cluster P1 markers


- cells with more ECM (MGP) are closer to day1 cells 
- cells with more cell cycle arrest (CDKN1A/C) are closer to day1 cells
- cells with more WNT signalling (CCN5) are closer to day1 and also undifferentiated day3 cells
- cells with antioxidant expression (SOD2) are closer to day1; but also to the SWAT trajectory

```{r tsne_p1_markers}
FeaturePlot(sbs,"MGP", reduction = "tsne",order = T, pt.size = 0.5) & scale_colour_gradientn(colours=spectral)#ECM marker
FeaturePlot(sbs,"CDKN1A", reduction = "tsne",order = T, pt.size = 0.5) & scale_colour_gradientn(colours=spectral) #cell cycle arrest
FeaturePlot(sbs,"CDKN1C", reduction = "tsne",order = T, pt.size = 0.5) & scale_colour_gradientn(colours=spectral)
FeaturePlot(sbs,"CCN5", reduction = "tsne",order = T, pt.size = 0.5) & scale_colour_gradientn(colours=spectral) #WNT signalling
FeaturePlot(sbs,"SOD2", reduction = "tsne",order = T, pt.size = 0.5) & scale_colour_gradientn(colours=spectral) #Superoxide dismutase
```

```{r}
VlnPlot(sbs, split.by="initial_clusters", features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), stack = T, flip = T, 
        group.by = "initial_clusters", cols=clusters)
```


```{r}
VlnPlot(sbs,  features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), stack = T, flip = T, 
        group.by = "time.progen_clusters", split.by="time.progen_clusters", cols=c(pal[1:3],prog_cols ))  +
  scale_x_discrete(limits=c(c(paste0("p",c(5,6,7,4,0,3,1,2)), "day0","day1","day3")))

VlnPlot(sbs,  features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), stack = T, flip = T, 
        group.by = "time.progen_clusters", split.by="time.progen_clusters", cols=rep(prog_cols[2],11))  +
  scale_x_discrete(limits=c(c(paste0("p",c(5,6,7,4,0,3,1,2)), "day0","day1","day3")))

VlnPlot(sbs,  features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), stack = T, flip = T, 
        group.by = "time.progen_clusters", split.by="time.progen_clusters", cols=c(pal[1:3],prog_cols))  +
  scale_x_discrete(limits=c(c(paste0("p",c(4,0,3,1,2)), "day1","day3"))) #remove day0 for example
#kansje highlight cluster with target gene expr (p1)
```

```{r}
VlnPlot(sbs,  features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), stack = T, flip = T, 
        group.by = "if_progen_cluster", split.by="if_progen_cluster", cols=c(prog_cols,clusters[c(1:4,6:14)])) 

VlnPlot(sbs,  features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), stack = T, flip = T, 
        group.by = "if_progen_cluster", split.by="if_progen_cluster", cols=c(prog_cols,clusters[c(1:4,6:14)])) +
  scale_x_discrete(limits=c(paste0("p",c(4,0,3,1,2)), paste0("a",c(0,7,2,3,5))))

VlnPlot(sbs,  features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), stack = T, flip = T, 
        group.by = "if_progen_cluster", split.by="if_progen_cluster", cols=c(prog_cols,clusters[c(1:4,6:14)])) +
  scale_x_discrete(limits=c(paste0("p",c(4,0,3,1,2)), paste0("a",c(0,7,2,3,5))))
```

```{r}
Idents(sbs) = "if_progen_cluster"
DoHeatmap(subset(sbs,downsample=100), features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), 
          group.colors= c(prog_cols,clusters[c(1:4,6:14)])) 

Idents(sbs) = "time.progen_clusters"
DoHeatmap(subset(sbs,downsample=500), features=c("MGP","CDKN1A","CDKN1C","CCN5","SOD2"), 
          group.colors= c(pal[1:3],prog_cols,clusters[c(1:4,6:14)])) 
```

## Cluster P2 trajectories

```{r tnse_p2_markers}
p2_markers = c("PHLDA1",#top
               "ICAM1",#previously referenced
               "NFKB2",
               "CXCL8", #
               "TNFAIP6" #top tnfa
               )
for (gene in p2_markers){
  p = FeaturePlot(sbs,gene, reduction = "tsne",order = T, pt.size = 0.5) & 
    scale_colour_gradientn(colours=spectral) & ggtitle(gene)
  print(p)
}
```

- PHLDA1 also highly expressed in day1 
- ICAM1 expressed at the border with day1; but not specific to cluster P2 (also cluster P0)
- NFKB2 - slightly more positive cells on the day0/day1 border
- CXCL8 & TNFAIP6 - specific localisation at the day0/day1 border, especially close to the undifferentiated trajectory

```{r}
VlnPlot(sbs,  features=p2_markers, stack = T, flip = T, 
        group.by = "if_progen_cluster", split.by="if_progen_cluster", cols=c(prog_cols,clusters[c(1:4,6:14)])) 
```

 
```{r}
VlnPlot(sbs,  features=p2_markers, stack = T, flip = T, 
        group.by = "if_progen_cluster", split.by="if_progen_cluster", cols=c(prog_cols,clusters[c(1:4,6:14)])) +
  scale_x_discrete(limits=c(paste0("p",c(4,0,3,1,2)), paste0("a",c(0,7,2,3,5))))
```
```{r}
VlnPlot(sbs,  features=p2_markers, stack = T, flip = T, 
        group.by = "time.progen_clusters", split.by="time.progen_clusters", cols=c(pal[1:3],prog_cols)) 

VlnPlot(sbs,  features=p2_markers, stack = T, flip = T, 
        group.by = "time.progen_clusters", split.by="time.progen_clusters", cols=c(pal[1:3],prog_cols))  +
  scale_x_discrete(limits=c(c(paste0("p",c(4,0,3,1,2)), "day1","day3"))) #remove day0 for example
```
### Marker of CD142 (aka F3)
```{r}
FeaturePlot(sbs,"F3", reduction = "tsne", pt.size = 0.5) & 
    scale_colour_gradientn(colours=spectral) 
```

## Load monocle pseudotime

```{r}
if (subset_data){
  mono = readRDS(here(odir,"10%_monocle_analysis_tsne.rds"))
} else {
  mono = readRDS(here(odir,"monocle_complete_analysis_tsne.rds"))
}
mono
```


```{r}
head(pro_clusters[colnames(mono)])
mono$progen_clusters = pro_clusters[colnames(mono)]
table(mono$progen_clusters, useNA="ifany")
```


```{r}
mono$if_progen_cluster = paste0("a",mono$initial_clusters)
mono$if_progen_cluster[!is.na(mono$progen_clusters)] = paste0("p",mono$progen_clusters[!is.na(mono$progen_clusters)])
table(mono$if_progen_cluster, useNA="ifany")
```

```{r}
mono$time.progen_clusters = mono$time
mono$time.progen_clusters[!is.na(mono$progen_clusters)] = paste0("p",mono$progen_clusters[!is.na(mono$progen_clusters)])
table(mono$time.progen_clusters)
```


```{r}
pick_root_node = function(cds,factor_name, start_bin){
  cell_ids = which(colData(cds)[,factor_name]==start_bin)
  closest_vertex <-  cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  summary(closest_vertex)
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <- igraph::V(principal_graph(cds)[["UMAP"]])$name[as.numeric(names
    (which.max(table(closest_vertex[cell_ids,]))))]
return(root_pr_nodes)
}
mono= order_cells(mono, root_pr_nodes = pick_root_node(mono, "initial_clusters", "5"))
```

```{r tsne_monocle_prog_cols}
colnames(colData(mono))
p = plot_cells(mono, color_cells_by = "time.progen_clusters",
           label_leaves=FALSE,
           trajectory_graph_color ="black",#"#F18486", #"#CB6773",
           trajectory_graph_segment_size = 0.5,
           label_branch_points=FALSE, cell_size = 0.5,
           label_cell_groups = F,
           label_roots = T, graph_label_size=4) + 
  scale_colour_manual(values =c(prog_cols, pal[2:3]), limits=c(paste0("p",0:7), paste0("day",c(1,3))))
p$layers[[1]]$aes_params$colour = 'transparent'
p
```

```{r tsne_monocle_initial_clusters}
p = plot_cells(mono, color_cells_by = "initial_clusters",
           label_leaves=FALSE,
           trajectory_graph_color ="black",#"#F18486", #"#CB6773",
           trajectory_graph_segment_size = 0.5,
           label_branch_points=FALSE, cell_size = 0.5,
           label_cell_groups = F,
           label_roots = T, graph_label_size=4) + 
  scale_colour_manual(values =c(clusters))
p$layers[[1]]$aes_params$colour = 'transparent'
p
```


```{r tsne_monocle_traj_pseudotime}

p = plot_cells(mono, color_cells_by = "pseudotime",
           label_leaves=FALSE,
           trajectory_graph_color ="#B31417",#"#F18486", #"#CB6773",
           trajectory_graph_segment_size = 0.5,
           label_branch_points=FALSE, cell_size = 0.5,
           label_cell_groups = F,
           label_roots = T, graph_label_size=4) + scale_colour_viridis(transform="reverse", direction=-1)
p$layers[[1]]$aes_params$colour = 'transparent'
p
```




```{r tsne_monocle_pseudotime}
summary(colnames(mono) == rownames(sbs@meta.data))
sbs$monocle_pseudotime =  scales::rescale(pseudotime(mono), to = c(1, 0)) #rescale from 1 to 0 (cos were tracing backwards from  day3)
FeaturePlot(sbs, "monocle_pseudotime", reduction = "tsne", pt.size = 0.5)+ scale_colour_viridis(direction=-1)
```
```{r}
rm(mono)
```

```{r}
VlnPlot(sbs, "monocle_pseudotime", pt.size = 0, cols =clusters, group.by = "initial_clusters")
VlnPlot(sbs, "monocle_pseudotime", pt.size = 0, cols =clusters, group.by = "initial_clusters") +
  scale_x_discrete(limits=c("11","4","7",9,10,1,0,2,3,6,8,5))
```




```{r}
VlnPlot(sbs, "monocle_pseudotime", pt.size = 0, cols =c(prog_cols, pal) )
VlnPlot(sbs, "monocle_pseudotime", pt.size = 0, cols =c(prog_cols, pal[c(2,3)]) ) + 
  scale_x_discrete(limits=c(paste0("p",c(7,0,5,6,4,3,1,2)), paste0("day",c(1,3))))
```
```{r}
VlnPlot(sbs, "monocle_pseudotime", pt.size = 0, cols =c(prog_cols,clusters), group.by = "if_progen_cluster")
VlnPlot(sbs, "monocle_pseudotime", pt.size = 0, cols =c(prog_cols,clusters), group.by = "if_progen_cluster") +
   scale_x_discrete(limits=c(paste0("p",c(4,0,3,1,2)), paste0("a",c(0,7,2,3,5))))
```

### velocity pseudotime

```{r}
if (subset_data){
  velo = readRDS(here(odir,"10%_dynamical_velocity_scexperiment.rds"))
} else {
  velo = readRDS(here(odir,"dynamical_velocity_scexperiment.rds"))
}
velo
```

velocity analysis dropped one cell from day3 somehow which makes matching a bit trickier

```{r}
pseudotime_vals <- rep(NA, length = length(colnames(sbs)))
names(pseudotime_vals) <- colnames(sbs)
existing_cells <- intersect(colnames(sbs), colnames(velo))
pseudotime_vals[existing_cells] <- colData(velo)[existing_cells,'velocity_pseudotime']
summary(pseudotime_vals)
```


```{r tsne_velocity_pseudotime}
sbs$velocity_pseudotime <- pseudotime_vals
FeaturePlot(sbs, "velocity_pseudotime",reduction = "tsne", pt.size = 0.5)+ scale_colour_viridis(direction=-1)
```


## Figure 6 ViolinPlot 

```{r velocity_monocle_violin}
VlnPlot(sbs, c("velocity_pseudotime","monocle_pseudotime"), stack=T, flip = T, pt.size = 0, cols =c(prog_cols, pal),
        group.by="time.progen_clusters",split.by="time.progen_clusters"
        ) +ylab("Pseudotime (z-score)") + theme_hp

VlnPlot(sbs,  c("velocity_pseudotime","monocle_pseudotime"),  group.by="time.progen_clusters",split.by="time.progen_clusters",
        stack=T, flip = T, pt.size = 0, cols =c(prog_cols, pal))+ 
  scale_x_discrete(limits=c(paste0("p",c(5,6,7,0,3,4,2,1)), paste0("day",c(1,3))))+ 
  ylab("Pseudotime (z-score)") + theme_hp
```

```{r velocity_monocle_flipped}
VlnPlot(sbs,  c("velocity_pseudotime","monocle_pseudotime"),  group.by="time.progen_clusters",split.by="time.progen_clusters",
        stack=T, pt.size = 0, cols =c(prog_cols, pal))+ 
  scale_y_discrete(limits=rev(c(paste0("p",c(5,6,7,0,3,4,2,1)), paste0("day",c(1,3)))))+ 
  scale_x_continuous(breaks=c(0,0.25,0.5,0.75,1), expand=expansion(mult=c(0,0.05))) &
  xlab("Pseudotime ") & theme_hp & 
  theme(legend.position = "none") 

VlnPlot(sbs,  c("velocity_pseudotime","monocle_pseudotime"),  group.by="time.progen_clusters",split.by="time.progen_clusters",
         ncol=1, pt.size = 0, cols =c(prog_cols, pal)) & coord_flip() &
  scale_x_discrete(limits=rev(c(paste0("p",c(5,6,7,0,3,4,2,1)), paste0("day",c(1,3))))) &
  ylab("Pseudotime") & theme_hp & 
  theme(axis.text.x = element_text(angle=0), plot.title = element_text(size=10),
        axis.text.y= element_text(size=10),legend.position = "none", axis.title = element_text(size=12)) &
   scale_y_continuous(expand=expansion(mult=c(0,0.05)))
```

